---
title: "A1 · Getting started"
description: "A five-minute first fit on a five-site toy dataset."
output:
  rmarkdown::html_vignette:
    fig_width: 7
    fig_height: 4.5
vignette: >
  %\VignetteIndexEntry{A1 · Getting started}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  echo     = TRUE,
  collapse = TRUE,
  comment  = "#>",
  message  = FALSE,
  warning  = FALSE
)
library(bayesEfron)
```

## Problem framing

`bayesEfron` addresses the canonical univariate random-effects
meta-analytic deconvolution problem. For each of $K$ sites the
analyst observes a point estimate $\hat\theta_i$ and a within-study
standard error $\sigma_i$, with $\sigma_i$ allowed to vary across
sites (heteroscedastic). The generative assumption is two-stage:
the latent site effects $\theta_i$ are drawn from a continuous
mixing distribution $g(\theta)$, and the observed estimates satisfy
$\hat\theta_i \mid \theta_i \sim \mathcal{N}(\theta_i,\,\sigma_i^2)$.

The package decomposes the meta-analyst's question into **two
distinct inferential targets**:

1. **The mixing distribution $g(\theta)$ itself.** A nonparametric
   density estimated through Efron's log-spline prior on a fixed
   grid of effect-size values. This is the "population"
   distribution of true site effects.
2. **The per-site posteriors $\theta_i \mid \hat\theta_i, \sigma_i$.**
   Empirical-Bayes shrunken estimates, with each $\theta_i$ pulled
   toward $g$ by an amount that depends on $\sigma_i$.

A single call to `bayes_efron_fit()` returns both. The remainder of
this vignette walks through a five-site toy fit and reads the two
targets off the returned object.

## Inputs

The toy dataset is five sites with effect estimates ranging from
$-0.21$ to $0.61$ and within-study standard errors between $0.15$
and $0.24$.

```{r inputs}
theta_hat <- c(-0.21, 0.04, 0.19, 0.38, 0.61)
sigma     <- c( 0.18, 0.15, 0.22, 0.19, 0.24)
```

## Convert with `as_bef_data()`

The package's input class normalizes the two parallel vectors and
records a `source` tag for provenance.

```{r as-bef}
dat <- as_bef_data(list(theta_hat = theta_hat, sigma = sigma))
dat
```

## The fit call

```{r fit-call, eval = FALSE}
fit <- bayes_efron_fit(
  theta_hat     = c(-0.21, 0.04, 0.19, 0.38, 0.61),
  sigma         = c( 0.18, 0.15, 0.22, 0.19, 0.24),
  L             = 51L,
  M             = 3L,
  chains        = 1L,
  iter_warmup   = 150L,
  iter_sampling = 100L,
  seed          = 1234L
)
```

`L = 51L` is the discrete-grid resolution and `M = 3L` is the
natural-cubic-spline degrees of freedom; both are deliberately small
for a smoke-scale fit. The warmup and sampling counts are short for
the same reason. Substantive analyses raise these to package
defaults (four chains × 1000 warmup × 3000 sampling, $L = 101$,
$M = 6$); see the **Scaling up** section at the end.

## Cached fixture

To keep this vignette free of a CmdStan dependency at render time
we load a cached fixture produced by exactly the call above.

```{r load-fit}
fit <- readRDS(
  system.file("examples", "cached_fit_re_smoke.rds",
              package = "bayesEfron")
)
class(fit)
```

## `summary(fit)`

```{r summary}
fit_summary <- summary(fit)
fit_summary
```

The printed object has three blocks. **Prior g** holds the
posterior-mean summary of the mixing distribution — the answer to
target (1). **Diagnostics** reports Hamiltonian Monte Carlo health
and the Stan-model fingerprint. **Theta summary** is the per-site
posterior table — the answer to target (2).

## The two targets, side by side

The same `fit` object carries the population-level and site-level
quantities. Pulling them into one table makes the contrast
explicit.

```{r two-targets}
ps         <- fit_summary$prior_summary
post_means <- coef(fit)

prior_row <- data.frame(
  level    = "Prior g (population)",
  site     = NA_integer_,
  estimate = ps$mean,
  scale    = ps$sd,
  note     = "mean and sd of g(theta)"
)

site_rows <- data.frame(
  level    = "Posterior theta_i (per site)",
  site     = seq_along(post_means),
  estimate = unname(post_means),
  scale    = fit_summary$theta_summary$sd,
  note     = "posterior mean and sd"
)

knitr::kable(
  rbind(prior_row, site_rows),
  digits  = 3,
  caption = paste(
    "Two inferential targets from one fit:",
    "the prior g (top row) and the per-site posterior latent",
    "effects (rows 1-5)."
  )
)
```

Two patterns to read off the table. First, every posterior
`estimate` falls between the corresponding raw $\hat\theta_i$ and
the prior mean of $g$ — the shrinkage signature. Site 5, the most
extreme observed estimate ($\hat\theta_5 = 0.61$), shrinks furthest
in absolute terms toward the prior mean. Second, each posterior
`scale` value is smaller than the corresponding $\sigma_i$,
reflecting the precision gain from partial pooling under $g$.

## Caterpillar plot

The caterpillar plot reads the posterior table visually.

```{r caterpillar, fig.cap = "Per-site posterior latent effects with 90 percent credible intervals from the cached five-site smoke fit. Sites are ordered by site index."}
plot(fit, type = "caterpillar")
```

The vertical scatter of the posterior means traces the rank order
of the raw estimates; the interval half-widths reflect $\sigma_i$
together with the regularization induced by $g$.

## Diagnostics, briefly

The cached fixture uses one chain with 100 sampling iterations — a
smoke-scale configuration. The **Diagnostics** block in
`summary(fit)` shows $\widehat{R} \approx 1.2$ and bulk ESS in the
low teens, both well outside the conventional release thresholds
($\widehat{R} < 1.01$, ESS $> 400$). That is expected for a
deliberately short run and is the reason this configuration is
reserved for compile-and-run verification, not for inference. See
the **A5 · Diagnostics in practice** vignette for the diagnostic
checklist used on substantive fits.

## What's next

Three follow-on vignettes carry the workflow forward:

- **A2 · Anatomy of a `bayesEfron` fit** — every field of the
  returned object, every S3 method on the parent / RE-child class
  hierarchy.
- **A3 · From `metafor::escalc()` to bayesEfron** — the
  end-to-end path starting from group treatment / control summaries
  through a working fit and its outputs.
- **M1 · The empirical-Bayes deconvolution problem** — the formal
  motivation for the log-spline prior, the role of fully Bayesian
  inference over Efron's original empirical-Bayes formulation, and
  the connection to the broader literature.

## Scaling up

The cached fit uses five sites, one chain, and 100 sampling
iterations. Production analyses should:

* Use the package defaults (`L = 101`, `M = 6`,
  `chains = 4`, `iter_warmup = 1000`, `iter_sampling = 3000`).
* Confirm $\widehat{R} < 1.01$ with bulk ESS above 400 for the
  posterior-of-interest summaries.
* Replace the toy vectors with `as_bef_data()` applied to either a
  named list of `theta_hat` / `sigma` or a `metafor::escalc()`
  object — the **A3** vignette walks through the latter.

The compile-and-cache layer ensures the Stan model is compiled
once; subsequent fits with matching model source, package version,
CmdStan version, and platform reuse the cached binary. The first
call in a fresh session pays the compilation cost (typically tens
of seconds); later calls do not.
