---
title: "A2 · Anatomy of a bayesEfron fit"
description: "Every field of the bef_fit_re object and every S3 method."
output:
  rmarkdown::html_vignette:
    fig_width: 7
    fig_height: 4.5
vignette: >
  %\VignetteIndexEntry{A2 · Anatomy of a bayesEfron fit}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  echo     = TRUE,
  collapse = TRUE,
  comment  = "#>",
  message  = FALSE,
  warning  = FALSE,
  fig.width  = 7,
  fig.height = 4.5
)
library(bayesEfron)

fit <- readRDS(system.file(
  "examples", "cached_fit_re_smoke.rds",
  package = "bayesEfron"
))
```

## Scope

The **A1 · Getting started** vignette ran a five-site fit, called
`summary()`, and read off the two inferential targets — the mixing
distribution $g(\theta)$ and the per-site posteriors $\theta_i \mid
\hat\theta_i, \sigma_i$. This vignette opens up the same object.
Every visible field of the returned `bef_fit_re`, every attribute
payload, and every S3 method is named, shown, and explained.

All examples below use the cached five-site smoke fit shipped with
the package:

```{r load-shown, eval = FALSE}
fit <- readRDS(system.file(
  "examples", "cached_fit_re_smoke.rds",
  package = "bayesEfron"
))
```

No code chunk in this vignette compiles or samples from Stan; the
fixture stands in for a live `bayes_efron_fit()` call.

## Object structure

The returned object carries two S3 classes — a family-agnostic
parent and an RE-specific child — and three top-level slots.

```{r class}
class(fit)
```

```{r top-level-structure}
str(fit, max.level = 1)
```

The three top-level fields are coordinated. `draws` is the raw
`posterior::draws_array` of MCMC draws across all monitored
parameters and generated quantities. `metadata` is a closed list of
exactly 13 named fields, with four additional payloads stored as
attributes. `posterior` is a tidy, by-parameter representation of
the draws that the S3 methods consume; it does not replace `draws`,
but flattens what those methods need into a faster-to-index form.

## The 13 visible metadata fields

The metadata contract is closed: `bayes_efron_fit()` always populates
exactly the same 13 names in the same order. Treat any departure as
a bug.

```{r metadata-names}
names(fit$metadata)
```

The 13 fields divide naturally into five groupings.

```{r metadata-table, echo = FALSE}
knitr::kable(
  data.frame(
    Field = c(
      "`model_family`",
      "`grid_method`",
      "`seed`, `cmdstan_version`, `stan_file_sha256`",
      "`data_list`",
      "`runtime_seconds`",
      "`mean_g_summary`, `var_g_summary`",
      "`theta_summary`",
      "`theta_rep_draws`",
      "`effective_params_summary`, `log_marginal_likelihood_summary`"
    ),
    Purpose = c(
      "Model family identifier; `\"RE\"` only in v0.1.",
      "Which of the four grid recipes was used.",
      "Reproducibility triple: effective seed, CmdStan version, SHA-256 of the locked Stan source.",
      "The seven-field Stan data block sent to CmdStan.",
      "Sampler wall-clock in seconds.",
      "Posterior summaries of two functionals of the mixing distribution $g$.",
      "Per-site posterior table: one row per site, columns include `mean`, `sd`, `hpdi_lower`, `hpdi_upper`, `map`.",
      "Replicated effects ($J \\times K$) used for the posterior-predictive `confint(type = \"theta\")` path.",
      "Model-quality summaries: effective parameter count and log marginal likelihood."
    ),
    stringsAsFactors = FALSE
  ),
  caption = "The 13 visible metadata fields of `bef_fit_re`, grouped by purpose."
)
```

The reproducibility triple is the part that matters for archival.
These fields, together with `data_list`, sampler settings, package
version, and a matching CmdStan/toolchain environment, make the
fit auditable and practically replayable. They should not be
described as a cross-platform guarantee of byte-identical draws.

`data_list` is a passthrough record of the Stan data block, exposed
so that downstream tools can recover the exact inputs CmdStan saw
without re-running the grid recipe:

```{r data-list-names}
names(fit$metadata$data_list)
```

`theta_summary` is the per-site table that `coef()`, `vcov()`,
`as.data.frame()`, and `confint(type = "theta")` ultimately read
from; `theta_rep_draws` is the matrix of posterior-predictive
replicates that powers `confint(type = "theta")`. Both are
documented as part of the metadata contract precisely so users can
reach for them directly when an unsupported aggregation is needed.

## The four attribute payloads

Four further objects live as attributes of `fit$metadata` rather
than as named slots: one secondary functional of $g$, one
sampler-diagnostic record, and two diagnostic bookkeeping flags.

```{r attribute-names}
setdiff(names(attributes(fit$metadata)), "names")
```

The package treats these as durable but not part of the print
surface. They are read through dedicated S3 methods rather than
direct `attr()` calls:

- `sd_g_summary` — posterior summary of $\mathrm{sd}(g)$; surfaced
  by `summary()` (as `prior_summary$sd`) and by
  `confint(type = "g")`.
- `diagnostics` — the HMC diagnostic record:
  `rhat`, `ess_bulk`, `ess_tail`, `divergences`, `max_treedepth`.
  Surfaced by `summary()` and by `diagnose()`.
- `diagnostic_skipped` — character vector of diagnostic names that
  the post-processor refused to compute (empty on a clean fit).
- `sampler_diagnostics_failed` — character vector of HMC sampler
  diagnostics that the underlying CmdStanMCMC call could not return
  (empty on a clean fit).

For the cached smoke fit both bookkeeping flags are empty character
vectors, which is the expected shape on any clean run:

```{r attribute-flags}
attr(fit$metadata, "diagnostic_skipped")
attr(fit$metadata, "sampler_diagnostics_failed")
```

A non-empty value in either flag is the post-processor's way of
recording that one or more diagnostic computations could not
complete; `summary()` and `diagnose()` propagate the names without
guessing at substitute values.

## The S3 method surface

The methods table from `?bayesEfron-methods` enumerates the
coordinated method set. The subsections below show one concrete
call per method against the cached fit.

### `summary(fit)` and `print(fit)`

`summary(fit)` is the user-facing entry point: it returns a
three-block `summary.bef_fit_re` object covering the prior $g$
summary, the HMC diagnostic record, and the per-site `theta_summary`
table. Its output was the subject of the **A1** walkthrough and is
not repeated here.

`print(fit)` returns the compact one-screen overview that fits in a
console snippet. It does not enumerate sites; for that, call
`summary()` or `as.data.frame()`.

```{r print-fit}
print(fit)
```

### `coef(fit)`: point estimates

`coef()` dispatches on the RE-specific child class and returns a
named numeric vector of per-site point estimates. The default,
`type = "mean"`, returns posterior means.

```{r coef-mean}
coef(fit)
```

`type = "map"` returns the posterior mode (maximum a posteriori)
column from the same table.

```{r coef-map}
coef(fit, type = "map")
```

The two columns rarely coincide exactly, but for a unimodal
per-site posterior on a continuous support they typically agree to
within a fraction of $\sigma_i$.

### `vcov(fit)`: diagonal posterior covariance

`vcov()` returns a $K \times K$ diagonal matrix whose entries are
the squared posterior standard deviations of the per-site
$\theta_i$. The matrix is diagonal by construction: the package does
not model posterior dependence across sites at the $\theta_i$ level.

```{r vcov-fit}
vcov(fit)
```

The dimnames are the site labels (or the numeric site indices when
labels are absent, as in this smoke fixture).

### `confint(fit)` and `confint(fit, type = "g")`

`confint(fit)` returns a data frame of credible intervals on the
per-site latent effects $\theta_i$. The default credible level is
$90\%$ (`level = 0.9`); the intervals are read directly from the
`theta_rep_draws` posterior-predictive matrix.

```{r confint-default}
confint(fit)
```

Switching `type = "g"` returns intervals on the three functionals
of the mixing distribution stored under `posterior`: the mean of
$g$, the variance of $g$, and the standard deviation of $g$. The
`level` argument controls the credible width independently of the
type.

```{r confint-g-95}
confint(fit, level = 0.95, type = "g")
```

The `parm` argument (not shown here) accepts numeric site indices,
character site labels, or character functional names
(`"mean_g"`, `"var_g"`, `"sd_g"`) for selecting a subset.

### `nobs(fit)`: site count

`nobs()` is dispatched on the parent class and returns the integer
site count $K$ pulled from `fit$metadata$data_list$K`.

```{r nobs-fit}
nobs(fit)
```

### `logLik(fit)`: log marginal likelihood

`logLik()` returns a base-R `logLik` object whose value is the
posterior mean of the log marginal likelihood. The `df` attribute
holds the posterior mean of the effective parameter count, and
`nobs` is the site count. The object is exposed for inspection and
for user-defined comparisons; formal AIC/BIC or LOO-CV guidance is
not part of v0.1.0.

```{r loglik-fit}
logLik(fit)
```

### `as.data.frame(fit)`: per-site table

`as.data.frame()` returns the `theta_summary` table directly, with
one row per site and the same `site`, `mean`, `sd`, `hpdi_lower`,
`hpdi_upper`, and `map` columns that `summary()` surfaces.

```{r as-data-frame-fit}
as.data.frame(fit)
```

This is the method to reach for when feeding a `bayesEfron` fit
into a downstream pipeline that expects a tabular per-site object;
it preserves the original column types and is stable across
releases.

### `posterior::as_draws(fit)`: draws array

For workflows that want to drive the draws directly through the
`posterior` or `bayesplot` ecosystems, the `posterior::as_draws()`
generic returns the underlying `draws_array`.

```{r as-draws}
da <- posterior::as_draws(fit)
class(da)
dim(da)
```

The array dimensions are `iterations × chains × variables`. For
the cached smoke fit this is $100 \times 1 \times 183$ — 100
sampling iterations from a single chain, with 183 monitored
parameters and generated quantities.

## Diagnostics object

`diagnose()` is the structured producer for the `bef_diagnostic`
class. It pulls the `diagnostics` attribute together with the
durable model-quality summaries and emits a single object that can
be passed around, printed, and inspected without re-reaching into
the metadata attributes.

```{r diagnose-fit}
diag <- diagnose(fit)
class(diag)
diag
```

The printed view shows the worst-case $\widehat{R}$, the smallest
bulk and tail ESS, the total divergence count, and the number of
maximum-treedepth hits, alongside the runtime, the Stan SHA-256,
and the effective parameter summary. The full diagnostic checklist
— including release thresholds and remediation patterns — is the
subject of the **A5 · Diagnostics in practice** vignette.

## What's next

The remaining release vignettes carry the workflow forward along
four orthogonal axes.

- **A3 · From `metafor::escalc()` to bayesEfron** — the end-to-end
  path that begins with raw treatment / control group summaries,
  computes effect sizes with `metafor::escalc()`, and lands at a
  fitted object whose anatomy is the one shown here.
- **A4 · Choosing a grid recipe** — when to use
  `"paper_realdata"`, when an oracle-aware
  `"paper_simulation"` or `"paper_sensitivity"` run is appropriate,
  and what the experimental `"kl_target_experimental"` recipe
  actually targets.
- **A5 · Diagnostics in practice** — the release-quality checklist
  for the `diagnostics` attribute and the `bef_diagnostic` object,
  including the conventional $\widehat{R}$ and ESS thresholds and
  how to act on divergence or treedepth flags.
- **A6 · Plotting `bayesEfron` fits** — the `plot.bef_fit_re()`
  surface, including the caterpillar plot used in **A1**, the
  prior-$g$ density, and the posterior-predictive plots that read
  off `theta_rep_draws`.
