---
title: "A3 · From metafor::escalc() to bayesEfron"
description: "End-to-end path from group-arm summaries through escalc to a fitted bayesEfron model."
output:
  rmarkdown::html_vignette:
    fig_width: 7
    fig_height: 4.5
vignette: >
  %\VignetteIndexEntry{A3 · From metafor::escalc() to bayesEfron}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  echo     = TRUE,
  collapse = TRUE,
  comment  = "#>",
  message  = FALSE,
  warning  = FALSE,
  fig.width  = 7,
  fig.height = 4.5
)
library(bayesEfron)
has_metafor <- requireNamespace("metafor", quietly = TRUE)
```

## Scope

The **A1 · Getting started** vignette began at the effect-size
scale: a length-five `theta_hat` vector and a matching `sigma`
vector dropped straight into `bayes_efron_fit()`. Many applied
meta-analyses do not begin there. The raw record is per-arm
summaries — group means, standard deviations, and sample sizes —
and the analyst's first task is to convert those summaries into a
per-site estimate and its sampling standard error. This vignette
walks the full path: from the raw aggregate data, through
`metafor::escalc()`, through the `as_bef_data()` bridge, into
`bayes_efron_fit()`, and out to the standard outputs.

## The starting point

Consider a randomized education trial run across $K = 10$ schools.
Each school contributes a treatment arm and a control arm, with
each arm reporting an outcome mean, the within-arm standard
deviation, and the per-arm student count. The outcome is a
standardized end-of-year test score; the program assignment was
randomized within each school. The substantive research question
has two parts. First, what does the *population* distribution of
school-level program effects look like — concentrated around a
common effect, or genuinely heterogeneous? Second, what is each
*individual* school's program effect after partial pooling absorbs
the noise in its small-sample estimate?

The raw record is the following data frame. Column names follow the
convention used by `metafor::escalc()` for two-sample summaries:
`m1i`, `sd1i`, `n1i` for the treatment arm and `m2i`, `sd2i`, `n2i`
for the control arm.

```{r raw-data}
trial <- data.frame(
  site = paste0("site_", 1:10),
  m1i  = c(0.42, 0.18, 0.05, 0.31, 0.54,
           0.12, 0.39, 0.27, 0.48, 0.16),
  sd1i = c(1.05, 1.12, 0.98, 1.08, 1.15,
           1.03, 1.10, 0.96, 1.20, 1.07),
  n1i  = c(42L, 39L, 44L, 41L, 40L,
           43L, 37L, 45L, 38L, 46L),
  m2i  = c(0.05, 0.02, -0.08, 0.04, 0.10,
           -0.03, 0.11, 0.00, 0.08, -0.02),
  sd2i = c(1.01, 1.09, 1.04, 1.02, 1.12,
           1.00, 1.06, 0.99, 1.16, 1.05),
  n2i  = c(40L, 38L, 42L, 39L, 41L,
           44L, 36L, 43L, 39L, 45L)
)
trial
```

Treatment-arm means range from $0.05$ to $0.54$, control-arm means
from $-0.08$ to $0.11$, and per-arm sample sizes from $36$ to $46$
students. Three features of the record matter for the bridge. The
two arms at each site are independent random samples, so the
sampling variance of the per-site mean difference is the sum of
the two per-arm sampling variances; `metafor::escalc()` computes
this for the analyst. The outcome scale is the standardized test
score, so the mean-difference effect size is interpretable on the
same scale as a Cohen's-$d$-style standardized mean difference
without an additional pooling step. The per-arm sample sizes are
modest enough that within-site sampling noise is large relative to
plausible true between-site variation — the regime in which a
hierarchical estimator earns its keep.

## Effect sizes via `metafor::escalc()`

The `metafor` package is the canonical R interface for the family
of per-study effect-size and sampling-variance computations that
upstream a meta-analysis. The `escalc()` constructor takes the
group-arm summaries and a `measure` argument that selects the
estimator. For continuous outcomes on a common scale `measure =
"MD"` (the raw mean difference) is the natural choice; for studies
that report on different outcome scales `measure = "SMD"` (the
standardized mean difference, Hedges's $g$) would normalize them to
a common standard-deviation unit. The trial above uses a single
standardized test, so the raw mean difference is what we want.

```{r escalc, eval = has_metafor}
esc <- metafor::escalc(
  measure = "MD",
  m1i  = m1i,  sd1i = sd1i, n1i = n1i,
  m2i  = m2i,  sd2i = sd2i, n2i = n2i,
  data = trial,
  slab = site
)
class(esc)
```

The returned object has class `c("escalc", "data.frame")` and
carries two new columns alongside the original ones.

```{r escalc-table, eval = has_metafor}
knitr::kable(
  as.data.frame(esc)[, c("site", "yi", "vi")],
  digits  = 4,
  caption = paste(
    "Per-site effect sizes from `metafor::escalc(measure = \"MD\")`.",
    "Column `yi` is the treatment-minus-control mean difference;",
    "column `vi` is its sampling variance, computed as the sum of",
    "the two per-arm variances of the mean."
  )
)
```

Read the two new columns directly. `yi` is the per-site estimate
$\widehat\theta_i = \bar y_{i,T} - \bar y_{i,C}$: the
treatment-minus-control mean difference for site $i$. `vi` is the
corresponding sampling variance
$\widehat{\mathrm{Var}}(\widehat\theta_i) =
s_{i,T}^2/n_{i,T} + s_{i,C}^2/n_{i,C}$: the textbook estimate of
the sampling variance of an independent-samples mean difference.
The sampling *standard error* — the quantity `bayesEfron` needs —
is $\sigma_i = \sqrt{v_i}$. The `slab` argument records the site
labels as an attribute of `yi`; they will travel into `escalc`'s
print method and into downstream `metafor` models, but `escalc`
itself does not promote them to the data frame's row names.

The per-site estimates from this trial run from $0.13$ to $0.44$
and the standard errors from roughly $0.21$ to $0.27$, all on the
standardized-score scale. The classical fixed-effect summary would
be the inverse-variance-weighted mean; the random-effects summary
would be a method-of-moments or REML estimate of the
between-site variance. Both throw away the per-site point
estimates after a single pass. `bayesEfron`'s contribution is to
keep them: it estimates a continuous mixing distribution $g(\theta)$
over the latent site effects and returns a posterior for each
$\theta_i$ separately, in addition to summaries of $g$ itself.

## The `as_bef_data()` bridge

The `as_bef_data()` generic adapts the `escalc` object into the
`bef_data` class that `bayes_efron_fit()` consumes. The method
extracts `theta_hat` from `yi`, computes `sigma` as `sqrt(vi)`
after a strict-positivity check on `vi`, and records a
provenance tag.

```{r as-bef-data, eval = has_metafor}
bef_dat <- as_bef_data(esc)
bef_dat
```

The compact print shows the four pieces that survived the
conversion: the site count, the source tag, and one-line min /
median / max summaries of `theta_hat` and `sigma`. The full
representation is a four-element list:

```{r as-bef-fields, eval = has_metafor}
names(bef_dat)
bef_dat$source
```

The `source = "metafor::escalc"` element is the provenance tag the
`escalc` method writes. Downstream consumers can branch on it when
the bridge matters — for example, to record the input shape in a
methods appendix without re-deriving it from object class.
`bef_dat$names` is `NULL` here because the `escalc` object's
row names are the integer-sequence default; the `slab` attribute
that travels on `yi` is the `metafor`-internal label, not a row
name, and is dropped by the bridge. Where the analyst wants the
site labels surfaced in the fit's per-site table, the simplest fix
is to set `row.names(esc) <- esc$site` before the
`as_bef_data()` call.

## Cached fixture

Running the full fit on the ten-site escalc data above requires a
working CmdStan installation and a non-trivial sampler run. So that
this vignette renders without a CmdStan dependency, the
posterior-bearing outputs below are read from the cached
five-site smoke fixture shipped at
`inst/examples/cached_fit_re_smoke.rds`. The fixture was produced
with $L = 51$, $M = 3$, one chain, 150 warmup iterations, and 100
sampling iterations on the five-site toy data of **A1**. It is
deliberately small and is reserved for compile-and-run
verification; the diagnostic profile reported below reflects that
budget, not a production fit.

The displayed outputs — `summary()`, `confint()`, `diagnose()`,
the caterpillar plot — therefore correspond to the five-site
smoke configuration, not to the ten-site escalc data computed in
the previous section. The intent is to show the *shape* of each
return value with a renderable fixture; the **Live alternative**
section below shows the call you would make against the ten-site
data when CmdStan is available locally.

```{r load-cached}
fit <- readRDS(system.file(
  "examples", "cached_fit_re_smoke.rds",
  package = "bayesEfron"
))
class(fit)
```

## Outputs

`summary(fit)` returns a three-block object with the posterior
summary of the mixing distribution $g$, the HMC diagnostic record,
and the per-site `theta_summary` table.

```{r summary-fit}
summary(fit)
```

`confint(fit)` returns the per-site credible intervals on the
latent effects $\theta_i$. The default credible level is $0.90$
and the default `type` is `"theta"`; the intervals are read from
the posterior-predictive `theta_rep_draws` matrix stored in
metadata. The same generic with `type = "g"` switches to credible
intervals on the functionals of the mixing distribution.

```{r confint-fit}
confint(fit, level = 0.9, type = "theta")
```

`diagnose(fit)` produces a `bef_diagnostic` object that bundles
the HMC sampler health record together with the durable
model-quality summaries (effective parameter count, log marginal
likelihood). The print view surfaces the worst-case
$\widehat{R}$, the smallest bulk and tail ESS, the divergence and
maximum-treedepth counts, and the Stan SHA-256. **A5**
documents the release-quality checklist this object is built to
support.

```{r diagnose-fit}
diagnose(fit)
```

`plot(fit, type = "caterpillar")` is the visual summary of the
per-site posterior table. Each site contributes one point estimate
and one credible interval, ordered by the site index.

```{r caterpillar, fig.cap = "Per-site posterior latent effects with 90 percent credible intervals from the cached five-site smoke fit. The fixture is a smoke-scale configuration (one chain, 100 sampling iterations) reserved for compile-and-run verification; production fits with the package defaults provide a larger post-warmup draw set and more reliable diagnostics."}
plot(fit, type = "caterpillar")
```

## Live alternative

The chunk below is the call that produces a fitted object for the
ten-site escalc data computed earlier in this vignette. It runs at
the same smoke-scale configuration as the cached fixture and
requires a working CmdStan installation. Set the environment
variable `BAYESEFRON_RUN_LIVE=1` before rendering to execute it
inline; otherwise it is skipped.

```{r live-fit, eval = has_metafor && identical(Sys.getenv("BAYESEFRON_RUN_LIVE"), "1")}
fit_live <- bayes_efron_fit(
  theta_hat     = bef_dat$theta_hat,
  sigma         = bef_dat$sigma,
  L             = 51L,
  M             = 3L,
  chains        = 1L,
  iter_warmup   = 150L,
  iter_sampling = 100L,
  seed          = 1234L
)
summary(fit_live)
```

The same four S3 methods shown above — `summary()`, `confint()`,
`diagnose()`, `plot()` — apply unchanged to `fit_live`. The
canonical executable form of this end-to-end path is installed at
`system.file("examples", "example-bayes-efron-fit.R", package =
"bayesEfron")` and is the script the vignette's outputs would
match in a live render.

## What's next

- **A4 · Choosing a grid recipe** — the four discrete-grid
  recipes that govern the support of $g$ and when each one is
  appropriate. The default `paper_realdata` recipe is the one a
  ten-site analysis like this vignette would adopt; the
  oracle-aware recipes are for simulation work where the true
  $g$ is known.
- **A5 · Diagnostics in practice** — the release-quality checklist
  for the `diagnose()` output, including the conventional
  $\widehat{R} < 1.01$ and bulk-ESS $> 400$ thresholds and the
  remediation patterns for divergences and treedepth hits.
- **A7 · Case study — Lee–Sui replication light** — the long-form
  case study that picks up where this vignette leaves off, taking
  a substantive dataset through the full pre-registered analysis
  and the methods-appendix template.
