---
title: "A4 · Choosing a grid"
description: "Side-by-side comparison of the four grid recipes."
output:
  rmarkdown::html_vignette:
    fig_width: 7
    fig_height: 5
vignette: >
  %\VignetteIndexEntry{A4 · Choosing a grid}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  echo     = TRUE,
  collapse = TRUE,
  comment  = "#>",
  message  = FALSE,
  warning  = FALSE,
  fig.width  = 7,
  fig.height = 5
)
library(bayesEfron)
has_ggplot2 <- requireNamespace("ggplot2", quietly = TRUE)
```

## Why the grid matters

`bayesEfron` represents the mixing distribution
$g(\theta)$ on a discrete support of length $L$ together with a
length-$M$ natural-cubic-spline basis built on that support. The Stan
model never sees a continuous $g$: it sees the log-spline
coefficients $\boldsymbol\alpha \in \mathbb{R}^M$ that map, via the
basis matrix $B \in \mathbb{R}^{L \times M}$, to the discrete
probabilities $g_\ell \propto \exp\{(B\boldsymbol\alpha)_\ell\}$ at
the $L$ grid points. The grid is therefore not a presentation
detail; it is the support on which $g$ is defined.

Three design choices govern that support — the lower and upper
endpoints, the number of points between them, and the spline
degrees of freedom. `make_efron_grid()` exposes four named recipes
that fix the first two for the analyst: `paper_realdata`,
`paper_simulation`, `paper_sensitivity`, and
`kl_target_experimental`. Each recipe encodes a different rule for
the endpoints, with `L` and $M$ left at sensible defaults.

When the grid is too narrow, the spline basis cannot represent mass
that the data place outside the support, and the per-site posteriors
develop pile-ups against the boundary. When it is too wide, the
basis spends degrees of freedom on regions where $g$ should be
exactly zero, and the posterior mean of $g$ acquires diffuse tails
that suppress shrinkage in the centre. The recipes below fix the
endpoints by formulas the package treats as part of its public
contract; this vignette shows what those formulas produce on a
shared toy dataset and how to choose between them.

## A shared toy dataset

The following seven-site dataset is used throughout. `theta_hat` is
the observed per-site estimate, `sigma` is the within-study standard
error, and `theta_true` is the oracle vector used by the two
simulation-only recipes.

```{r toy-data}
theta_hat  <- c(-0.45, -0.20, 0.05, 0.20, 0.40, 0.60, 0.90)
sigma      <- c( 0.18,  0.22, 0.20, 0.16, 0.19, 0.24, 0.21)
theta_true <- c(-0.50, -0.10, 0.00, 0.25, 0.45, 0.70, 1.00)

toy <- data.frame(
  site       = paste0("site_", 1:7),
  theta_hat  = theta_hat,
  sigma      = sigma,
  theta_true = theta_true
)
toy
```

The observed range of `theta_hat` spans $1.35$ units; the oracle
range of `theta_true` spans $1.50$ units. The recipes differ in
which of the two ranges they read and by how much they widen it.

## Recipe 1 — `paper_realdata`

The default recipe takes the observed range
$[\min\hat\theta_i, \max\hat\theta_i]$ and expands it symmetrically
by `expansion` times the range width. With the package default
`expansion = 0.5` the support extends $50\%$ of the observed range
past each endpoint.

```{r grid-realdata}
g_real <- make_efron_grid(
  theta_hat = theta_hat,
  sigma     = sigma,
  grid_method = "paper_realdata"
)

g_real$grid_method
c(L = g_real$L, M = g_real$M, expansion = g_real$expansion)
range(g_real$grid)
g_real$attribution
```

The returned list carries the same eight fields for every recipe:
`grid`, `B`, `M`, `L`, `expansion`, `kappa`, `grid_method`, and
`attribution`. The first four describe the discrete support and
basis. `expansion` records the range-relative widening factor that
was applied (an `NA_real_` when the recipe does not use one).
`kappa` is the KL target (`NULL` here; only meaningful for the
experimental recipe). `attribution` records the formula tag and the
source file lineage so downstream consumers can audit which recipe
produced the grid.

The plot below shows the grid points as a rug along the bottom of
the panel, the seven observed `theta_hat` values as labelled points,
and the six spline-basis curves overlaid. The basis is the
$L \times M$ matrix returned in `g_real$B`; each column is one
natural-cubic-spline basis function evaluated on the grid.

```{r plot-realdata, eval = has_ggplot2, fig.height = 3.5, fig.alt = "Natural-cubic-spline basis for the paper_realdata grid on the seven-site toy dataset; rug marks grid points, black markers show observed theta_hat values, coloured curves show the basis functions."}
plot_grid <- function(g, theta_hat, label) {
  basis_df <- data.frame(
    x     = rep(g$grid, ncol(g$B)),
    y     = as.numeric(g$B),
    basis = factor(rep(seq_len(ncol(g$B)), each = length(g$grid)))
  )
  ggplot2::ggplot() +
    ggplot2::geom_line(
      data = basis_df,
      ggplot2::aes(x = x, y = y, colour = basis),
      alpha = 0.7
    ) +
    ggplot2::geom_rug(
      data = data.frame(x = g$grid),
      ggplot2::aes(x = x), sides = "b", alpha = 0.3
    ) +
    ggplot2::geom_point(
      data = data.frame(x = theta_hat, y = 0),
      ggplot2::aes(x = x, y = y), colour = "black", size = 2
    ) +
    ggplot2::labs(
      title = label,
      x = expression(theta),
      y = "spline basis"
    ) +
    ggplot2::guides(colour = "none") +
    ggplot2::theme_minimal()
}

plot_grid(g_real, theta_hat, label = "paper_realdata, expansion = 0.5")
```

## Recipe 2 — `paper_simulation`

The simulation recipe is oracle-aware: it reads the oracle vector
`theta_true` and sets the endpoints to
$[\min\theta_i^{\text{true}} - 0.5,\,\max\theta_i^{\text{true}} +
0.5]$. The widening is an *absolute* $0.5$-unit pad, not a
range-relative fraction; this matches the formula used in the
paper's simulation study. `expansion` is consequently reported as
`NA_real_`.

```{r grid-simulation}
g_sim <- make_efron_grid(
  theta_hat   = theta_hat,
  sigma       = sigma,
  theta_true  = theta_true,
  grid_method = "paper_simulation"
)

g_sim$grid_method
c(L = g_sim$L, M = g_sim$M)
g_sim$expansion
range(g_sim$grid)
g_sim$attribution
```

The recipe requires `theta_true`; calling it without an oracle
returns a `bef_err_grid_oracle_required` condition. That is by
design — the simulation rule is only well-defined when a known
truth exists, and silently falling back to the observed range would
mask the distinction between the two regimes.

```{r plot-simulation, eval = has_ggplot2, fig.height = 3.5, fig.alt = "Natural-cubic-spline basis for the paper_simulation grid (absolute +/- 0.5 pad on theta_true) on the seven-site toy dataset; rug marks grid points, black markers show observed theta_hat values, coloured curves show the basis functions."}
plot_grid(g_sim, theta_hat, label = "paper_simulation, absolute +/- 0.5")
```

Because the oracle range exceeds the observed range here, the
simulation grid is wider on the left endpoint than the real-data
grid. In small-$K$ simulations the observed range can be
substantially narrower than the support of $g$, which is exactly
the case the absolute-pad rule is designed for.

## Recipe 3 — `paper_sensitivity`

The sensitivity recipe is also oracle-aware but uses a
range-relative widening, parameterised by `bound_expansion`. The
endpoints are
$[\min\theta_i^{\text{true}} - b \cdot W,\,\max\theta_i^{\text{true}}
+ b \cdot W]$ where $W$ is the width of the oracle range and $b$ is
`bound_expansion`. The default is $b = 0.5$.

```{r grid-sensitivity}
g_sens <- make_efron_grid(
  theta_hat       = theta_hat,
  sigma           = sigma,
  theta_true      = theta_true,
  grid_method     = "paper_sensitivity",
  bound_expansion = 0.5
)

g_sens$grid_method
c(L = g_sens$L, M = g_sens$M, bound_expansion = g_sens$expansion)
range(g_sens$grid)
g_sens$attribution
```

The `expansion` slot reports the effective `bound_expansion` here,
which is the source of the formula tag
`paper_sensitivity_relative` in the attribution record. Sweeping
`bound_expansion` over a small grid (e.g. $\{0.25, 0.5, 1.0\}$) is
the canonical way to read off how sensitive the per-site posterior
means and the posterior of $g$ are to the choice of support — the
exercise A6's sensitivity section walks through end to end. As with
`paper_simulation`, an oracle is required.

```{r plot-sensitivity, eval = has_ggplot2, fig.height = 3.5, fig.alt = "Natural-cubic-spline basis for the paper_sensitivity grid (bound_expansion = 0.5) on the seven-site toy dataset; rug marks grid points, black markers show observed theta_hat values, coloured curves show the basis functions."}
plot_grid(g_sens, theta_hat, label = "paper_sensitivity, bound_expansion = 0.5")
```

## Recipe 4 — `kl_target_experimental`

The fourth recipe derives the grid *spacing* from a Kullback-Leibler
target $\kappa$ between adjacent Gaussian kernels of width
$\min(\sigma)$, then sets $L$ to the integer count that fills the
expanded observed range at that spacing. The formula is
$d = 2 \cdot \min(\sigma) \cdot \sqrt{\exp(2\kappa) - 1}$ and the
default for $\kappa$ is $1/K$ where $K = $ `length(theta_hat)`.
The function emits a once-per-session disclaimer noting that this
calibration is experimental for heteroscedastic inputs; the call
below wraps it in `suppressMessages()` so the message does not
appear in the rendered vignette.

```{r grid-kl}
g_kl <- suppressMessages(
  make_efron_grid(
    theta_hat   = theta_hat,
    sigma       = sigma,
    grid_method = "kl_target_experimental"
  )
)

g_kl$grid_method
c(L = g_kl$L, M = g_kl$M, expansion = g_kl$expansion, kappa = g_kl$kappa)
range(g_kl$grid)
g_kl$attribution
```

The recipe also enforces $L \in [51, 300]$: if the KL spacing
implies a count outside this window, $L$ is clipped to the nearest
boundary. The experimental disclaimer is emitted once per session;
an additional informational message is emitted when the derived
count is clipped to the `[51, 300]` bounds. With the toy dataset
above the implied count falls below $51$, so $L$ is held at the
lower bound. The
attribution string records the underlying formula and credits the
`ebnm::ebnm_scale_npmle()` calibration that motivated it.

```{r plot-kl, eval = has_ggplot2, fig.height = 3.5, fig.alt = "Natural-cubic-spline basis for the kl_target_experimental grid on the seven-site toy dataset; rug marks grid points, black markers show observed theta_hat values, coloured curves show the basis functions."}
plot_grid(g_kl, theta_hat, label = "kl_target_experimental")
```

## Side-by-side comparison

The 2 × 2 panel below assembles the four grids on a shared
$\theta$-axis. The basis curves are normalized to a common $[0, 1]$
y-range within each panel so the panels can be read side by side
without the basis amplitude dominating the comparison; the rug
along the bottom shows the grid points themselves, and the black
points are the seven observed `theta_hat` values.

```{r side-by-side, eval = has_ggplot2, fig.width = 7, fig.height = 5, dpi = 300, fig.cap = "Side-by-side comparison of the four grid recipes on the same seven-site toy dataset. Each panel shows the support points (rug), the observed theta_hat values (black points), and the natural-cubic-spline basis (coloured curves, normalized within panel)."}
grid_long <- function(g, recipe) {
  B_norm <- apply(g$B, 2, function(col) {
    rng <- range(col)
    if (diff(rng) <= 0) col else (col - rng[1]) / diff(rng)
  })
  data.frame(
    recipe = recipe,
    x      = rep(g$grid, ncol(B_norm)),
    y      = as.numeric(B_norm),
    basis  = factor(rep(seq_len(ncol(B_norm)), each = length(g$grid)))
  )
}

basis_all <- rbind(
  grid_long(g_real, "paper_realdata"),
  grid_long(g_sim,  "paper_simulation"),
  grid_long(g_sens, "paper_sensitivity"),
  grid_long(g_kl,   "kl_target_experimental")
)
basis_all$recipe <- factor(
  basis_all$recipe,
  levels = c(
    "paper_realdata", "paper_simulation",
    "paper_sensitivity", "kl_target_experimental"
  )
)

rug_all <- rbind(
  data.frame(recipe = "paper_realdata",         x = g_real$grid),
  data.frame(recipe = "paper_simulation",       x = g_sim$grid),
  data.frame(recipe = "paper_sensitivity",      x = g_sens$grid),
  data.frame(recipe = "kl_target_experimental", x = g_kl$grid)
)
rug_all$recipe <- factor(rug_all$recipe, levels = levels(basis_all$recipe))

obs_all <- do.call(rbind, lapply(
  levels(basis_all$recipe),
  function(r) data.frame(recipe = r, x = theta_hat, y = 0)
))
obs_all$recipe <- factor(obs_all$recipe, levels = levels(basis_all$recipe))

ggplot2::ggplot() +
  ggplot2::geom_line(
    data = basis_all,
    ggplot2::aes(x = x, y = y, colour = basis),
    alpha = 0.7
  ) +
  ggplot2::geom_rug(
    data = rug_all,
    ggplot2::aes(x = x), sides = "b", alpha = 0.25
  ) +
  ggplot2::geom_point(
    data = obs_all,
    ggplot2::aes(x = x, y = y), colour = "black", size = 1.5
  ) +
  ggplot2::facet_wrap(~ recipe, ncol = 2) +
  ggplot2::labs(
    x = expression(theta),
    y = "spline basis (normalized within panel)"
  ) +
  ggplot2::guides(colour = "none") +
  ggplot2::theme_minimal()
```

Three patterns are visible. First, `paper_realdata` and
`kl_target_experimental` produce identical endpoints because both
read the observed range and apply the same default $50\%$
expansion; they differ only in $L$ (and therefore in the density of
the rug and the resolution of the spline basis on the same
interval). Second, `paper_simulation` and `paper_sensitivity` both
read the oracle range, but the simulation rule pads by an absolute
$0.5$ units on each side while the sensitivity rule pads by a
fraction of the oracle width. Third, the basis shape itself is
governed entirely by the support endpoints and $M = 6$: the
interior knot placement is determined by `splines::ns()`, so when
two recipes produce the same endpoints they produce the same basis
shape and differ only in the resolution of the underlying support.

## A small results table

```{r results-table}
results <- data.frame(
  recipe       = c("paper_realdata", "paper_simulation",
                   "paper_sensitivity", "kl_target_experimental"),
  L            = c(g_real$L, g_sim$L, g_sens$L, g_kl$L),
  M            = c(g_real$M, g_sim$M, g_sens$M, g_kl$M),
  expansion    = c(g_real$expansion, g_sim$expansion,
                   g_sens$expansion, g_kl$expansion),
  kappa        = c(NA_real_, NA_real_, NA_real_, g_kl$kappa),
  support_min  = c(min(g_real$grid), min(g_sim$grid),
                   min(g_sens$grid), min(g_kl$grid)),
  support_max  = c(max(g_real$grid), max(g_sim$grid),
                   max(g_sens$grid), max(g_kl$grid)),
  stringsAsFactors = FALSE
)
knitr::kable(
  results,
  digits = 3,
  caption = paste(
    "Grid summaries for the four recipes on the shared seven-site",
    "toy dataset. `expansion` is NA for `paper_simulation` because",
    "the absolute-pad rule does not use a range-relative factor;",
    "`kappa` is NA for the three paper-faithful recipes."
  )
)
```

## Picking a recipe

The decision tree the package documentation defends is short and
data-driven.

- **Real data, no oracle.** Use `paper_realdata`. This is the
  default for `make_efron_grid()` and `bayes_efron_fit()`. The
  $50$ percent range-relative expansion is the rule used in the
  published real-data analysis and is intended to keep the
  posterior mass away from the grid boundary without making the
  support unnecessarily diffuse.

- **Simulation with known truth, paper match.** Use
  `paper_simulation`. The absolute $\pm 0.5$ pad on the oracle
  range is the rule used in the paper's simulation study, and
  matching it is the right choice for any replication of those
  numbers or for an extension of the simulation design.

- **Sensitivity check on grid width.** Use `paper_sensitivity` and
  sweep `bound_expansion` over a small set (e.g.
  $\{0.25, 0.5, 1.0\}$). Posterior summaries that move materially
  across the sweep indicate that the grid endpoints are doing
  inferential work; summaries that are stable indicate that the
  choice of endpoint is not load-bearing for the inference.
  Requires an oracle.

- **Experimental KL tuning.** Use `kl_target_experimental` with
  caution. The calibration is taken from the `ebnm` package's
  homoscedastic-input recipe, and its behavior on the
  heteroscedastic inputs `bayesEfron` accepts is not validated
  against a published number. The once-per-session disclaimer is
  part of the recipe's contract; downstream comparisons should
  preserve it so that consumers know the recipe's status.

A pragmatic default is to fit `paper_realdata` first, then re-fit
under one or two alternative recipes (or under the same recipe with
a wider `expansion`) and inspect whether the per-site posterior
means, the posterior summary of $g$, and the diagnostic record
shift in any consequential way. A6's sensitivity section is the
formal version of that check.

## What's next

- **A5 · Diagnostics in practice** — the release-quality checklist
  for the `diagnostics` attribute and the `bef_diagnostic` object,
  including the diagnostic signatures of a too-narrow or too-wide
  grid.
- **A6 · Plotting and visualization** — the visual diagnostics for
  $g$ and the per-site posteriors, including a worked sensitivity
  sweep across `paper_sensitivity` settings.
- **M4 · Grid construction and spline basis** — the methodological
  vignette with the formal specification of each recipe, the
  monotonicity and rank-deficiency postconditions enforced by
  `make_efron_grid()`, and the connection to Efron's original
  log-spline construction.
