---
title: "A5 · Diagnostics in practice"
description: "How to read diagnose() and decide whether a fit is releasable."
output:
  rmarkdown::html_vignette:
    fig_width: 7
    fig_height: 4.5
vignette: >
  %\VignetteIndexEntry{A5 · Diagnostics in practice}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  echo     = TRUE,
  collapse = TRUE,
  comment  = "#>",
  message  = FALSE,
  warning  = FALSE,
  fig.width  = 7,
  fig.height = 4.5
)
library(bayesEfron)
has_ggplot2 <- requireNamespace("ggplot2", quietly = TRUE)

fit <- readRDS(system.file(
  "examples", "cached_fit_re_smoke.rds",
  package = "bayesEfron"
))
```

## Why diagnostics matter

`bayesEfron` fits its random-effects deconvolution model with
Stan's No-U-Turn Sampler (NUTS). Every posterior summary the
package surfaces — the mean of $g(\theta)$, the per-site
intervals on $\theta_i$, the marginal log-likelihood — is a
Monte Carlo estimate that depends on the sampler having explored
the posterior well enough for those estimates to be trustworthy.
A pathological sampler can produce summaries that look reasonable
in tabular form but reflect a chain that has not converged, has
explored only one mode, or has lost numerical resolution in
high-curvature regions.

Sampler-health diagnostics are the standard mechanism for ruling
those failures out. The post-processor records the conventional
NUTS diagnostics on every fit, validates them against a fixed
schema, and surfaces them through a dedicated producer
(`diagnose()`) and S3 methods. This vignette walks through the
schema field by field, anchors each field to its conventional
release threshold, and shows what flagged diagnostics look like
by reading them off the cached smoke fixture, which is engineered
to fail the thresholds.

## The `diagnose()` producer

`diagnose()` is the structured producer for the `bef_diagnostic`
class. It does not re-process draws or run new computations; it
surfaces the diagnostic payload that the fit pipeline stored as
attributes of `fit$metadata`, behind a stable access path.

```{r diagnose-call}
diag <- diagnose(fit)
class(diag)
```

The returned object is a list of eleven fields with a single S3
class and is validated internally before return.

```{r diagnose-structure}
str(diag, max.level = 1)
```

The sampler-health fields (`rhat`, `ess_bulk`, `ess_tail`,
`divergences`, `max_treedepth`) are numeric vectors; on a
multi-chain real fit they carry per-parameter or per-chain
entries. The model-quality field `effective_params_summary` is a
posterior summary list. The provenance fields record what was
sampled and how long it took. The two bookkeeping flags
(`diagnostic_skipped`, `sampler_diagnostics_failed`) name
diagnostics the post-processor intentionally did not compute or
could not return, and are empty character vectors on a clean fit.

The producer-versus-method distinction matters for orientation.
`diagnose()` is the producer: a function that emits a
`bef_diagnostic` object. `summary()`, `print()`, and `format()`
are S3 methods that consume that object and emit human-readable
views. The graphical companion view is reached through
`plot(fit, type = "diagnostic")`.

## Each diagnostic field

The five sampler-health fields divide into three conceptual
groups: convergence (`rhat`), mixing (`ess_bulk`, `ess_tail`),
and sampler pathology (`divergences`, `max_treedepth`). The
release thresholds below are the conventional ones used in the
Stan documentation and reflected in the package's verification
ledger.

### `rhat`

`rhat` is the per-parameter $\widehat{R}$ statistic (in the
rank-normalized form recommended for current Stan workflows). It
compares the between-chain variance to the within-chain variance
for each monitored parameter; values close to $1$ indicate that
the chains have converged to a common stationary distribution.
The conventional release threshold is $\widehat{R} < 1.01$; the
$1.01$–$1.05$ band is a warning, above $1.05$ is a blocker.

When $\widehat{R}$ exceeds the threshold the conventional
remediation is to extend warmup, increase the sampling count, or
investigate whether the chains have settled into different modes
via trace plots (e.g.\ `posterior::as_draws(fit)` and
`bayesplot::mcmc_trace()`). For the cached smoke fixture:

```{r rhat-cached}
diag$rhat
```

The value is well above the $1.01$ threshold. The fixture was
sampled with a single chain and a very small iteration count
(`chains = 1`, `iter_warmup = 150`, `iter_sampling = 100`), which
makes the smoke-scale $\widehat{R}$ a deliberate teaching example
rather than an indictment of the model.

### `ess_bulk`

`ess_bulk` is the per-parameter bulk effective sample size — the
Monte Carlo information available for estimating posterior means.
A bulk ESS of $400$ means the autocorrelated draws carry as much
information as $400$ independent draws would. The conventional
release threshold is $\mathrm{ESS}_{\text{bulk}} > 400$.

When bulk ESS falls below the threshold the conventional
remediation is to extend sampling. Persistent low bulk ESS for a
specific parameter is a sign of a model identifiability problem
rather than a sampler-tuning issue. For the cached fixture:

```{r ess-bulk-cached}
diag$ess_bulk
```

The value is two orders of magnitude below the threshold, again
because of the smoke-scale iteration count.

### `ess_tail`

`ess_tail` is the per-parameter tail effective sample size, the
ESS relevant for the credible-interval endpoints that `confint()`
returns. The conventional release threshold mirrors the bulk
threshold: $\mathrm{ESS}_{\text{tail}} > 400$.

Tail ESS is typically smaller than bulk ESS, because tail
quantiles depend on rare excursions in the tails of the chain. A
passing bulk ESS paired with a failing tail ESS indicates that
posterior means are stable but credible intervals are not. For
the cached fixture:

```{r ess-tail-cached}
diag$ess_tail
```

The value is below the threshold but well above the bulk ESS,
which is the expected ordering when both fail at smoke scale.

### `divergences`

`divergences` counts the number of divergent transitions per
chain — the sampler's record of numerical failure where the
leapfrog integrator's step size is too large for the local
posterior curvature. Divergences bias posterior summaries: the
affected draws cluster in regions the sampler could not traverse
faithfully, and discarding them after the fact does not restore
unbiasedness. The conventional release threshold is $0$: a clean
fit has no divergent transitions.

When divergences appear the conventional remediation is to raise
the `adapt_delta` control argument (which lowers the adapted
step size), reparameterise the model, or both. For the cached
fixture:

```{r divergences-cached}
diag$divergences
```

The fixture's divergence count is zero, so this is the one
diagnostic that does not flag — the smoke-scale problem the
fixture exercises is convergence and mixing, not sampler
pathology.

### `max_treedepth`

`max_treedepth` counts the number of times the NUTS tree-building
algorithm hit its maximum depth without resolving a U-turn. This
is an efficiency warning, not a correctness one: the transition
is still valid, but the sampler is doing more work than expected
and may be missing a faster exploration path. The conventional
release threshold is $0$, with a small nonzero count tolerable
when the other diagnostics pass.

When max-treedepth hits appear the conventional remediation is to
raise the `max_treedepth` control argument or reparameterise to
reduce posterior correlation. For the cached fixture:

```{r treedepth-cached}
diag$max_treedepth
```

The fixture's count is zero, matching the divergence result.

## `summary(diag)` — extreme-value view

`summary()` on a `bef_diagnostic` is a triage view. It compresses
each sampler-health field to its worst-case value: `rhat` to its
maximum, `ess_bulk` and `ess_tail` to their minima, and
`divergences` and `max_treedepth` to their totals across chains.

```{r summary-diag}
summary(diag)
```

The compressed view is what to scan first on a real fit: if the
worst-case $\widehat{R}$ is below $1.01$, the minimum bulk and
tail ESS are above $400$, and the total divergence and
max-treedepth counts are zero, the fit passes the headline
checks. The full per-parameter or per-chain vectors remain on the
`bef_diagnostic` object itself for deeper investigation when the
headline fails. The view also carries the unchanged model-quality
summaries and the two bookkeeping flags, so a single `summary()`
call gives the analyst everything needed for a release decision.

## `plot(fit, type = "diagnostic")`

The graphical companion view consumes the same payload through
`plot(fit, type = "diagnostic")`. It shows the five sampler-health
metrics as a bar chart, with the worst-case value for each metric
on the vertical axis. The view is meant for quick visual triage
during interactive work, not as a replacement for reading the
numeric fields.

The function uses a `ggplot2` backend when `ggplot2` is installed
and the environment variable `BAYESEFRON_NO_GGPLOT2` is not
exactly `"1"`, and a base-graphics backend otherwise. The
`ggplot2` backend returns a `ggplot2::ggplot` object that can be
extended with `+ theme()` and `+ labs()`; the base backend draws
on the active graphics device and returns `invisible(NULL)`.

```{r plot-diagnostic, eval = has_ggplot2, fig.height = 4, fig.alt = "Diagnostic view of the cached five-site smoke fit: bar chart of Rhat, bulk ESS, tail ESS, divergences, and max-treedepth flags; ESS bars dominate the value axis."}
plot(fit, type = "diagnostic")
```

In the fixture's bar chart the bulk and tail ESS bars dwarf the
others on the value axis — the visual signature of a smoke-scale
fit, where ESS is the information-bearing diagnostic and the one
that scales fastest with iteration count.

## What the cached fixture flags reveal

The cached fixture was produced by the call documented in **A1**:
one chain, $150$ warmup iterations, $100$ sampling iterations.
These are far below what NUTS needs to converge on a
deconvolution problem of this shape. The fixture exists to keep
the package's vignettes and examples free of a CmdStan dependency
at render time, and was sized for fast-loading rather than for
inferential quality.

The three sampler-health failures the fixture exhibits are
therefore expected and informative:

- `rhat` $\approx 1.20$, well above the $1.01$ threshold.
- `ess_bulk` $\approx 12$, two orders of magnitude below $400$.
- `ess_tail` $\approx 105$, well below $400$.

The two pathology counters (`divergences` and `max_treedepth`)
are both zero: the sampler is healthy in the local-geometry
sense, but the chain is too short and too solitary to support
convergence or mixing conclusions.

For substantive analyses the package's verification ledger
(documented in **M6**) treats any one of the three failures the
fixture exhibits as a release blocker. The recommended response
is to re-fit with the package's default configuration (four
chains, $1000$ warmup, $3000$ sampling) and re-check; persistent
failures after that are model-diagnostic rather than
sampler-diagnostic and call for a closer look at the data, the
grid recipe (see **A4**), or the spline degrees of freedom $M$.

## A diagnostic checklist

The bullets below condense the conventional thresholds into a
release checklist. A fit that ticks every box can be released
without further investigation; a fit that fails any box requires
the corresponding remediation before its summaries are acted on.

- [ ] All per-parameter $\widehat{R}$ below $1.01$
  (`max(diag$rhat) < 1.01`).
- [ ] All per-parameter bulk ESS above $400$
  (`min(diag$ess_bulk) > 400`).
- [ ] All per-parameter tail ESS above $400$
  (`min(diag$ess_tail) > 400`).
- [ ] Zero divergent transitions
  (`sum(diag$divergences) == 0`).
- [ ] Zero max-treedepth saturations
  (`sum(diag$max_treedepth) == 0`).
- [ ] Runtime within the expected band for the chosen
  configuration (`diag$runtime_seconds`).
- [ ] Both bookkeeping flags empty
  (`length(diag$diagnostic_skipped) == 0` and
  `length(diag$sampler_diagnostics_failed) == 0`).

## What's next

- **A6 · Plotting `bayesEfron` fits** — the four-view plot
  surface, including the caterpillar plot used in **A1**, the
  prior-$g$ density, the sensitivity overlay, and the diagnostic
  bar chart shown above.
- **M5 · The Stan model** — the methodological vignette
  documenting the locked Stan source, the
  log-spline-deconvolution likelihood, and the parameterisation
  choices that determine the sampler's geometry.
- **M6 · Verification and calibration** — the verification
  ledger that fixes the release thresholds used in the checklist
  above and documents the calibration runs the package's
  reference fits were validated against.
