---
title: "A6 · Plotting and visualization"
description: "Four diagnostic-and-summary views of a fitted bayesEfron model."
output:
  rmarkdown::html_vignette:
    fig_width: 7
    fig_height: 4.5
vignette: >
  %\VignetteIndexEntry{A6 · Plotting and visualization}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  echo     = TRUE,
  collapse = TRUE,
  comment  = "#>",
  message  = FALSE,
  warning  = FALSE,
  fig.width  = 7,
  fig.height = 4.5
)
library(bayesEfron)
has_ggplot2 <- requireNamespace("ggplot2", quietly = TRUE)
if (has_ggplot2) {
  library(ggplot2)
}

fit <- readRDS(system.file(
  "examples", "cached_fit_re_smoke.rds",
  package = "bayesEfron"
))
```

## Four views, four questions

`plot.bef_fit_re()` is the package's graphical surface. A single
S3 method dispatches on a `type` argument to one of four views,
each tuned to a question a meta-analyst routinely asks of a
deconvolution fit:

| `type`          | Question answered                                                       |
|:----------------|:------------------------------------------------------------------------|
| `"caterpillar"` | Where do the per-site posterior means sit relative to one another, and how wide are their credible intervals? |
| `"g"`           | What does the underlying mixing distribution $g(\theta)$ look like, with uncertainty? |
| `"sensitivity"` | How much does the per-site interval width move when the credible level is changed? |
| `"diagnostic"`  | Is the sampler healthy enough for the headline summaries to be trusted? |

The four views read from the same fit object — there is no
re-fitting and no auxiliary data load. The caterpillar and
sensitivity views consume the per-site posterior draws stored in
`fit$metadata$theta_rep_draws`; the prior-summary view reads the
discretised $g$ draws in `fit$draws`; the diagnostic view consumes
the same payload that `diagnose()` and `summary()` consume,
producing a bar-chart triage view of the sampler-health fields.

The cached five-site smoke fit loaded in the setup chunk is the
fixture used throughout. It was sampled deliberately small (one
chain, 100 sampling iterations) and so exhibits diagnostic
failures that the `type = "diagnostic"` view makes immediately
visible; see **A5** for the numeric thresholds and remediation
recipe. All four chunks below render against this fixture, so the
visual output a reader sees is exactly what the documented call
produces.

## `type = "caterpillar"`

The caterpillar plot is the canonical per-site posterior view.
For each site $i$ the plot draws a horizontal segment spanning the
credible interval at the requested `level`, an inner heavier
segment spanning the central 50% interval (the inter-quartile
range from the posterior draws), and a filled point at the
posterior mean. A dashed vertical reference line marks the
posterior mean of $g$ — the population-level prior mean. The plot
shows posterior means and intervals plus the population reference
line; raw $\hat\theta_i$ values are not drawn in the v0.1 plot
method.

```{r caterpillar-default, fig.alt = "Caterpillar plot of per-site posterior latent effects (90 percent credible intervals) from the cached five-site smoke fit, sites sorted by posterior mean."}
plot(fit, type = "caterpillar")
```

The `sort_by` argument controls the row ordering. The default,
`sort_by = "mean"`, orders sites by posterior mean (ascending from
bottom). `sort_by = "sigma"` orders by within-study standard error
$\sigma_i$, which makes the relationship between $\sigma_i$ and
interval width directly readable: sites with the largest
$\sigma_i$ have the longest segments. `sort_by = "none"` preserves
input order, which is useful when sites carry an analyst-meaningful
identifier (a state code, a cohort year) that the analyst wants on
the vertical axis in source order.

```{r caterpillar-sigma, fig.alt = "Caterpillar plot of per-site posterior latent effects (95 percent credible intervals) from the cached five-site smoke fit, sites sorted by within-study standard error sigma."}
plot(fit, type = "caterpillar", sort_by = "sigma", level = 0.95)
```

The `level` argument controls the outer interval; the inner
heavier segment is fixed at the 25th–75th percentile range. The
plot above uses `level = 0.95`, widening the outer segments
relative to the default 90% intervals shown in the first plot.

## `type = "g"`

The prior-summary view shows the posterior estimate of the mixing
distribution $g(\theta)$ — the population distribution of latent
site effects. When the draws object carries the discretised $g$
fields (`g[1]` through `g[L]`, with $L$ the grid resolution), the
view draws the posterior mean of $g$ on the grid as a solid line
with a credible band at the requested `level`. When those fields
are not stored, the view falls back to a marginal interval plot of
the prior moments through `confint(x, type = "g")`.

```{r g-view, fig.alt = "Prior-summary view of the cached fit showing the posterior-mean estimate of the mixing distribution g(theta) with uncertainty bands."}
plot(fit, type = "g")
```

For tabular use the same population-level summary lives in
`fit$metadata`: `mean_g_summary` and `var_g_summary` are named
fields, while `sd_g_summary` is stored as a durable attribute of
`fit$metadata`. All three carry the posterior mean, standard
deviation, and conventional quantile summaries of the prior mean,
prior variance, and prior standard deviation respectively. The
dashed reference line in the caterpillar view above is the
`mean_g_summary$mean` entry. **A2** documents the metadata block in
full.

## `type = "sensitivity"`

The sensitivity view is a caterpillar plot at a non-default
credible level, rendered with the title "bayesEfron sensitivity"
to distinguish it from the standard caterpillar. Side-by-side
inspection of the two views answers a single question: how much
does the per-site interval width move as the credible level moves?
For posteriors that are approximately Gaussian on the latent
scale, the move from 90% to 95% widens each interval by a factor
of roughly $1.96/1.645 \approx 1.19$. Departures from that ratio
are diagnostic of either tail asymmetry or sparse posterior
sampling in the interval endpoints.

```{r sensitivity-default, fig.alt = "Sensitivity view of the cached fit at default 90 percent credible level: per-site posterior intervals with inner inter-quartile segments and outer 90 percent segments."}
plot(fit, type = "sensitivity")
```

```{r sensitivity-95, fig.alt = "Sensitivity view of the cached fit at the 95 percent credible level: same posterior means and inner segments as the default view, with widened outer segments."}
plot(fit, type = "sensitivity", level = 0.95)
```

The two plots share the same posterior means and inner
inter-quartile segments; only the outer segments move. The cached
fixture's intervals are very wide in absolute terms — a consequence
of the smoke-scale iteration count rather than the credible-level
choice — so the relative widening is the comparison to focus on.

## `type = "diagnostic"`

The diagnostic view is a bar chart of the five sampler-health
metrics that `diagnose(fit)` and `summary(fit)$diagnostics`
produce in numeric form: $\widehat{R}$, bulk ESS, tail ESS,
divergence count, and max-treedepth-saturation count. It is a
companion to the numeric view rather than a replacement for it;
**A5** walks through each metric, its release threshold, and the
remediation recipe when the threshold is missed.

```{r diagnostic-view, fig.height = 4, fig.alt = "Diagnostic bar chart of the cached fit showing Rhat, bulk ESS, tail ESS, divergences, and max-treedepth; ESS bars dominate the value axis at smoke scale."}
plot(fit, type = "diagnostic")
```

On the cached fixture the bulk and tail ESS bars dominate the
vertical scale — the visual signature of a smoke-scale fit where
ESS is the information-bearing diagnostic and the one that scales
fastest with iteration count.

## Backends

`plot.bef_fit_re()` ships two backends. The `ggplot2` backend is
used when `ggplot2` is installed and the environment variable
`BAYESEFRON_NO_GGPLOT2` is not exactly `"1"`; it returns a
`ggplot2::ggplot` object visibly, so the object can be assigned
and extended with layered grammar. The base-graphics backend is
used otherwise; it draws on the active graphics device and returns
`invisible(NULL)`. The two backends produce visually similar
output but are not pixel-identical: the base backend uses
`graphics::segments()` and `graphics::points()` with package
defaults, the `ggplot2` backend layers `geom_segment()`,
`geom_point()`, and `geom_vline()` on a `theme_minimal()` base.

Forcing the base backend is a one-line operation:

```{r toggle-base, eval = FALSE}
Sys.setenv(BAYESEFRON_NO_GGPLOT2 = "1")
plot(fit, type = "caterpillar")
Sys.unsetenv("BAYESEFRON_NO_GGPLOT2")
```

Downstream code that chains `ggplot2` layers should guard against
the base path with `inherits(p, "ggplot")`. The guard is cheap and
keeps the calling code working in either backend regime:

```{r ggplot-chain, eval = has_ggplot2, fig.alt = "Caterpillar plot of the cached fit rendered through the ggplot2 backend with theme_minimal and added title, subtitle, and caption, demonstrating that the plot.bef_fit_re ggplot output can be customised with standard ggplot2 layers."}
p <- plot(fit, type = "caterpillar")
if (inherits(p, "ggplot")) {
  p +
    theme_minimal() +
    labs(
      title    = "Per-site posterior intervals",
      subtitle = "Cached five-site smoke fit, 90% credible intervals",
      caption  = "Dashed line is the posterior mean of g."
    )
}
```

The chained call returns a fully customised `ggplot` object that
behaves like any other; further `+ scale_*()`, `+ facet_*()`, and
`+ coord_*()` layers compose as expected. The `attr(p,
"bef_plot_payload")` attribute carries the data the view consumed,
which is occasionally useful when extending the view with a layer
that needs the same numeric inputs.

## What's next

- **A7 · Case study — Lee–Sui replication light** — an end-to-end worked
  analysis on a real multisite dataset, exercising the four views
  in their substantive roles.
- **M2 · Efron's log-spline prior** — the methodological vignette
  documenting the prior $g(\theta)$ that the `type = "g"` view
  visualises, including the role of the spline degrees of freedom
  $M$ and the discretisation grid $L$.
- **M3 · Bayesian random-effects hierarchy** — the formal write-up of the
  $\theta_i \mid g$ / $\hat\theta_i \mid \theta_i, \sigma_i$
  hierarchy whose two layers the caterpillar and prior-$g$ views
  display.
