---
title: "A7 · Case study — Lee–Sui replication light"
description: "One K=50 replication of the Lee–Sui benchmark, end to end."
output:
  rmarkdown::html_vignette:
    fig_width: 7
    fig_height: 4.5
vignette: >
  %\VignetteIndexEntry{A7 · Case study — Lee–Sui replication light}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  echo     = TRUE,
  collapse = TRUE,
  comment  = "#>",
  message  = FALSE,
  warning  = FALSE,
  fig.width  = 7,
  fig.height = 4.5
)
library(bayesEfron)
has_ggplot2 <- requireNamespace("ggplot2", quietly = TRUE)

fit <- readRDS(system.file(
  "examples", "cached_fit_re_smoke.rds",
  package = "bayesEfron"
))
```

## Background — the Lee–Sui benchmark

The Lee–Sui benchmark is the v0.1 release-blocking calibration test
for the random-effects deconvolution fit. It targets a single,
narrow question: do the posterior `theta_rep` 90% credible
intervals contain the latent site effects at the nominal rate? The
test is necessary because the package's headline summaries are
posterior intervals on per-site effects $\theta_i$, and a user
relying on a 90% interval needs evidence that the long-run
coverage rate is close to 90% under known data-generating
conditions.

The benchmark holds the data-generating mechanism fixed and varies
the number of sites $K$ across five values: $K \in \{50, 100, 200,
500, 1500\}$. Each $K$ defines a scenario; within a scenario the
benchmark draws 20 independent replication datasets from the
locked Lee–Sui twin-towers generator and fits the random-effects
model to each. For each replication the test computes the per-site
coverage rate of the nominal 90% `theta_rep` interval — the
fraction of sites whose `theta_true` falls inside the interval —
and the headline statistic is the mean of those 20 replication-level
rates. The acceptance band is `[0.87, 0.92]`. The band is two-sided
on purpose: aggregate coverage outside that band — either too
permissive or too anti-conservative — falls outside the release
gate.

The v0.1.0 release ledger passed at all five $K$ values. The
repaired $K=1500$ result, the last to clear, recorded an aggregate
coverage of 0.876433333333333 against the same `[0.87, 0.92]`
band; the four smaller-$K$ scenarios cleared with more headroom. The five passing
rows are recorded in `tests/testthat/verification_targets.csv` as
`tier3_theta_rep_coverage_paper_realdata_K50` through
`tier3_theta_rep_coverage_paper_realdata_K1500`. The Tier 3 evidence
is benchmark-scoped: it underwrites calibration of `theta_rep`
intervals on the locked Lee–Sui all-20-replication fixtures and the
default fit configuration, not a general-purpose calibration claim.
Sampler diagnostic flags for `rhat`, `ess_bulk`, and `ess_tail` are
preserved as calibration-tier metadata on the accepted fits.

The benchmark's choice of five $K$ values is deliberate. The two
smallest values, $K=50$ and $K=100$, exercise the small-$K$ regime
in which discrete-grid deconvolution is most sensitive to grid and
prior choices and in which sampler-level finite-sample noise has
the largest effect on per-site intervals. The mid-range values,
$K=200$ and $K=500$, exercise the regime in which most applied
multisite analyses operate. The largest value, $K=1500$, exercises
the regime in which the underlying mixing distribution is densely
sampled and the model's ability to recover the shape of $g(\theta)$
becomes a leading constraint on coverage. Reading the five rows
together — rather than any single row in isolation — is the
intended use of the release evidence.

This vignette walks through one $K=50$ replication end to end. It
stops short of the full Tier 3 release-evidence matrix — that
matrix, with all five $K$ values and all 20 replications per $K$,
is the subject of M6.

## The K=50 fixture

The locked $K=50$ fixture lives at
`tests/testthat/_fixtures/lee_sui_K50.rds` and is the smallest of
the five Tier 3 fixtures by site count. The fixture is bundled
with the package source tree under `tests/testthat/_fixtures/`,
not under `inst/`, and is therefore not accessible through
`system.file()` to a vignette built outside the source tree. The
present vignette describes the fixture's structure from prior
inspection and uses a smaller cached fit for the rendered output;
the next section explains the stand-in framing.

The fixture is a list of 16 named elements. The top-level scalars
record the scenario: `K = 50`, the primary `replication` index,
the integer `seed`, and the `grid_method = "paper_realdata"` grid
recipe with the locked $L = 51$ grid length and $M = 6$ spline
degrees of freedom. The primary replication's per-site arrays are
`theta_hat`, `sigma`, and `theta_true`, each of length 50, along
with the integer `source_row_id` and the `tower` factor that
records which side of the Lee–Sui twin-towers density each site
was drawn from.

A `replications` slot carries the full all-20-replication payload:
20 sub-lists, each with the same 15-field replication schema as the
primary entry (`replication`, `dataset_id`, `seed`, `K`,
`theta_hat`, `sigma`, `theta_true`, `source_row_id`, `tower`, `L`,
`M`, `grid_method`, `expansion`, `bound_expansion`, and the
per-replication `grid` metadata). The 20 datasets share the
generator and the locked grid configuration but use distinct
sampling seeds, so each replication is a genuine independent draw
from the data-generating mechanism.

The fixture's `metadata$coverage` block records the protocol the
release ledger enforces: the nominal level is 0.9, the aggregation
is the mean of replication-level site-coverage rates over 20
replications, the per-replication denominator is the per-site
posterior interval count, failed replications are not silently
dropped, replication identity is keyed by unique input-payload
hashes, and sampler seeding follows the Tier 3 sampler-seed policy.
Each of these six policy fields binds a choice that, left implicit,
could move the headline aggregate by an amount comparable to the
half-width of the acceptance band; recording them on the fixture
makes the release decision rule auditable.

The fixture's `metadata$source` block records the lineage of the
underlying scenario, including the public data-generation script,
the public Stan model file, the public small-$K$ pipeline, the
original development script, and the archived appendix result that
serves as the canonical input for the $K$ vector. Each artifact is
recorded with a path, a relevant-lines range, and a SHA-256 digest,
so the provenance of the generated `theta_hat`, `sigma`, and
`theta_true` arrays is reconstructible from the fixture alone.

## Loading the cached fit (stand-in)

The vignette uses the five-site cached smoke fit shipped under
`inst/examples/cached_fit_re_smoke.rds` for all chunks that produce
posterior summary or plot output. The reason is purely a
renderability constraint: the locked $K=50$ fixture lives in the
test-fixtures tree and is not on the install path that a vignette
sees at build time. The cached smoke fixture is a real
`bef_fit_re` object produced by the package's eight-stage fit
pipeline on a five-site toy dataset, so it carries the same
metadata schema, the same generated-quantity payload, and the same
S3 method behaviour as a $K=50$ Lee–Sui fit. The differences are
the site count, the iteration budget (one chain, 100 sampling
iterations), and the data-generating mechanism. A $K=50$ Lee–Sui
fit would expose output of the same shape, scaled to 50 rows
instead of 5 in the per-site tables and to 4 chains with the full
sampling budget in the diagnostic block.

```{r load-fit, eval = FALSE}
fit <- readRDS(system.file(
  "examples", "cached_fit_re_smoke.rds",
  package = "bayesEfron"
))
```

## Posterior summaries on the cached stand-in

The `summary()` method returns a `summary.bef_fit` object that
collects the three population-level summaries of the prior
$g(\theta)$, the sampler-health diagnostic block, the runtime, the
locked Stan SHA-256, and the per-site posterior summary table.

```{r summary-fit, eval = TRUE}
summary(fit)
```

The per-site theta table here has five rows. A genuine $K=50$
Lee–Sui fit would expose the same column set — `site`, `mean`,
`sd`, `lower`, `upper`, `map` — with 50 rows; the calibration test
consumes the `lower` and `upper` columns of that 50-row table
against the held-out `theta_true` vector to compute the replication's
per-site coverage rate.

The `confint()` method returns the same per-site interval table as a
plain data frame, which is the form the calibration test consumes:

```{r confint-theta, eval = TRUE}
confint(fit, type = "theta")
```

On the $K=50$ fixture this call would return a 50-row data frame.
The benchmark drives the same call against the held-out
`theta_true` vector for the replication, counts the fraction of
rows for which `lower <= theta_true <= upper` holds, and records
that fraction as the replication's per-site coverage rate. The
aggregate over 20 replications is the headline statistic for the
$K=50$ row of the release ledger.

## Coverage interpretation

The Tier 3 coverage statistic is built in two stages. Within a
single replication, the posterior fit produces a posterior
distribution over $\theta_i$ for each of the $K$ sites; the
post-processor reduces those draws to a 90% interval per site, and
the per-site coverage rate is the fraction of the $K$ sites for
which the held-out `theta_true_i` lies inside the interval. Across
replications, the aggregate coverage statistic is the mean of those
20 per-replication rates, weighted equally. The mean-of-rates
aggregation, rather than a pooled fraction over all $20 \times K$
sites, is the locked rule recorded under
`metadata$coverage$aggregation` in the fixture and matches the
release-ledger convention. Failed replications are not silently
dropped: a replication that fails to fit is recorded as a failure
and the release decision is made on the population of 20 replications
that the pipeline actually accepted.

The pre-specified acceptance band is `[0.87, 0.92]`. The band is
two-sided. The lower bound, 0.87, gives the model a small but
explicit budget for under-coverage: posterior intervals that are
slightly narrow under the data-generating mechanism remain in band
provided the deficit is bounded. The upper bound, 0.92, denies the
model credit for over-conservatism: a release that buys nominal
coverage by inflating intervals well beyond 90% is also out of
band.

At $K=50$ the band is wider than the worst-case sampling noise
under the protocol. With 20 replications and a per-replication
denominator of $K=50$ sites, simulation noise on the aggregate is
small relative to the half-width of the band, so a band excursion
at $K=50$ is informative about calibration rather than about
sampling variability. The release ledger records that the $K=50$
scenario cleared the band with healthy headroom, as did $K=100$,
$K=200$, and $K=500$; the repaired $K=1500$ result, 0.876, cleared
the lower bound with the slimmest margin of the five and is the
documented worst-case scenario for the v0.1.0 release.

The Tier 3 evidence is intentionally benchmark-scoped: it
underwrites the calibration claim on the locked Lee–Sui
all-20-replication fixtures under the default fit configuration. It
does not underwrite a generic frequentist coverage claim for
arbitrary data-generating mechanisms; users analysing data whose
mixing distribution differs materially from the Lee–Sui twin-towers
generator should treat the band as suggestive rather than
guaranteed and consider running a scenario-specific calibration
test of their own.

## Diagnostics on the case-study fit

The `diagnose()` method returns the validated `bef_diagnostic`
payload that the post-processor stored on the fit object. Running
it on the cached stand-in surfaces the small-fit signature
deliberately:

```{r diagnose-fit, eval = TRUE}
diagnose(fit)
```

The cached fit was sampled with one chain and 100 sampling
iterations and so trips the conventional thresholds on $\widehat{R}$
and bulk ESS; A5 documents those thresholds and the remediation
recipe. A real $K=50$ Lee–Sui fit uses four chains, 1000 warmup
iterations, and 3000 sampling iterations under the default fit
configuration; that envelope makes diagnostics reviewable on the
post-warmup draws. Sampler-flag metadata for `rhat`, `ess_bulk`,
and `ess_tail` that survives the larger-$K$ refits is preserved as
calibration-tier metadata and does not unilaterally invalidate the
Tier 3 coverage acceptance: the release ledger treats the
aggregate coverage and the diagnostic flags as separable evidence
streams.

## Caterpillar plot of latent effects

The caterpillar view is the canonical per-site posterior summary.
Each row shows the credible interval for one site, with a heavier
inner segment at the inter-quartile range and a filled point at the
posterior mean; a dashed vertical reference line marks the
posterior mean of $g$.

```{r caterpillar, eval = has_ggplot2, fig.alt = "Caterpillar plot of per-site posterior latent effects with 90 percent credible intervals for one K=50 replication of the Lee-Sui benchmark from the cached fit."}
plot(fit, type = "caterpillar")
```

On the cached five-site stand-in the plot has five rows. A $K=50$
Lee–Sui fit would produce a 50-row caterpillar; the v0.1 plot
method draws posterior means and credible intervals for each row
but does not currently overlay the held-out `theta_true` marks.
For per-site coverage the workflow consumes the numeric
`theta_rep_draws` rather than the plot. The aggregate coverage
statistic is the count of in-coverage rows divided by 50, averaged
across the 20 replications of the scenario. A6 documents the four S3 plot views
in full, including the prior-summary, sensitivity, and diagnostic
views that complement the caterpillar.

## What's next

- **M3 · Bayesian random-effects hierarchy** — the mathematical write-up of
  the $\theta_i \mid g$ / $\hat\theta_i \mid \theta_i, \sigma_i$
  hierarchy that the Lee–Sui benchmark stress-tests.
- **M5 · The Stan model and the cache** — the locked Stan source,
  the eight-stage fit pipeline, and the cache contract behind the
  $K=50$ fit object whose schema this vignette traces.
- **M6 · Verification and calibration** — the full Tier 3 release
  evidence matrix across all five $K$ values and all 20
  replications per $K$, including the repaired $K=1500$ row that
  closed the release ledger.
