---
title: "Getting started with bayesEfron"
description: "A first fit: the distribution of effects in 19 teacher expectancy studies and the effect of each study."
output:
  rmarkdown::html_vignette:
    fig_width: 7
    fig_height: 4.5
bibliography: references.bib
link-citations: true
vignette: >
  %\VignetteIndexEntry{Getting started with bayesEfron}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
library(bayesEfron)
options(bayesEfron.use_cli = FALSE, width = 80)
num <- function(x, digits = 2) formatC(x, format = "f", digits = digits)
```

@raudenbush1985 analyzed 19 studies of the effect of teachers' expectations
on their pupils' IQ scores. In each study teachers were led to expect that
certain pupils, in fact chosen at random, would show unusual intellectual
growth. Each study gives an estimate of the effect of that expectation, as
a standardized mean difference, and a standard error. The studies differ a
great deal in size, so their standard errors differ too.

A meta-analysis of such studies usually asks first what the average effect
is. Often a second question is of more interest: how the true effects are
distributed over studies, and what the effect in each study is once the
other studies are taken into account. The random-effects model answers the
second question by assuming that the true effects follow a normal
distribution [@raudenbush1985]. For these studies, however, there is reason
to doubt that assumption. @raudenbush1984 argued that an expectation that
the experimenter tries to induce carries less conviction with a teacher who
already knows the pupils, and that the effect should therefore be found
mainly in the studies where teacher and pupils had met only recently. If a
minority of the studies have sizable effects and the others have effects
near zero, the distribution of true effects is skewed.

bayesEfron fits the same two-level model as a random-effects
meta-analysis, but it does not take the distribution of true effects,
called $g$ below, to be normal: $g$ may have any smooth shape that a small
number of spline functions can describe [@efron2016; @lee2025]. This
vignette fits the model to the 19 studies and interprets the two things
that the fit returns: the estimated $g$, and the effect of each study.

## The studies

The data are in the package as `raudenbush1985`.

```{r data}
transform(
  raudenbush1985[, c("author", "year", "weeks", "yi", "sei")],
  sei = round(sei, 3)
)
```

`yi` is the standardized mean difference and `sei` its standard error.
`weeks` is the number of weeks for which the teacher had known the pupils
when the expectation was induced.

```{r by-weeks}
recent <- raudenbush1985$weeks <= 1
contact <- ifelse(recent, "at most 1 week", "2 weeks or more")
tapply(raudenbush1985$yi, contact, mean)
```

The estimates average `r num(mean(raudenbush1985$yi[recent]))` in the
`r sum(recent)` studies with at most a week of prior contact and
`r num(mean(raudenbush1985$yi[!recent]))` in the other `r sum(!recent)`,
which is the pattern that Raudenbush expected. The model below has no
covariates and does not use `weeks`. It is given only the estimates and
their standard errors, and the question is what it makes of the
distribution of effects from those alone.

`as_bef_data()` takes the two columns out of the data frame and checks
them. The package calls the units sites, since they may equally be the
sites of a multisite trial; here each site is one study. There must be at
least five of them, with no missing values and with positive standard
errors. Other forms of input are described in `vignette("data-input")`.

```{r as-bef-data}
dat <- as_bef_data(raudenbush1985)
dat
```

In the printed object and in what follows, `theta_hat` holds the estimates
and `sigma` their standard errors.

## Fitting the model

Fitting needs the cmdstanr package, which is not on CRAN, and CmdStan,
which needs a C++ compiler. Both are installed once.

```{r install, eval = FALSE}
install.packages("cmdstanr",
                 repos = c("https://stan-dev.r-universe.dev", getOption("repos")))
cmdstanr::install_cmdstan()
```

The model is then fitted with one call.

```{r fit-call, eval = FALSE}
fit <- bayes_efron_fit(theta_hat = dat$theta_hat, sigma = dat$sigma,
                       seed = 1985)
```

The model is fitted by Markov chain Monte Carlo, which produces a sample of
values, called draws, from the posterior distribution of every quantity in
the model; the estimates and intervals below are computed from the draws.
By default four chains are run, each with 1,000 warmup iterations, which
are discarded, and 3,000 sampling iterations, which are kept. The first
call compiles the Stan model, which can take up to a minute. For these data
the sampling then took `r num(raudenbush_fit$metadata$runtime_seconds, 1)`
seconds.

The model represents $g$ by the probabilities that it gives to 101 equally
spaced values of the effect, called the grid. The default grid was used
here; `vignette("choosing-a-grid")` discusses it, and `vignette("model")`
sets out the model.

So that this vignette can be built without CmdStan, the result of the call
above is stored in the package as `raudenbush_fit`, which keeps one
posterior draw in fifteen.

```{r fit}
fit <- raudenbush_fit
fit
```

The line of diagnostics gives the worst value of each convergence
diagnostic over all the quantities in the model. R-hat and the effective
sample sizes are satisfactory and there is no divergent transition. The
warning says that an iteration reached the maximum tree depth, which
concerns the efficiency of the sampler and not the validity of the draws.
`vignette("diagnostics")` explains the diagnostics and what to do when one
of them fails.

## The distribution of effects

```{r g-summary}
confint(fit, type = "g")
```

```{r g-numbers, include = FALSE}
g_ci <- confint(fit, type = "g")
rownames(g_ci) <- g_ci$site
grid <- fit$metadata$data_list$grid
g_mean <- colMeans(exp(posterior::as_draws_matrix(fit)[, sprintf("log_g[%d]", seq_along(grid))]))
peak <- grid[which.max(g_mean)]
```

The rows are the mean, the variance and the standard deviation of $g$, each
with its posterior mean (`point`) and a 90% posterior interval, which runs
from the 5% to the 95% quantile of the draws. Such an interval is often
called a credible interval; `level = 0.95` gives a 95% interval, the level
that metafor uses by default for its confidence intervals. The mean of $g$
is the average effect over studies of this kind. It is estimated as
`r num(g_ci["mean_g", "point"])`, with an
interval from `r num(g_ci["mean_g", "lower"])` to
`r num(g_ci["mean_g", "upper"])`. The standard deviation of $g$ is
estimated as `r num(g_ci["sd_g", "point"])`, with an interval from
`r num(g_ci["sd_g", "lower"])` to `r num(g_ci["sd_g", "upper"])`.

```{r g-plot, fig.alt = "The estimated distribution of effects on the grid, with a 90 percent band."}
plot(fit, type = "g")
```

The line is the posterior mean of the probability that $g$ gives to each
point of the grid, and the band is a pointwise 90% interval. The estimated
distribution is concentrated near zero and has a long tail to the right.
About `r round(100 * sum(g_mean[abs(grid) <= 0.2]))`% of its probability
lies between $-0.2$ and $0.2$, and about
`r round(100 * sum(g_mean[grid > 0.5]))`% lies above $0.5$. It is not
symmetric about its peak at `r num(peak)`:
`r round(100 * sum(g_mean[grid > peak + 0.4]))`% of the probability lies
more than 0.4 above the peak and
`r round(100 * sum(g_mean[grid < peak - 0.4]))`% more than 0.4 below it.

The model was not told which studies had recent contact. The posterior
mean of $g$ nevertheless puts most effects near zero and a few well above
it, although the band shows how little 19 studies say about the right tail.
The curve also turns up slightly at the two ends of the grid, where there
are no estimates. The data hardly bear on $g$ there, so the band is wide
and the shape of the curve reflects the prior and the spline functions.

`predict()` turns the estimated $g$ into statements about a study that has
not been done. It draws the effect of a new study from $g$, with a
different posterior draw of $g$ each time, so that the uncertainty about
$g$ is part of the result.

```{r predict}
theta_new <- predict(fit, seed = 1)
quantile(theta_new, c(0.05, 0.25, 0.5, 0.75, 0.95))
mean(theta_new > 0.2)
```

The last number estimates the probability that the true effect in a new
study of this kind exceeds 0.2.

## The effect in each study

`coef()` returns the posterior mean of each study's effect, and `confint()`
returns a 90% interval for it.

```{r sites}
sites <- cbind(
  raudenbush1985[, c("author", "weeks", "yi", "sei")],
  posterior_mean = coef(fit),
  confint(fit)[, c("lower", "upper")]
)
sites$sei <- round(sites$sei, 3)
sites[c("posterior_mean", "lower", "upper")] <-
  round(sites[c("posterior_mean", "lower", "upper")], 2)
sites
```

```{r site-numbers, include = FALSE}
positive <- which(sites$lower > 0)
largest <- which.max(sites$yi)
naive <- sites$yi[largest] + c(-1, 1) * qnorm(0.95) * raudenbush1985$sei[largest]
theta_largest <- as.numeric(
  posterior::as_draws_matrix(fit)[, sprintf("theta_rep[%d]", largest)]
)
grid_comparison <- readRDS(system.file(
  "examples", "grid_comparison_example.rds", package = "bayesEfron"
))
sensitivity <- grid_comparison$raudenbush_sensitivity
study4_range <- range(sensitivity$study4_mean)
# The sentences below are about these data; stop if they no longer hold.
stopifnot(
  all(abs(coef(fit) - peak) <= abs(raudenbush1985$yi - peak) + 0.005),
  all(raudenbush1985$weeks[positive] <= 1),
  largest == 4,
  diff(unlist(sites[largest, c("lower", "upper")])) > diff(naive),
  # nine specifications, none as much as two standard errors behind the best
  nrow(sensitivity) == 9,
  all(abs(sensitivity$elpd_diff) < 2 * sensitivity$se_diff |
        sensitivity$se_diff == 0)
)
```

Each posterior mean is drawn from the study's own estimate toward the
values where $g$ puts most of its probability, here the peak just above
zero. How far depends on the standard error. Studies 5 and 17 have similar
estimates, `r num(sites$yi[5])` and `r num(sites$yi[17])`, but standard
errors of `r num(sites$sei[5], 3)` and `r num(sites$sei[17], 3)`, and their
posterior means are `r num(sites$posterior_mean[5])` and
`r num(sites$posterior_mean[17])`.

Because $g$ has a long right tail, a large positive estimate is not drawn
all the way back, and the interval excludes zero for `r length(positive)`
studies (`r paste(sites$author[positive], collapse = "; ")`). All of them
had at most a week of prior contact.

For study `r largest`, which has the largest estimate, the interval runs
from `r num(sites$lower[largest])` to `r num(sites$upper[largest])` and is
wider than the one that its estimate and standard error give by themselves,
`r num(naive[1])` to `r num(naive[2])`. With a normal $g$ of known mean and
variance that cannot happen. Here the model is undecided between a large
true effect and a small one that was overestimated: it gives probability
`r num(mean(theta_largest < 0.3))` to an effect below 0.3 and
`r num(mean(theta_largest > 0.8))` to one above 0.8. The result for this
study is also the one that depends most on the choice of the grid: over
the nine specifications compared in `vignette("choosing-a-grid")` its
posterior mean ranges from `r num(study4_range[1])` to
`r num(study4_range[2])`, and cross-validation does not tell those
specifications apart. A conclusion about a study as extreme as this one
should be checked against that comparison before it is reported.

The limits of the intervals are quantiles of posterior draws of the effect.
The draws take values on the grid, whose spacing here is
`r num(diff(grid)[1])`, so the limits should be read to no finer a
resolution than that. A limit that lies between two grid points is an
interpolation between two draws.

```{r caterpillar, fig.height = 5, fig.alt = "Posterior means and intervals of the 19 study effects."}
plot(fit, type = "caterpillar")
```

The points are the posterior means, the thin lines are the 90% intervals
and the thick lines are the central 50% intervals. The studies are numbered
as in the data, and the dashed line marks the mean of $g$.

## What the fitted object holds

`fit` is a list with three elements. `fit$draws` holds the posterior draws
of every quantity in the model, in the format of the posterior package.
`fit$posterior` holds the draws of the mean, variance and standard
deviation of $g$ and of two summaries of the fit, and `fit$metadata` holds
the data, the settings and summaries of the draws. `?bef_fit` describes
them and lists the methods. `summary()` prints the summaries that are used
most.

```{r summary}
summary(fit)
```

`Prior g` is the distribution of true effects, which is the prior
distribution of each study's effect. `Effective params` and
`Log marginal lik.` are explained in `vignette("model")`. In the table of
study effects, `sd` is the posterior standard deviation, `lower` and
`upper` are the limits of the 90% interval and `map` is the posterior mode,
the point of the grid with the largest posterior probability. Where the
posterior distribution is skewed the mode and the mean are far apart. For
study `r largest` the mode is `r num(coef(fit, type = "map")[largest])` and
the mean `r num(coef(fit)[largest])`, which is one more sign that the model
is undecided about that study.

Any quantity in the model can be summarized from the draws with the
functions of the posterior package.

```{r draws}
posterior::summarize_draws(
  posterior::subset_draws(posterior::as_draws(fit),
                          variable = c("mean_g", "sd_g", "lambda"))
)
```

## References
