---
title: "Choosing the grid and comparing models"
description: "What the grid and the spline basis do, and how to check whether the results depend on them."
output:
  rmarkdown::html_vignette:
    fig_width: 7
    fig_height: 4.5
bibliography: references.bib
link-citations: true
vignette: >
  %\VignetteIndexEntry{Choosing the grid and comparing models}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
library(bayesEfron)
options(bayesEfron.use_cli = FALSE, width = 80)
has_loo <- requireNamespace("loo", quietly = TRUE)
num <- function(x, digits = 2) formatC(x, format = "f", digits = digits)
stored <- readRDS(system.file("examples", "grid_comparison_example.rds",
                              package = "bayesEfron"))
```

The model represents $g$, the distribution of true effects, by the
probabilities that it gives to the points of a grid, and it describes the
logarithms of those probabilities by a small number of spline functions.
Three things are therefore chosen before the model is fitted: where the
grid begins and ends, how many points it has, and how many spline
functions describe the shape of $g$. The defaults follow the application
in @lee2025 [Section 7.2] and its replication code: a grid of 101 points
that extends beyond the estimates by half their range on each side, and
six spline functions. For the teacher expectancy studies the comparisons
below do not separate the choices, although the result for the study with
the largest estimate depends on them. For data whose true effects fall
into two groups, the number of spline functions matters.

## Where the grid begins and ends

A true effect is taken to lie on the grid. If the grid is too narrow, an
effect outside it cannot be represented, and the posterior distribution of
that site is pressed against the end of the grid. If it is much too wide,
part of it covers values that no site can plausibly have. The knots of the
spline functions are spread evenly over the grid, so a wider grid also
leaves fewer of them where the effects are. The default rule widens the
range of the estimates on each side by `expansion` times its width.

```{r grid}
grid <- make_efron_grid(raudenbush1985$yi, raudenbush1985$sei)
range(raudenbush1985$yi)
range(grid$grid)
```

With the default `expansion = 0.5` the grid is twice as wide as the range
of the estimates.

```{r expansion}
ends <- sapply(c(0.25, 0.5, 1), function(expansion) {
  grid <- make_efron_grid(raudenbush1985$yi, raudenbush1985$sei,
                          expansion = expansion)
  range(grid$grid)
})
colnames(ends) <- paste("expansion", c(0.25, 0.5, 1))
ends
```

An estimate with a large standard error can lie far from its true effect,
so the estimates usually spread more widely than the true effects do. Room
beyond the estimates is needed all the same: the true effect of the site
with the largest estimate may be larger still, and a grid that ended at
that estimate would not let its posterior distribution say so.

@lee2025 [Appendix C] varied the width of the grid in a simulation with
1,500 sites, widening the range of the true effects by a quarter of its
width, by half and by its full width, and found that the accuracy of the
estimates and the coverage of the intervals changed little. The model was
fitted there with a variational approximation and not by Markov chain
Monte Carlo. The last section of this vignette looks at the same question
for 19 studies, where the answer is less comfortable.

`grid_method` selects other rules. `"paper_simulation"` and
`"paper_sensitivity"` place the grid around the true effects, which are
known only in a simulation; they are there so that the simulations of the
paper can be repeated, and they are of no use in the analysis of real data.
`"kl_target_experimental"` keeps the default end points and chooses the
number of points from the smallest standard error, by a rule taken from
the ebnm package [@willwerscheid2025]. In ebnm the rule bounds what is
lost when the distribution is approximated by a mixture of normal
components on a grid, provided that the standard errors are equal. The
bound has not been shown to hold for the model of this package, in which
the distribution consists of probabilities on the grid points and is
restricted to a spline family. Because the method uses at most 300 points,
the spacing can also be wider than the rule asks for. The method prints a
message that says so, and `?make_efron_grid` gives the formulas.

## How many points and how many spline functions

The number of grid points, `L`, sets the resolution. The posterior
distribution of an effect is a distribution on the grid, whose spacing is
`r num(diff(grid$grid)[1])` for the teacher expectancy studies with the
default of 101 points, and the limits of the intervals should be read to
no finer a resolution than that. @lee2025 [Appendix C]
compared 51, 101 and 201 points in the same simulation and found the
results nearly the same, although the computing time grew roughly in
proportion to the number of points.

The number of spline functions, `M`, sets how much shape $g$ can have. The
logarithm of the grid probabilities is a combination of `M` natural cubic
spline functions [@efron2016], which `make_efron_grid()` returns as the
columns of `B`.

```{r basis, fig.alt = "The six natural cubic spline functions on the default grid."}
matplot(grid$grid, grid$B, type = "l", lty = 1,
        xlab = "Effect", ylab = "Spline function")
```

With few functions $g$ has only broad features, although even four
functions allow two peaks. With more it can follow narrower ones, such as
a thin tail or two groups of effects that lie close together. The prior
distribution of the coefficients draws them toward zero by an amount that
is estimated with them (`vignette("model")` describes it), which guards
against more flexibility than the data support. It does not make `M` a
matter of indifference, however, as the second example below shows.

## How well one fit predicts

To compare choices you need a number that says how well a fitted model
describes the data. For each site and each posterior draw the model
computes the density of the estimate $\hat\theta_i$ given $g$,

$$
p(\hat\theta_i \mid \alpha) = \sum_{j=1}^{L} g_j(\alpha)\,
  \phi(\hat\theta_i;\, \tau_j, \sigma_i),
$$

where $\tau_j$ are the grid points, $g_j(\alpha)$ their probabilities and
$\phi(\cdot\,; \tau, \sigma)$ the normal density with mean $\tau$ and
standard deviation $\sigma$. The true effect of the site is summed out.
From these values `loo()` estimates how well the model would predict the
estimate of each site if that site had been left out of the fit
[@vehtari2017].

```{r loo, eval = has_loo}
fit_loo <- loo::loo(raudenbush_fit)
fit_loo
```

`elpd_loo` is the sum over sites of the log predictive densities. A larger
value is better, but the value means little by itself; it serves to compare
fits of the same data. `p_loo` is what the loo package calls the effective
number of parameters, and it is not the `effective_params` that
`summary()` prints for a fit. The method is an approximation, and the
Pareto k values say for each site whether it can be relied on. A large
value means that leaving the site out would change the posterior
distribution a good deal, so that the approximation is poor for that site.
@vehtari2017 take 0.7 as the limit, and recent versions of the loo package
lower it for a fit with few draws (to 0.66 for the 800 draws that
`raudenbush_fit` keeps). `plot()` shows the values by site.

```{r loo-plot, eval = has_loo, fig.height = 3.5, fig.alt = "Pareto k values of the 19 studies."}
plot(raudenbush_fit, type = "loo")
```

## Comparing specifications

Call a combination of `L`, `M` and `expansion` a specification. Each
specification is a different model for the same data.
`compare_efron_grids()` fits the model for every combination of the values
given for the three arguments and puts the results of `loo()` in one
table. With its defaults it compares 51, 101 and 201 grid points and four,
six and eight spline functions, nine fits in all.

```{r compare-call, eval = FALSE}
compare_efron_grids(raudenbush1985$yi, raudenbush1985$sei, seed = 1985)
```

The result of this call is stored in the package, together with the other
comparisons that are discussed below.

```{r compare-raudenbush}
stored <- readRDS(system.file("examples", "grid_comparison_example.rds",
                              package = "bayesEfron"))
stored$raudenbush
```

```{r compare-numbers, include = FALSE}
cmp <- as.data.frame(stored$raudenbush)
by_m <- function(table, m, column) range(abs(table[[column]][table$M == m]))
```

The rows are in the order of `elpd_loo`, the best first. `elpd_diff` is
the difference from the first row and `se_diff` is the standard error of
that difference. A difference of less than about two standard errors does
not separate two specifications. With as few as 19 sites the standard
error is itself uncertain, so the rule is a rough guide.

For the teacher expectancy studies nothing stands out. The number of grid
points hardly matters, because for a given `M` the three values of `L`
give nearly the same `elpd_loo`. The fits with eight spline functions come
first, and those with six follow at a distance of about
`r num(mean(by_m(cmp, 6, "elpd_diff")), 1)` with a standard error of about
`r num(mean(by_m(cmp, 6, "se_diff")), 1)`. Those with four are
`r num(by_m(cmp, 4, "elpd_diff")[1], 1)` to
`r num(by_m(cmp, 4, "elpd_diff")[2], 1)` behind, still less than two
standard errors. They are also the rows that the note under the table
names: their fits had divergent transitions, and a fit with such a problem
should not be chosen whatever its `elpd_loo`. For these data the
comparison gives no reason to depart from the default.

The choice matters more when the true effects form separate groups. The
second stored comparison is for simulated data of that kind, with
`r nrow(stored$two_groups_data)` sites, true effects in two groups and
unequal standard errors, following the design of the simulations in
@lee2025 [Section 5].

```{r two-groups-data, fig.height = 3.5, fig.alt = "Histograms of the simulated true effects and of their estimates."}
sim <- stored$two_groups_data
op <- par(mfrow = c(1, 2))
hist(sim$theta, breaks = 30, main = "True effects", xlab = "")
hist(sim$theta_hat, breaks = 30, main = "Estimates", xlab = "")
par(op)
```

The two groups are plain in the true effects and blurred in the estimates,
because many of the standard errors are large compared with the distance
between the groups.

```{r compare-two-groups}
stored$two_groups
```

```{r compare-numbers-2, include = FALSE}
cmp2 <- as.data.frame(stored$two_groups)
four <- cmp2[cmp2$M == 4, ]
# Spacing of the default grid with 51 points, in units of the smallest
# standard error.
spacing_51 <- 2 * diff(range(sim$theta_hat)) / 50 / min(sim$sigma)
# The paragraph below is about these fits; stop if it no longer holds.
stopifnot(
  all(four$max_rhat > 1.05),
  all(cmp2$max_rhat[cmp2$M != 4] < 1.01)
)
```

The fits with four spline functions are
`r num(min(abs(four$elpd_diff)), 1)` to `r num(max(abs(four$elpd_diff)), 1)`
behind the best one, which is `r num(min(abs(four$elpd_diff) / four$se_diff), 1)`
to `r num(max(abs(four$elpd_diff) / four$se_diff), 1)` standard errors, but
they did not converge (`vignette("diagnostics")` looks at them more
closely), and the scores of fits that have not converged cannot be relied
on. What the table shows is that four functions are not a specification to
use for these data, and not how much worse they predict. The fits with six
and eight functions converged and are close to each other, with
differences of about `r num(mean(by_m(cmp2, 6, "elpd_diff")), 1)` and
standard errors of about `r num(mean(by_m(cmp2, 6, "se_diff")), 1)`. The
number of grid points again makes no difference that can be seen, although
with 51 points the spacing of this grid is `r num(spacing_51, 1)` times the
smallest standard error.

Every row of such a table is a full fit, so a comparison takes as long as
that many fits, and the column `sampler_seconds` shows how the time grows
with `L`. Since the number of grid points made no difference in either
example or in the simulation of the paper, we suggest varying `M` first.

## What the choices change in the results

`elpd_loo` measures how well the fitted $g$ predicts the estimates
$\hat\theta_i$. That is not the same as how accurately each true effect is
estimated, and two specifications that predict equally well can give
different results for a site. The results themselves can be compared
directly. The stored file holds nine more fits of the teacher expectancy
studies, for three widths of the grid and six, eight and ten spline
functions.

```{r sensitivity}
sens <- stored$raudenbush_sensitivity
shown <- sens[, c("expansion", "M", "mean_g", "sd_g", "study18_mean",
                  "study4_mean", "study4_lower", "study4_upper",
                  "elpd_diff", "se_diff")]
names(shown)[5:8] <- c("study18", "study4", "lower", "upper")
round(shown, 2)
```

`study18` and `study4` are the posterior means of the effects of two
studies, and `lower` and `upper` are the limits of the 90% interval of
study 4.

```{r sensitivity-numbers, include = FALSE}
span <- function(x) paste(num(min(x)), "to", num(max(x)))
between <- function(x) paste(num(min(x)), "and", num(max(x)))
worst <- which.min(sens$elpd_diff)
# The sentence on study 4 below is about these fits; stop if it no longer
# holds: smaller with a wider grid for each M, larger with more spline
# functions for each width.
falls_with_width <- tapply(sens$study4_mean, sens$M, function(x) all(diff(x) < 0))
rises_with_m <- tapply(sens$study4_mean, sens$expansion, function(x) all(diff(x) > 0))
stopifnot(all(falls_with_width), all(rises_with_m))
```

The mean of $g$ lies between `r between(sens$mean_g)` in the nine fits and
its standard deviation between `r between(sens$sd_g)`. The posterior mean
of study 18, whose estimate is the most precise, lies between
`r between(sens$study18_mean)`. These results hardly depend on the
specification. The posterior mean of study 4 does. That study has the
largest estimate, `r num(raudenbush1985$yi[4])`, and a large standard
error, and its posterior mean ranges from `r span(sens$study4_mean)`; it is
smaller the wider the grid and the fewer the spline functions. The last
two columns do not say which of these fits to prefer: the largest
difference in `elpd_loo` from the best of the nine is
`r num(abs(sens$elpd_diff[worst]), 1)`, with a standard error of
`r num(sens$se_diff[worst], 1)`.

How far an extreme and imprecise estimate is drawn in depends on the tail
of $g$, and 19 studies leave the tail uncertain. The intervals for study 4
are wide in every fit and say as much. When a conclusion rests on such a
site, we suggest reporting its effect under two or three specifications
and not under one.

For the `r nrow(sim)` simulated sites the specification matters much less.
Three widths of the grid and six or eight spline functions give the same
results to within a few hundredths.

```{r sensitivity-two-groups}
sens2 <- stored$two_groups_sensitivity
round(sens2[, c("expansion", "M", "mean_g", "sd_g", "coverage", "rmse",
                "treedepth_hits")], 2)
```

`coverage` is the share of the true effects that lie inside the 90%
intervals and `rmse` is the root mean squared error of the posterior
means. The true effects have mean `r num(mean(sim$theta))` and standard
deviation `r num(sd(sim$theta))`. The last column shows a cost of a wide
grid: the wider the grid, the more of the 12,000 iterations reach the
maximum tree depth.

## References
