---
title: "Using metafor and data frames as input"
description: "Getting estimates and standard errors into the model from a data frame, a list or metafor, and labeling the results."
output:
  rmarkdown::html_vignette:
    fig_width: 7
    fig_height: 4.5
bibliography: references.bib
link-citations: true
vignette: >
  %\VignetteIndexEntry{Using metafor and data frames as input}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
library(bayesEfron)
options(bayesEfron.use_cli = FALSE, width = 80)
has_metafor <- requireNamespace("metafor", quietly = TRUE)
num <- function(x, digits = 2) formatC(x, format = "f", digits = digits)
```

The model needs two numbers for each site: an estimate of the effect and
the standard error of that estimate. Data seldom arrive in that form. A
table may hold variances where standard errors are wanted, or the estimates
may still have to be computed from the means, standard deviations and
sample sizes of two groups. The sites usually have names as well, which you
will want to see again next to the results. `as_bef_data()` deals with the
first two points and keeps the names.

## Estimates already in a data frame

The teacher expectancy studies of @raudenbush1985 come as a data frame
with one row for each study.

```{r data}
studies <- raudenbush1985[, c("author", "yi", "vi", "sei")]
studies$sei <- round(studies$sei, 4)
head(studies, 4)
```

`yi` is the estimate, a standardized mean difference, `vi` is its sampling
variance and `sei` its standard error. `as_bef_data()` finds the two
columns it needs by their names.

```{r by-name}
dat <- as_bef_data(raudenbush1985)
dat
```

For the estimates it takes the first of the names `theta_hat`, `yi`,
`estimate`, `effect` and `y` that the data frame has, and for the standard
errors the first of `sigma`, `se`, `std_error`, `stderr` and `sei`. Columns
with other names are given in the arguments `theta_hat` and `sigma`. If
your second column holds variances, as `vi` does, set `variance = TRUE` and
the function takes the square roots.

```{r variance}
dat_v <- as_bef_data(raudenbush1985, theta_hat = "yi", sigma = "vi",
                     variance = TRUE)
all.equal(dat_v$sigma, dat$sigma)
```

The function stops if a value is missing, if a standard error is not
positive or if there are fewer than five sites. It does not drop incomplete
rows, because the number of sites would then change without notice.

```{r missing}
incomplete <- raudenbush1985
incomplete$sei[3] <- NA
try(as_bef_data(incomplete))
```

## From group summaries to estimates

Often the data are one step further back, and each study reports the mean,
the standard deviation and the size of two groups. `metafor::escalc()`
computes an estimate and its sampling variance from such summaries
[@viechtbauer2010], and `as_bef_data()` accepts its result. The example is
the nine studies of the length of hospital stay of stroke patients under
specialized and under routine care that @normand1999 analyzed. The data
are in the metadat package, which metafor loads.

```{r escalc, eval = has_metafor, message = FALSE}
library(metafor)
stroke <- escalc(
  measure = "MD",
  m1i = m1i, sd1i = sd1i, n1i = n1i,
  m2i = m2i, sd2i = sd2i, n2i = n2i,
  data = dat.normand1999
)
stroke[, c("source", "m1i", "m2i", "yi", "vi")]
```

With `measure = "MD"` the estimate `yi` is the difference between the two
means, here in days, and `vi` is the sum of the two squared standard errors
of the means, $s_1^2 / n_1 + s_2^2 / n_2$. A difference in the original
units is the natural choice when every study measures the outcome in the
same way, as these do. When the studies use different instruments, as the
teacher expectancy studies did with their IQ tests, the standardized mean
difference (`measure = "SMD"`) puts the estimates on a common scale.

```{r escalc-convert, eval = has_metafor}
rownames(stroke) <- stroke$source
stroke_dat <- as_bef_data(stroke)
stroke_dat
```

For an `escalc` object the estimates are the column `yi` and the standard
errors are the square roots of `vi`. The site labels are taken from the row
names, which is why the row names were set first; labels given to
`escalc()` in its argument `slab` are not used.

Nine studies are enough for the function to run, but they can say little
about the shape of the distribution of effects, and with so few the results
depend more on the choices that `vignette("choosing-a-grid")` describes.
The check of the intervals that `vignette("implementation")` reports starts
at 50 sites; we have no evidence on how they perform with fewer.

## A list

When the estimates and standard errors are two vectors, they can be passed
to `bayes_efron_fit()` directly. `as_bef_data()` accepts them as a list as
well, which is a way to have them checked beforehand.

```{r list}
as_bef_data(list(
  theta_hat = c(0.03, 0.12, -0.14, 1.18, 0.26),
  sigma = c(0.125, 0.147, 0.167, 0.373, 0.369),
  names = c("Rosenthal", "Conn", "Jose", "Pellegrini 1", "Pellegrini 2")
))
```

## Fitting, and putting the names back

The elements `theta_hat` and `sigma` of the result are the two arguments of
`bayes_efron_fit()`.

```{r fit-call, eval = FALSE}
dat <- as_bef_data(raudenbush1985, site = "author")
fit <- bayes_efron_fit(dat$theta_hat, dat$sigma, seed = 1985)
```

The fitted object numbers the sites in the order of the data and does not
keep their labels, so the labels have to be put back by hand. The object
`raudenbush_fit` in the package holds the result of this fit and keeps one
posterior draw in fifteen.

```{r labels}
dat <- as_bef_data(raudenbush1985, site = "author")
fit <- raudenbush_fit
results <- cbind(study = dat$names, as.data.frame(fit))
results$mean <- round(results$mean, 2)
head(results[, c("study", "mean", "hpdi_lower", "hpdi_upper")])
```

`mean` is the posterior mean of the effect. `hpdi_lower` and `hpdi_upper`
are the 5% and 95% posterior quantiles, the limits of a 90% interval; in
spite of their names the interval is equal-tailed. `confint(fit,
level = 0.95)` gives 95% intervals, the level that metafor uses by default
for its confidence intervals, and `?bef_fit` describes all the columns.

## What differs from a random-effects model

metafor fits the random-effects model to the same two columns. The
difference between the two analyses lies in what is assumed about the
distribution of true effects. The random-effects model takes it to be
normal, with a mean and a standard deviation that are estimated. bayesEfron
estimates the distribution itself, $g$, and reports its mean and standard
deviation as summaries.

```{r normal, eval = has_metafor}
normal_fit <- rma(yi, vi, data = raudenbush1985)
c(mean = unname(coef(normal_fit)), sd = sqrt(normal_fit$tau2))
confint(fit, type = "g")[c(1, 3), ]
```

```{r outside, include = FALSE}
grid <- fit$metadata$data_list$grid
g <- colMeans(exp(posterior::as_draws_matrix(fit)[, sprintf("log_g[%d]", seq_along(grid))]))
outside <- grid < min(raudenbush1985$yi) | grid > max(raudenbush1985$yi)
mean_g <- sum(g * grid)
share <- sum(g[outside] * (grid[outside] - mean_g)^2) / sum(g * (grid - mean_g)^2)
# The next paragraph says "about twice"; stop if that no longer holds.
if (has_metafor) {
  g_summary <- confint(fit, type = "g")
  ratio <- g_summary$point[g_summary$site == "sd_g"] / sqrt(normal_fit$tau2)
  stopifnot(length(ratio) == 1, abs(ratio - 2) < 0.4)
}
```

The two means are close, whereas the standard deviation of $g$ is about
twice that of the normal model. Both numbers describe how much the true
effects vary between studies, and they differ because the two models allow
different shapes for that variation. Part of the difference is the long
right tail of the estimated $g$ (it is shown in `vignette("bayesEfron")`),
which a normal distribution cannot have. The other part is that $g$ keeps a
little probability beyond the smallest and the largest estimate, where the
data hardly bear on it: `r num(sum(g[outside]))` in all, which accounts for
`r round(100 * share)`% of the variance of the estimated distribution, the
posterior mean of $g$. A standard deviation is sensitive to such
probability, so that of $g$ says as much about the tails that the model
allows as about the bulk of the effects.

The site effects differ too. `blup()` gives the prediction of the normal
model for each study, which is drawn toward the overall mean by a factor
that depends only on the standard error. With bayesEfron, how far an
estimate is drawn in depends on the shape of $g$ as well.

```{r blup, eval = has_metafor}
compare <- data.frame(
  study = raudenbush1985$author,
  estimate = raudenbush1985$yi,
  normal = round(blup(normal_fit)$pred, 2),
  bayesEfron = round(coef(fit), 2)
)
compare[order(-compare$estimate)[1:6], ]
```

For most of the studies the two columns are close, but they part for the
largest estimates. Under the normal model the largest estimate, which has a
large standard error, is taken to be mostly noise and is drawn most of the
way to the mean. Under the estimated $g$, with its long right tail, a true
effect of that size is less surprising, and the estimate is drawn in less
far. Which of the two is nearer the truth cannot be read from 19 studies.
The comparison shows that the answer for the extreme studies depends on
what is assumed about the distribution.

## References
