---
title: "Checking convergence and plotting"
description: "Reading the convergence diagnostics of a fit, acting on the warnings, and the four plots."
output:
  rmarkdown::html_vignette:
    fig_width: 7
    fig_height: 4.5
bibliography: references.bib
link-citations: true
vignette: >
  %\VignetteIndexEntry{Checking convergence and plotting}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
library(bayesEfron)
options(bayesEfron.use_cli = FALSE, width = 80)
has_ggplot2 <- requireNamespace("ggplot2", quietly = TRUE)
has_loo <- requireNamespace("loo", quietly = TRUE)
num <- function(x, digits = 2) formatC(x, format = "f", digits = digits)
count <- function(n) if (n == 0) "none" else format(n, big.mark = ",")
```

Every number that a fit reports is computed from the draws of a Markov
chain, and it describes the posterior distribution only if the chains have
reached that distribution and have moved through it often enough. Whether
they have cannot be seen from the results themselves: a fit that has not
converged prints a table of estimates that looks like any other. The
diagnostics are the evidence on that point, and they should be read before
the estimates are.

## What `diagnose()` returns

The fit examined here is `raudenbush_fit`, the model fitted to the teacher
expectancy studies with the default settings.

```{r diagnose}
fit <- raudenbush_fit
diag <- diagnose(fit)
diag
```

`diagnose()` collects what `bayes_efron_fit()` computed when the model was
fitted, and the printed lines give the worst value of each diagnostic.
`raudenbush_fit` stores one posterior draw in fifteen, but its diagnostics
are those of the full run of four chains with 3,000 sampling iterations
each. `?diagnose` lists the elements of the object. Two of them name the
diagnostics that call for attention: `sampler_diagnostics_failed` names
those that fail a check, and `sampler_diagnostics_warned` those that only
deserve a look.

```{r flags}
diag$sampler_diagnostics_failed
diag$sampler_diagnostics_warned
```

## Reading the diagnostics

The sampler moves through the posterior distribution in a sequence of
small steps, whose size it sets during warmup, and each iteration is a
path of many steps [@hoffman2014]. R-hat and the effective sample sizes
describe the draws that result. The other three diagnostics describe the
paths.

R-hat compares the chains with one another. If they have all settled on
the same distribution, the variation between chains is no larger than the
variation within them, and R-hat is close to 1. The effective sample size
says how many independent draws the correlated draws of the chains are
worth. The bulk effective sample size concerns the center of the
distribution, and with it the posterior means, whereas the tail effective
sample size concerns the 5% and 95% quantiles, and with them the
intervals. @vehtari2021 recommend that the draws be used only if, for
every quantity, R-hat is below 1.01 and both effective sample sizes are at
least 400 when four chains are run. `diagnose()` lists a warning when a
value falls short of these recommendations. It counts an R-hat above 1.05
as a failed check, and `bayes_efron_fit()` then gives a warning as soon as
the fit ends.

A divergent transition means that the sampler met a region of the
posterior distribution that it could not follow with the step size it had
chosen. The draws may then miss part of the distribution, so `diagnose()`
counts even one divergent transition as a failed check. An iteration
reaches the maximum tree depth when its path is cut short at the greatest
length allowed. That makes the sampler less efficient, which shows in the
effective sample sizes, but it does not bias the draws, and it is listed as
a warning and not as a failure. The energy Bayesian fraction of missing
information (E-BFMI) is computed for each chain and is low when the chain
moves slowly between the high and the low levels of the log density.
`diagnose()` lists a warning below 0.2, whereas cmdstanr prints a message
of its own below 0.3, so a message can appear for a fit that `diagnose()`
does not flag.

When a fit has fewer than 400 draws in all, R-hat and the effective sample
sizes are too unreliable to act on. Their values are still returned, but
they are not compared with the thresholds, and the element
`diagnostic_skipped` says so. Divergent transitions, the maximum tree depth
and E-BFMI are checked however short the run.

```{r ess-reported}
reported <- c("mean_g", "sd_g", sprintf("theta_mean[%d]", c(4, 10, 18)))
data.frame(
  rhat = round(diag$rhat[reported], 3),
  ess_bulk = round(diag$ess_bulk[reported]),
  ess_tail = round(diag$ess_tail[reported])
)
```

For the teacher expectancy studies the largest R-hat is
`r num(max(diag$rhat, na.rm = TRUE), 3)`. The smallest bulk effective
sample size, `r round(min(diag$ess_bulk, na.rm = TRUE))`, belongs to
`r names(which.min(diag$ess_bulk))`, the log density, whereas the
quantities that are reported, such as those in the table, have larger
ones. There is no divergent transition. The maximum tree depth was reached
in `r sum(diag$max_treedepth)` of the 12,000 iterations, and since the
effective sample sizes are adequate nothing needs to be done about it. The
smallest E-BFMI is
`r num(min(diag$ebfmi))`, which lies between the two thresholds: cmdstanr
printed its message when the model was fitted, and `diagnose()` lists no
warning.

## What to do about a warning

The remedies are arguments of `bayes_efron_fit()`.

```{r remedies, eval = FALSE}
theta_hat <- raudenbush1985$yi
sigma <- raudenbush1985$sei

# Divergent transitions: smaller steps.
bayes_efron_fit(theta_hat, sigma, adapt_delta = 0.99)

# Iterations that reach the maximum tree depth: allow longer paths.
bayes_efron_fit(theta_hat, sigma, max_treedepth = 12)

# Small effective sample sizes, or R-hat between 1.01 and 1.05: run longer.
bayes_efron_fit(theta_hat, sigma, iter_warmup = 2000, iter_sampling = 6000)
```

These remedies address the sampler. The grid can be a cause as well. When
a large share of the iterations reaches the maximum tree depth, the grid
may be wider than it needs to be. For the simulated data of
`vignette("choosing-a-grid")` a narrower grid (a smaller `expansion`)
removed those iterations without changing the results, and for the fits
with 1,500 sites in `vignette("implementation")` it brought R-hat below
1.05.

When R-hat is well above 1.05, the specification of the model deserves a
look. The simulated data with two separate groups of effects give an
example. The table shows three fits of those data: with four spline
functions and the default length of run, with four spline functions and a
run four times as long, and with six spline functions.

```{r not-converged}
stored <- readRDS(system.file("examples", "grid_comparison_example.rds",
                              package = "bayesEfron"))
runs <- stored$two_groups_runs
shown <- runs[, c("M", "iter_sampling", "draws", "max_rhat", "min_ess_bulk",
                  "treedepth_hits", "coverage")]
shown$max_rhat <- round(shown$max_rhat, 2)
shown$min_ess_bulk <- round(shown$min_ess_bulk)
shown
```

```{r not-converged-numbers, include = FALSE}
short <- runs[1, ]
long <- runs[2, ]
six <- runs[3, ]
# The sentences below are about these fits; stop if they no longer hold.
stopifnot(
  short$M == 4, long$M == 4, six$M == 6, long$draws == 4 * short$draws,
  short$max_rhat > 1.05, six$max_rhat < 1.01,
  diff(range(runs$coverage)) < 0.02
)
```

With four spline functions the default run does not converge: R-hat is
`r num(short$max_rhat)` and the smallest effective sample size is
`r round(short$min_ess_bulk)` out of
`r format(short$draws, big.mark = ",")` draws. The run four times as long
brings R-hat down to `r num(long$max_rhat)`, but its smallest effective
sample size is still only `r round(long$min_ess_bulk)` out of
`r format(long$draws, big.mark = ",")` draws, and
`r format(long$treedepth_hits, big.mark = ",")` of its iterations reached
the maximum tree depth. The chains move very slowly, and more iterations
buy little. With six spline functions the default run converges. Why four
functions make the sampler struggle with these data is not established
here; the practical point is that the remedy was a different specification
and not a longer run.

The last column is the share of the true effects that lie inside the 90%
intervals, which can be computed here because the data are simulated. It
is nearly the same in the three fits. The diagnostics, then, do not measure
how wrong a fit is. They say whether there is evidence that the chains
have sampled the posterior distribution, and for the first two fits the
evidence is that they have not.

```{r benchmark, include = FALSE}
benchmark <- utils::read.csv(
  system.file("verification", "benchmark-coverage.csv", package = "bayesEfron"),
  check.names = FALSE
)
large <- benchmark[benchmark$K == max(benchmark$K), ]
smaller <- benchmark[benchmark$K < max(benchmark$K), ]
```

Convergence also becomes harder as the number of sites grows. In the
benchmark that `vignette("implementation")` describes, which uses the
default sampler settings, `r count(sum(smaller[["n_rhat_above_1.05"]]))` of
the `r sum(smaller$n_replications)` fits with up to
`r format(max(smaller$K), big.mark = ",")` sites had an R-hat above 1.05,
whereas `r count(large[["n_rhat_above_1.05"]])` of the
`r large$n_replications` fits with `r format(large$K, big.mark = ",")`
sites did.

## Plots of a fit

`plot()` draws four plots, chosen with `type`. With ggplot2 installed it
returns a `ggplot` object, which can be changed like any other;
`backend = "base"` uses base graphics.

The first two show what the fit estimates. `"caterpillar"`, the default,
shows the posterior mean of each site effect with an interval of
probability `level` and the central 50% interval. The sites are in the
order of their posterior means unless `sort_by` says otherwise: `"sigma"`
orders them by standard error, which shows how the intervals widen as the
estimates become less precise, and `"none"` keeps the order of the data.

```{r caterpillar, fig.height = 5, fig.alt = "Site effects ordered by standard error, with 95 percent intervals."}
plot(fit, type = "caterpillar", sort_by = "sigma", level = 0.95)
```

`"g"` shows the estimated distribution of effects, that is, the posterior
mean of the probability of each grid point, with a pointwise band.

```{r g-plot, eval = has_ggplot2, fig.alt = "The estimated distribution of effects with a relabeled axis."}
plot(fit, type = "g") +
  ggplot2::labs(title = NULL, x = "Standardized mean difference",
                y = "Probability of the grid point")
```

The other two show diagnostics. `"diagnostic"` draws the largest R-hat,
the smallest bulk and tail effective sample sizes and the numbers of
divergent transitions and of iterations at the maximum tree depth, the
values that `diagnose()` prints. E-BFMI is not drawn.

```{r diagnostic-plot, fig.height = 3, fig.alt = "The worst value of each convergence diagnostic."}
plot(fit, type = "diagnostic")
```

`"loo"` shows the Pareto k value of each site from leave-one-out
cross-validation, with reference lines at 0.5 and 0.7. A large value means
that leaving the site out would change the posterior distribution a good
deal, so that the approximation is poor for that site. Here the values are
all small. `vignette("choosing-a-grid")` says more about this criterion.

```{r loo-plot, eval = has_loo, fig.height = 3.5, fig.alt = "Pareto k values by site."}
plot(fit, type = "loo", backend = "base")
```

## References
