---
title: "Implementation and checks"
description: "Reproducing a fit, the Stan model, the checks the package makes, coverage on a benchmark with known effects, and the notation of the paper."
output:
  rmarkdown::html_vignette:
    fig_width: 7
    fig_height: 4.5
bibliography: references.bib
link-citations: true
vignette: >
  %\VignetteIndexEntry{Implementation and checks}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
library(bayesEfron)
options(bayesEfron.use_cli = FALSE, width = 80)
num <- function(x, digits = 2) formatC(x, format = "f", digits = digits)
count <- function(n) if (n == 0) "none" else format(n, big.mark = ",")
listing <- function(x) {
  if (length(x) < 2) return(as.character(x))
  paste(paste(x[-length(x)], collapse = ", "), "and", x[length(x)])
}
fit <- raudenbush_fit
```

Someone who reports results from this package may be asked whether the fit
can be reproduced and what evidence there is that the intervals mean what
they say.

## Reproducing a fit

A fitted object records what produced its draws.

```{r reproduce}
fit$metadata$seed
fit$metadata$cmdstan_version
substr(fit$metadata$stan_file_sha256, 1, 12)
```

`seed` is the seed of the sampler. If none is given to `bayes_efron_fit()`
one is taken from the clock, and it is stored either way. The last value is
the beginning of a hash of the Stan file. The settings of the sampler and
the versions of bayesEfron and cmdstanr are attributes of `metadata`.

```{r sampler-settings}
str(attr(fit$metadata, "sampler_settings"))
attr(fit$metadata, "package_version")
attr(fit$metadata, "cmdstanr_version")
```

The data, the grid and the spline basis are in `fit$metadata$data_list`.
With the same data, settings, seed and versions on the same machine, the
draws are the same. Between machines or versions of CmdStan they can
differ, and the results then agree to within Monte Carlo error only.

## The Stan model

The model is one Stan file, which can be read at

```{r stan-file, eval = FALSE}
system.file("stan", "efron_re.stan", package = "bayesEfron")
```

Its comments state the model, which `vignette("model")` describes. The
file is compiled the first time a model is fitted. The compiled model is
kept in the user's cache directory and used again in later sessions, and it
is recompiled when the Stan file or the version of CmdStan changes.
`?bayes_efron_compile` says where the directory is and how to clear it.

## Checks against known answers

The tests of the package check its computations in two steps. First the
formulas of the model are written out in plain R and compared with results
obtained in another way: the posterior distribution of a site effect under
a normal $g$ with the conjugate normal result, and the mean and variance of
$g$ with Gauss-Hermite quadrature. Then the quantities that the Stan model
wrote for the stored fit `raudenbush_fit` are recomputed with those R
formulas from the draws of the spline coefficients and compared: the
probabilities of $g$, its mean and variance, the posterior mean, standard
deviation and mode of each site, the log likelihood and the effective
number of parameters.

Other tests compare fits with deconvolveR, the implementation by
@narasimhan2020 of the empirical Bayes estimator of @efron2016, on three
simulated data sets with equal standard errors and 20, 50 and 100 sites.
The two methods are not expected to coincide. The tests require that the
posterior means of the site effects lie within one standard error of the
estimates of deconvolveR, with a correlation of at least 0.99, and that the
two estimates of $g$ for the largest data set differ by at most 0.25 in
total variation distance.

These tests show that the pieces are computed correctly. They do not show
that the intervals have the right coverage.

## A benchmark with known effects

The benchmark repeats, with the package and with Markov chain Monte Carlo,
a check that @lee2025 [Appendix E] made with a variational approximation.
The true effects are the 1,500 values of their simulation study, in two
separate groups: 500 between $-1.7$ and $-0.7$ and 1,000 between $0.7$ and
$2.7$. The standard errors are unequal, the largest being 9 times the
smallest. Data sets of 50, 100, 200, 500 and 1,500 sites are fitted, 20 of
each size, with the default sampler settings, the default rule for the
ends of the grid and the numbers of grid points used in the paper (51, 71,
81, 101 and 101).

The 20 data sets of 50 to 500 sites are stratified random samples of the
sites of one simulated panel of 1,500. They differ in the sites that they
hold, and they overlap, more so the larger they are. The 20 data sets of
1,500 sites hold the same true effects, each with its own assignment of
the standard errors to the sites and its own errors. The coverage of a fit
is the share of sites whose true effect lies inside the 90% interval that
`confint()` returns.

```{r benchmark}
benchmark <- utils::read.csv(
  system.file("verification", "benchmark-coverage.csv", package = "bayesEfron"),
  check.names = FALSE
)
shown <- benchmark[, c("K", "n_replications", "coverage", "coverage_se",
                       "max_rhat", "n_rhat_above_1.05", "min_ess_bulk",
                       "total_divergences", "total_treedepth_hits")]
shown[c("coverage", "coverage_se", "max_rhat")] <-
  round(shown[c("coverage", "coverage_se", "max_rhat")], 3)
shown$min_ess_bulk <- round(shown$min_ess_bulk)
shown
```

```{r benchmark-numbers, include = FALSE}
large <- benchmark[benchmark$K == max(benchmark$K), ]
small <- benchmark[benchmark$K < max(benchmark$K), ]
warned <- small$K[small$max_rhat > 1.01 | small$min_ess_bulk < 400]
failed_small <- sum(small[["n_rhat_above_1.05"]])
iterations <- 12000 * large$n_replications
```

The table was computed on `r benchmark$generated_on[1]` with CmdStan
`r benchmark$cmdstan_version[1]`. `coverage_se` is the standard deviation of
the coverage between data sets divided by the square root of their number.
It measures the variation between the 20 data sets of one size and nothing
more: the data sets of up to `r format(max(small$K), big.mark = ",")` sites
are samples of the sites of one simulated panel, and those of
`r format(large$K, big.mark = ",")` sites share one set of true effects, so
the variation that another panel or another set of true effects would
bring is not in it. The last five columns describe convergence: the
largest R-hat and the smallest bulk effective sample size over the fits,
the number of fits with an R-hat above 1.05, and the total numbers of
divergent transitions and of iterations that reached the maximum tree
depth.

With up to `r format(max(small$K), big.mark = ",")` sites the coverage is
about `r num(mean(small$coverage), 2)` at every size, a little below the
nominal 0.90. For the same sizes @lee2025 report 88.7% to 89.6% with the
variational approximation.
`r if (failed_small == 0) "None" else failed_small` of the
`r sum(small$n_replications)` fits `r if (failed_small <= 1) "has" else "have"`
an R-hat above 1.05, although at `r listing(warned)` sites the largest
R-hat exceeds 1.01 or the smallest bulk effective sample size is below 400,
so some of those fits would have drawn a warning from `diagnose()`.

For `r format(large$K, big.mark = ",")` sites the coverage is
`r num(large$coverage, 3)`, but `r count(large[["n_rhat_above_1.05"]])` of
the `r large$n_replications` fits did not converge with the default
settings: the largest R-hat is
`r num(large$max_rhat)`, the smallest bulk effective sample size is
`r round(large$min_ess_bulk)`, and
`r format(large$total_treedepth_hits, big.mark = ",")` of the
`r format(iterations, big.mark = ",")` iterations reached the maximum tree
depth. The row therefore describes what the default settings give at that
size and is not evidence about the coverage of a converged fit.

We refitted the data sets of the fits that did not converge in two ways,
with a narrower grid and with a larger maximum tree depth.

```{r remedies}
remedies <- utils::read.csv(
  system.file("verification", "benchmark-remedies.csv", package = "bayesEfron")
)
shown <- remedies[, c("replication", "setting", "max_rhat", "min_ess_bulk",
                      "treedepth_hits", "coverage")]
shown$max_rhat <- round(shown$max_rhat, 2)
shown$min_ess_bulk <- round(shown$min_ess_bulk)
shown$coverage <- round(shown$coverage, 3)
shown
```

```{r remedies-numbers, include = FALSE}
narrow <- remedies[remedies$setting == "expansion = 0.05", ]
deeper <- remedies[remedies$setting == "max_treedepth = 12", ]
# The sentence on the narrower grid is about these fits; stop if it no
# longer holds.
stopifnot(nrow(narrow) == large[["n_rhat_above_1.05"]], all(narrow$max_rhat < 1.05))
```

With a grid that ends just beyond the estimates (`expansion = 0.05` in
place of 0.5), R-hat was at most `r num(max(narrow$max_rhat))` and no fit
failed the check, although `r sum(narrow$min_ess_bulk < 400)` of the
`r nrow(narrow)` still had an effective sample size below 400. With the
default grid and a maximum tree depth of 12,
`r count(sum(deeper$max_rhat > 1.05))` of the `r nrow(deeper)` fits still
had an R-hat above 1.05, and each took several times as long.

In these data many of the standard errors are large, so the estimates
spread over a much wider range than the true effects, and the default
grid, which extends beyond the estimates by half their range on each side,
is wider than it needs to be. This cannot be made a general rule, however.
When the standard errors are small, the estimates spread little more than
the true effects, and room beyond them is needed
(`vignette("choosing-a-grid")`).

The benchmark covers one shape of $g$ and one pattern of standard errors.
It says nothing about data whose standard errors are much larger relative
to the spread of the effects, or about fewer than 50 sites. The scripts and
the data are in `data-raw/benchmark/` in the source repository.

## Relation to the paper

The package implements the fully Bayesian model of @lee2025 [Section 4].
It writes the quantities of the paper as follows.

| Paper | Package | Meaning |
|:---|:---|:---|
| $\Theta_i$ | $\theta_i$ in the documentation | True effect of site $i$ |
| $\hat\theta_i$, $\sigma_i$ | `theta_hat`, `sigma` | Estimate and its standard error |
| $\mathcal{T} = \{\theta_1, \ldots, \theta_m\}$ | `grid`, with `L` points $\tau_j$ | Grid on which $g$ is defined |
| $Q$, an $m \times p$ matrix | `B`, an `L` by `M` matrix | Spline basis on the grid |
| $\alpha$, $\lambda$ | `alpha`, `lambda` | Spline coefficients and their prior precision |
| $g_j$ | `exp(log_g)` | Probability of grid point $j$ |
| $\pi_{ij}$ | $w_{ij}$ in `vignette("model")`; not stored | Posterior probability of grid point $j$ for site $i$, given $\alpha$ |

The grid methods `"paper_realdata"`, `"paper_simulation"` and
`"paper_sensitivity"` are the rules of the code that accompanies the
paper, <https://github.com/joonho112/fully-bayes-efron-prior>, for the
application, the simulation study and the sensitivity analysis. Where the
text of the paper and its code differ in a detail, the package follows the
code. The Stan appendix of the paper, for example, shows a grid that
extends 0.5 beyond the estimates, whereas the simulation code builds it
from the true effects.

Not in the paper are the effective number of parameters, the log
likelihood of each site with the cross-validation built on it, `predict()`
and the experimental grid method. The table of site effects follows the
definitions of the paper [@lee2025, Section 4.3 and Appendix A]: the
posterior mean, the posterior standard deviation with both parts of the
variance, the posterior mode on the grid and an equal-tailed interval.

To cite the package, cite the paper; `citation("bayesEfron")` gives the
reference.

## References
