---
title: "M1 · The empirical-Bayes deconvolution problem"
description: "The Robbins-Efron formulation, prior-art landscape, and motivation for the fully Bayesian log-spline approach."
output:
  rmarkdown::html_vignette:
    fig_width: 7
    fig_height: 4.5
vignette: >
  %\VignetteIndexEntry{M1 · The empirical-Bayes deconvolution problem}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  echo     = TRUE,
  collapse = TRUE,
  comment  = "#>",
  message  = FALSE,
  warning  = FALSE,
  fig.width  = 7,
  fig.height = 4.5
)
library(bayesEfron)
has_ggplot2 <- requireNamespace("ggplot2", quietly = TRUE)
```

## Scope of this vignette

This is the first vignette in the methodological track. Its
purpose is to motivate the empirical-Bayes deconvolution problem
that `bayesEfron` solves and to justify the particular modelling
choices that distinguish this package from its closest relatives.
Subsequent vignettes in the methodological track (M2–M6) descend
through the parameterisation, the random-effects hierarchy, grid
construction, the Stan implementation, and the verification
programme. A short reading map at the end indicates where each
load-bearing claim is either proved or computationally checked.

## Notation registry

| Symbol | Meaning |
|:-------|:--------|
| $K$ | Number of sites |
| $\theta_i$ | Latent site effect |
| $\hat\theta_i$ | Observed effect estimate |
| $\sigma_i$ | Within-study standard error |
| $g(\theta)$ | Mixing distribution |
| $L$ | Grid length |
| $M$ | Spline degrees of freedom |
| $\alpha$ | Spline coefficient vector |
| $\lambda$ | Smoothness precision (half-Cauchy prior) |

## The setup

Index $K$ independent sites by $i = 1, \ldots, K$. From the design at
site $i$ the analyst extracts a real-valued effect estimate
$\hat\theta_i$ together with an associated within-study standard
error $\sigma_i$. The standard errors are treated as known inputs:
they are computed from the site-level design (typically by
`metafor::escalc()` or its analogue at the user's site) and are not
themselves parameters of the model. They are also generally
heteroscedastic — different sites have different sample sizes,
designs, and within-site variances, so the $\sigma_i$ vary
substantially across $i$ in practice.

The generative assumption is two-stage. First the latent site effects
are drawn independently from a common mixing distribution $g$:
$$
\theta_i \mid g \;\stackrel{\mathrm{iid}}{\sim}\; g,
\qquad i = 1, \ldots, K.
$$
Conditional on the latent effect, the observed estimate is normally
distributed around $\theta_i$ with the known within-study standard
error:
$$
\hat\theta_i \mid \theta_i \;\sim\; \mathcal{N}(\theta_i, \sigma_i^2),
\qquad i = 1, \ldots, K.
$$
The two stages together imply that the marginal density of
$\hat\theta_i$ is the convolution of $g$ with a normal kernel:
$$
f_i(\hat\theta) \;=\; \int \varphi(\hat\theta - \theta;\; \sigma_i)\;
g(\theta)\,\mathrm{d}\theta,
$$
where $\varphi(\,\cdot\,; \sigma)$ is the centred normal density with
standard deviation $\sigma$. The inference problem is therefore a
**deconvolution problem** in the classical sense: recover the mixing
distribution $g$ from observations of a convolved marginal $f_i$,
under heteroscedastic Gaussian noise. The heteroscedasticity matters:
different sites contribute marginal densities with different
smoothing kernels, so the convolved marginals do not admit a single
common decomposition.

## The Robbins-Efron formulation

The empirical-Bayes literature begins with Robbins's (1956)
compound-decision insight: when the same decision rule is applied
in parallel to $K$ related sub-problems, the optimal rule depends on
the marginal distribution of the parameters $\theta_i$ across the
$K$ sites, not on any individual $\theta_i$ in isolation. In the
deconvolution setting this insight implies that good per-site
predictions, shrinkage estimators, and tail-probability statements
all factor through the unknown mixing distribution $g$. Knowing $g$
is therefore not a secondary curiosity; it is a prerequisite for any
per-site Bayes rule.

Efron's contribution in a sequence of papers culminating in his 2016
*Biometrika* article ("Empirical Bayes deconvolution estimates")
recast the Robbins programme in explicitly density-modelling terms.
The framing is the marginal-density identity
$$
f(\hat\theta) \;=\; \int \varphi(\hat\theta - \theta;\; \sigma)\;
g(\theta)\,\mathrm{d}\theta,
$$
together with the explicit acknowledgement that the inferential
target — the latent prior $g$ — appears only inside an integral
convolved with the Gaussian sampling kernel. Three operational
consequences follow.

First, $g$ is identified up to the regularity of the convolution: the
Gaussian kernel is strictly positive and analytic, so distinct $g$
that generate distinct marginal densities are recoverable in
principle, but the recovery is ill-conditioned at fine scales. The
practical implication is that some form of regularisation — either
parametric form, a smoothness penalty, or a prior on the parameters
of $g$ — is required to obtain a stable estimate of $g$ at finite
$K$.

Second, the per-site posterior at any candidate $g$ is given by
Bayes's rule applied to the convolution: $p(\theta_i \mid
\hat\theta_i, g) \propto g(\theta_i)\,\varphi(\hat\theta_i -
\theta_i; \sigma_i)$. The shape of $g$ enters the posterior linearly
in $\theta$-space, so the per-site shrinkage and the location of
posterior mass depend tightly on how the estimator of $g$ has been
regularised.

Third, the empirical-Bayes step that estimates $g$ from the
$\{\hat\theta_i\}$ and plugs $\hat g$ into the per-site Bayes rule
treats $\hat g$ as if it were known. At large $K$ the sampling
uncertainty in $\hat g$ vanishes and the plug-in is asymptotically
honest; at the moderate $K$ that dominate applied work, the plug-in
omits a genuine source of posterior uncertainty. This finite-sample
gap is the proximate motivation for the fully Bayesian treatment
discussed in §"Fully Bayesian versus empirical-Bayes" below.

## The prior-art landscape

Several families of estimator for $g$ have been studied in the
deconvolution and empirical-Bayes literature. Four are directly
relevant to the design choices in `bayesEfron`.

**NPMLE.** The nonparametric maximum-likelihood estimator places no
parametric form on $g$ and maximises the marginal likelihood over
the space of all probability distributions. Lindsay's (1995) result
establishes that the solution is a discrete distribution with at most
$K$ atoms, located at data-determined support points. The NPMLE has
attractive properties — consistency, no functional-form bias, and a
convex optimisation that scales reasonably with $K$ — and is
implemented in `REBayes`, `ashr`, and `mixSQP`. Its disadvantage in
the present setting is precisely the discreteness of the solution.
When the true $g$ is a smooth density (as it is in essentially every
applied meta-analytic context where the underlying causal mechanism
varies continuously across sites), the NPMLE places mass on a finite
collection of points and is structurally unable to represent the
true density's continuous support. The discrete approximation can be
adequate for ranking and selection problems, but it is awkward for
inference on functionals of $g$ — such as the variance, the
quartiles, or the tail probabilities — that the meta-analyst
typically wants to report.

**Laplace prior.** Closed-form Laplace shrinkage assumes $g$ is a
Laplace density and exploits the resulting analytical posterior. The
estimator is fast, transparent, and admits closed-form credible
intervals. It commits, however, to a specific tail behaviour
(double-exponential decay) and to a unimodal symmetric shape, neither
of which is plausible across the heterogeneity patterns observed in
education and policy meta-analyses. The Laplace family also lacks a
natural mechanism for representing the asymmetry, skew, or multiple
modes that are common in real cross-site distributions of treatment
effects.

**Point-normal prior.** The point-normal (spike-and-slab) prior
models $g$ as a mixture of a point mass at zero (or another anchor)
and a Gaussian slab over the rest of $\mathbb{R}$. The
specification is parsimonious and useful when the true effects are
genuinely sparse — for example, in screening problems where most
effects are exactly null. As a default for meta-analytic
deconvolution it sits uneasily, because it mixes a parametric
sparsity assumption with the nonparametric flexibility the
deconvolution problem was meant to provide. When the true $g$ has no
mass at zero, the spike component forces the model to absorb dense
near-zero structure into a degenerate atom, distorting both the prior
estimate and the per-site posteriors.

**Log-spline prior.** Efron's chosen prior, and the one this package
implements, represents $g$ as a discretised exponential family on a
fixed support grid $\{\theta_1, \ldots, \theta_L\}$. The
unnormalised log-density at the grid points is a linear combination
of natural-cubic-spline basis functions: $\log\tilde g_j =
B_j^\top \alpha$, with $B \in \mathbb{R}^{L \times M}$ a fixed basis
matrix and $\alpha \in \mathbb{R}^M$ the parameter. The normalised
density is $g = \mathrm{softmax}(B\alpha)$. The log-spline prior is
parametric in the spline coefficient vector $\alpha$ but
nonparametric in the resulting family of densities: with even a
modest spline dimension $M$ the basis can represent skewed,
asymmetric, and modestly multimodal densities, and the implicit
smoothness of the cubic-spline basis prevents the pathological
peakiness that an unregularised NPMLE can produce. The fixed-grid
construction also makes the model amenable to off-the-shelf Bayesian
samplers, which the NPMLE — with its data-determined atoms — is not.

## Why log-spline

The log-spline prior is not the only smooth nonparametric density
family one might use for $g$, but it has three properties that
together account for its prominence in the deconvolution literature.

First, **flexibility**. The natural-cubic-spline basis on a fixed
grid can approximate a wide range of unimodal and modestly
multimodal smooth densities to arbitrary accuracy as the spline
dimension $M$ grows, while remaining tractable at small $M$. For the
applied meta-analytic distributions that motivate this package — a
few modes, possibly skewed, possibly heavy-tailed but not extremely
so — values of $M$ between 5 and 8 typically suffice. The package's
default $M = 6$ is calibrated to the benchmark distributions used in
the paper that motivates the package.

Second, **identifiability**. The softmax map is invariant to adding
a constant to every component of $B\alpha$, so an intercept column
in the basis would be unidentifiable. The standard remedy — used
here — is to construct the basis with `splines::ns(grid, df = M,
intercept = FALSE)`, which removes the redundant constant direction.
The resulting parameterisation is identifiable on the spline
coefficients, and the proper hyperprior on the smoothness scale
(see below) ensures a proper joint posterior on $(\alpha, \lambda)$
under standard regularity conditions.

Third, **exponential-family structure**. The log-density is linear in
$\alpha$, so the marginal likelihood of the data under $g$ is a
mixture-of-Gaussians density in $\alpha$ — a tractable target for
Hamiltonian Monte Carlo. The smoothness penalty enters as a
zero-mean Gaussian prior on $\alpha$ with precision $\lambda$.
The log-density is smooth in $\alpha$, and the Gaussian prior
contributes a quadratic penalty. Together with the log-sum-exp
marginal likelihood, this gives Stan a differentiable target for
HMC, but it is not a conjugate update. Together these properties
make the log-spline prior the natural compromise between the
unrealistic discreteness of the NPMLE and the rigid parametric form
of the Laplace or point-normal alternatives.

## Fully Bayesian versus empirical-Bayes

Efron's original implementation of the log-spline prior chose the
smoothness penalty by cross-validation: $\alpha$ was estimated by
penalised maximum marginal likelihood, with the penalty constant
selected on a held-out scoring rule. The procedure is fast and
delivers a point estimate $\hat g$, but the credible intervals it
produces — built around the linearised variance of $\hat\alpha$ —
condition on the estimated prior as if it were the true prior. At
finite $K$ this conditioning omits a genuine source of uncertainty,
and the resulting empirical-Bayes intervals can undercover. The
mechanism is structural: the delta-method linearisation around
$\hat\alpha$ captures first-order sampling variability in the
penalised estimator but discards both higher-order curvature and any
prior uncertainty about $\alpha$ itself.

The fully Bayesian formulation in `bayesEfron` closes this gap by
placing a hyperprior on the smoothness precision and integrating
over both layers of uncertainty. Specifically, $\alpha \mid \lambda
\sim \mathcal{N}(0, \lambda^{-1} I_M)$ and $\lambda$ is assigned a
half-Cauchy hyperprior. The half-Cauchy citation is to Gelman's
(2006, *Bayesian Analysis*) weakly informative scale-prior
rationale; the package follows Lee and Sui in placing it on the
smoothness precision $\lambda$, with
$\sigma_\alpha = \lambda^{-1/2}$ connecting the two
parameterisations. The joint posterior over
$(\alpha, \lambda)$ is sampled by Hamiltonian Monte Carlo in Stan,
and every functional of $g$ — including the per-site posterior
mean, posterior standard deviation, and posterior replicate
$\theta_i^{\mathrm{rep}}$ — inherits the integration over both
layers. The trade-off is computational: the fully Bayesian fit is
an HMC run rather than an optimisation, with corresponding
increases in wall-clock time and diagnostic burden. The
compensating benefit is that credible intervals on
$\theta_i^{\mathrm{rep}}$ are calibrated jointly over the prior and
the per-site uncertainty. Lee and Sui (2025, *Mathematics*) report
near-nominal $88$–$91\%$ coverage for $90\%$ credible intervals
under the package's fully Bayesian formulation; the package's own
Tier 3 release verification reproduces this calibration result with
aggregate `theta_rep` interval coverage in the $[0.87, 0.92]$
acceptance band on the locked Lee-Sui benchmark fixtures across
$K \in \{50, 100, 200, 500, 1500\}$. Vignette M6 documents that
evidence in full.

## A small illustration

To fix ideas, consider a bimodal mixing distribution sampled at the
default grid resolution. The figure below shows a continuous reference
density and its discretisation onto a length-$L$ grid constructed by
`make_efron_grid()`. The grid is the fixed support on which the
log-spline parameterisation operates; the discrete bars are the
target the model would aim to recover under perfect data.

```{r grid-figure, fig.cap = "Continuous bimodal mixing distribution g and its discretisation onto the length-L grid produced by make_efron_grid()."}
set.seed(2026)
theta_hat_demo <- c(seq(-1.0, 1.0, length.out = 9), -0.8, 0.6, 0.85)
sigma_demo     <- rep(0.15, length(theta_hat_demo))

g_obj <- make_efron_grid(
  theta_hat = theta_hat_demo,
  sigma     = sigma_demo,
  L         = 101L,
  M         = 6L
)

theta_grid <- g_obj$grid
g_cont <- function(t) {
  0.55 * dnorm(t, mean = -0.4, sd = 0.20) +
    0.45 * dnorm(t, mean =  0.6, sd = 0.18)
}
g_grid <- g_cont(theta_grid)
g_grid <- g_grid / sum(g_grid)
theta_dense <- seq(min(theta_grid), max(theta_grid), length.out = 401)

if (has_ggplot2) {
  df_cont <- data.frame(theta = theta_dense, density = g_cont(theta_dense))
  df_grid <- data.frame(theta = theta_grid, mass = g_grid)
  ggplot2::ggplot() +
    ggplot2::geom_segment(
      data = df_grid,
      ggplot2::aes(x = theta, xend = theta, y = 0, yend = mass),
      colour = "grey40", linewidth = 0.4
    ) +
    ggplot2::geom_line(
      data = df_cont,
      ggplot2::aes(x = theta, y = density / sum(density) *
                     length(theta_dense) / length(theta_grid)),
      colour = "black", linewidth = 0.6
    ) +
    ggplot2::labs(
      x = expression(theta),
      y = expression(g(theta)),
      title = NULL
    ) +
    ggplot2::theme_minimal(base_size = 11)
} else {
  par(mar = c(4, 4, 1, 1))
  plot(theta_grid, g_grid, type = "h", col = "grey40", lwd = 1,
       xlab = expression(theta), ylab = expression(g(theta)))
  lines(theta_dense,
        g_cont(theta_dense) / sum(g_cont(theta_dense)) *
          length(theta_dense) / length(theta_grid),
        col = "black", lwd = 1.5)
}
```

The deconvolution problem is to recover such a $g$ — or a smooth
approximation thereof representable in the log-spline basis — from a
set of noisy estimates $(\hat\theta_i, \sigma_i)$ drawn from the
convolved marginals $f_i$.

## Reading map for the methodological track

The remaining methodological vignettes carry the development through
the implementation and verification of the model sketched above.

- **M2 — The Efron log-spline prior.** Detailed treatment of the
  log-spline parameterisation: the natural-cubic-spline basis, the
  softmax map, the role of the intercept-exclusion convention in
  identifiability, and the relation between the spline dimension
  $M$ and the expressible density family.
- **M3 — The random-effects hierarchy.** Embedding of the
  log-spline prior into the four-level random-effects model:
  sampling level, latent-effect level, spline-coefficient level,
  and smoothness hyperprior. Statement and verification of the
  proper-posterior conditions.
- **M4 — Grid construction.** The four grid recipes exposed by
  `make_efron_grid()`, the trade-offs between grid width and
  resolution, and the KL-bounded experimental grid rule.
- **M5 — Stan model and cache.** The byte-locked Stan source, the
  numerical-stability disciplines (`log_sum_exp`, max-shift
  normalisation, two-pass variance), and the cache that prevents
  redundant compilation.
- **M6 — Verification and calibration.** End-to-end coverage
  verification against the benchmark distributions, the
  release-blocking acceptance band on the per-site replicate
  intervals, and the diagnostic surface of the fitted object.

A reader who follows the track end-to-end will exit with a complete
picture of the model, its implementation, and the empirical case for
the calibration claims it makes.
