---
title: "M2 · The Efron log-spline prior"
description: "Derivation, simplex normalisation, half-Cauchy hyperprior, and a smoothness sweep."
output:
  rmarkdown::html_vignette:
    fig_width: 7
    fig_height: 4.5
vignette: >
  %\VignetteIndexEntry{M2 · The Efron log-spline prior}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  echo     = TRUE,
  collapse = TRUE,
  comment  = "#>",
  message  = FALSE,
  warning  = FALSE,
  fig.width  = 7,
  fig.height = 4.5
)
library(bayesEfron)
has_ggplot2 <- requireNamespace("ggplot2", quietly = TRUE)
```

## Scope of this vignette

This is the second vignette in the methodological track. M1
established the deconvolution setting and motivated the log-spline
family as the prior on the mixing distribution $g$. M2 specifies
that prior. The goal is a complete account of the parameterisation
that the package's Stan source implements: the continuous form on
$\mathbb{R}$, the discretisation onto the grid, the softmax
normalisation that yields a proper simplex, the identifiability
convention that fixes the intercept-induced ridge, the Gaussian
conditional prior on the spline coefficients, the half-Cauchy
hyperprior on the smoothness precision, and the visual character
of the family under varying precision.

## Notation reminder

M1 introduced the symbols. They recur unchanged: $K$ is the number of
sites; $\theta_i$ is the latent effect at site $i$ and $\hat\theta_i$
its observed estimate; $\sigma_i$ is the known within-study standard
error; $g$ is the mixing distribution to be recovered; $L$ is the
grid length; $M$ is the spline dimension; $\alpha \in \mathbb{R}^M$ is
the spline-coefficient vector; and $\lambda$ is the smoothness
precision that governs the Gaussian prior on $\alpha$. No new symbols
are introduced in this vignette except auxiliary intermediates used in
the derivations.

## The continuous form

On the real line, the log-spline prior writes the unnormalised
log-density of $g$ at $\theta$ as a linear function of $\alpha$:
$$
\log \tilde g_{\alpha}(\theta) \;=\; B(\theta)^{\top} \alpha,
\qquad
g_{\alpha}(\theta) \;\propto\; \exp\!\big(B(\theta)^{\top}\alpha\big),
$$
where $B(\theta) = (B_1(\theta), \ldots, B_M(\theta))^{\top}$ is the
length-$M$ natural-cubic-spline basis vector evaluated at $\theta$.
The basis functions $B_k(\cdot)$ are fixed by choosing knot locations
and the degree of the polynomial pieces; only $\alpha$ is a
parameter.

The exponential parameterisation is not a cosmetic choice. Three of
its properties shape every downstream computation in the package.

First, the map $\alpha \mapsto g_{\alpha}$ is unconstrained on
$\mathbb{R}^M$. Replacing the linear form $B(\theta)^{\top}\alpha$
with $g_{\alpha}(\theta)$ directly would require sign constraints to
keep $g$ non-negative, which is awkward both analytically and
algorithmically. The $\exp(\,\cdot\,)$ wrapper enforces positivity
identically in $\alpha$, so $\alpha$ can range freely over
$\mathbb{R}^M$ without boundary handling.

Second, the family $\{g_{\alpha} : \alpha \in \mathbb{R}^M\}$ is an
exponential family with natural parameter $\alpha$ and sufficient
statistic $B(\theta)$. Its log-density is linear in $\alpha$, and the
log-partition function
$$
\phi(\alpha) \;=\; \log \int \exp\!\big(B(\theta)^{\top}\alpha\big)\,
\mathrm{d}\theta
$$
is convex on $\mathbb{R}^M$ (this is a standard consequence of
Hölder's inequality and is recorded in the package's derivation
appendix). Convexity of $\phi$ supplies stable derivatives for the
normalised log-spline density. HMC performance on the $\alpha$
block at fixed $\lambda$ is therefore an empirical diagnostic
question, not a consequence of global log-concavity: the marginal
likelihood under the deconvolution mixture is a log of a weighted
Gaussian mixture, which is not generally log-concave in $\alpha$.

Third, the gradient of the log-density with respect to $\alpha$ is
just $B(\theta) - \nabla\phi(\alpha)$, and $\nabla\phi(\alpha)$ is the
expectation of $B(\Theta)$ under $g_{\alpha}$. Reverse-mode autodiff
in Stan computes this gradient at every leapfrog step in a single
matrix-vector multiplication of cost $O(LM)$.

## Discretisation onto a grid

`bayesEfron` does not work with the continuous form directly. Instead
the package fixes a length-$L$ grid $\{\theta_1, \ldots, \theta_L\}
\subset \mathbb{R}$ chosen before sampling, replaces the integral in
$\phi$ with a sum, and treats $g$ as a probability mass on the grid.
The discrete prior is
$$
g_l(\alpha) \;=\;
\frac{\exp\!\big(B_l^{\top}\alpha\big)}
     {\sum_{j=1}^{L} \exp\!\big(B_j^{\top}\alpha\big)},
\qquad l = 1, \ldots, L,
$$
where $B_l \in \mathbb{R}^M$ is the $l$-th row of the basis matrix
$B \in \mathbb{R}^{L \times M}$ evaluated at the grid:
$B = \mathrm{ns}(\theta_1, \ldots, \theta_L;\ \mathrm{df} = M,\
\mathrm{intercept} = \mathrm{FALSE})$. The denominator is the
finite-sum log-partition
$$
\phi(\alpha) \;=\;
\log \sum_{j=1}^{L} \exp\!\big(B_j^{\top}\alpha\big)
\;=\; \mathrm{log\_sum\_exp}(B\alpha),
$$
and the discrete prior is the softmax of $B\alpha$:
$g(\alpha) = \mathrm{softmax}(B\alpha)$, lying in the simplex
$\Delta^{L-1}$ by construction.

Two properties of this discretisation are worth stating explicitly.

The normalisation is **exact** on the grid, not approximate. The
continuous-form integral $\int \exp(B(\theta)^{\top}\alpha)\,
\mathrm{d}\theta$ would itself need to be approximated by a
quadrature rule, but the discrete model has no integral to
approximate: it replaces the continuum by the grid and normalises the
mass on the grid exactly. The relation between the discrete prior
and the continuous form is the usual Riemann-sum limit
$\sum_{j:\theta_j \le t} g_j \to \int_{-\infty}^{t} g(\theta)\,
\mathrm{d}\theta$ as the grid spacing tends to zero, but the
**inferential** target is the discrete prior $g$; vignette M4
discusses the grid-resolution trade-off in detail.

The Stan source computes $\log g = \log \mathrm{softmax}(B\alpha) =
B\alpha - \phi(\alpha)\,\mathbf{1}_L$ via Stan's built-in
`log_softmax`, which is implemented using the max-shift identity to
avoid overflow when $B\alpha$ spans tens of orders of magnitude. The
quantity $\log g$ is what enters the mixture log-likelihood (M3);
the simplex $g = \exp(\log g)$ is what enters the prior-summary
generated quantities.

## Identifiability of the spline coefficients

The softmax is invariant to a constant shift of its argument:
$$
\mathrm{softmax}(B\alpha + c\,\mathbf{1}_L) \;=\;
\mathrm{softmax}(B\alpha)
$$
for any $c \in \mathbb{R}$. If the basis matrix $B$ has the constant
vector $\mathbf{1}_L$ in its column space — that is, if there exists
$v_0 \in \mathbb{R}^M$ with $B v_0 = \mathbf{1}_L$ — then $\alpha$
and $\alpha + t v_0$ produce identical $g$ for every $t$, and the
joint posterior is degenerate along that direction. The package
handles this two ways.

The first is parameterisation. The basis matrix is constructed with
`splines::ns(grid, df = M, intercept = FALSE)`, where
`intercept = FALSE` instructs `splines::ns()` to omit the constant
column from the natural-cubic-spline basis. The resulting $B$ does
not include the constant direction by construction, so the most
direct form of the shift invariance is removed at the basis
construction step.

The second is a runtime guard. Whether the columns of $B$ span the
orthogonal complement of $\mathbf{1}_L$ — and hence whether
$\mathbf{1}_L \notin \mathrm{col}(B)$ — is not an automatic algebraic
consequence of `intercept = FALSE` for every admissible grid
configuration. The package therefore enforces a rank postcondition
in `make_efron_grid()`: the augmented matrix $[\mathbf{1}_L \;\;\; B]$
must have rank $M + 1$. The check is `qr(cbind(1, B), tol =
sqrt(.Machine$double.eps))$rank == M + 1L`. Failure raises the
typed condition `bef_grid_rank_deficient` before any Stan
compilation occurs. With the rank postcondition satisfied, $\alpha$
is identified on $\mathbb{R}^M$ and the joint posterior on
$(\alpha, \lambda)$ is proper.

## The smoothness hyperprior

The conditional prior on $\alpha$ is a centred Gaussian whose
variance is parameterised by the inverse of $\lambda$:
$$
\alpha \mid \lambda \;\sim\;
\mathcal{N}\!\left(\mathbf{0},\, \lambda^{-1}\,\mathbf{I}_M\right).
$$
The convention is variance-form: the second argument of
$\mathcal{N}(\mu, \sigma^2)$ is variance, so each component of
$\alpha$ has variance $1/\lambda$ and standard deviation
$\lambda^{-1/2}$. The Stan source declares the same prior in
standard-deviation form as `alpha ~ normal(0, inv_sqrt(lambda))`,
which Stan's `normal(mu, sd)` interprets identically.

The role of $\lambda$ is shrinkage on $\alpha$. Larger $\lambda$
concentrates $\alpha$ near zero, so $B\alpha$ is small for every
$\theta$, the softmax is close to uniform, and $g$ is smooth and
flat. Smaller $\lambda$ admits larger $\alpha$, so $B\alpha$ can vary
substantially across the grid and the resulting $g$ can exhibit
peaks, asymmetry, and modest multimodality.

Because $\lambda$ controls the effective smoothness of $g$ but is
itself unknown, the package places a hyperprior on it. The choice is
the half-Cauchy:
$$
\lambda \;\sim\; \mathrm{Cauchy}^{+}(0, 5),
$$
a half-Cauchy on $\lambda$ with scale parameter $5$. Two properties
of the half-Cauchy commend it for this role, following the
weakly-informative-prior tradition canonically framed by Gelman
(2006, *Bayesian Analysis*).

The first is heavy tails. The half-Cauchy has no moments, so neither
its mean nor its variance is defined; the implicit prior weight at
large $\lambda$ decays only as $\lambda^{-2}$. This is heavy enough
that the data can override the prior when shape evidence is strong,
pulling $\lambda$ either upward (more smoothing) or downward (less
smoothing) without resistance from prior-induced curvature. The
heavy upper tail in particular means that the prior does not force
$\lambda$ to be small even when the data are uninformative; the
prior admits "smooth" as a default consistent with weak data.

The second is mass at small $\lambda$. The half-Cauchy assigns
non-negligible density to small positive values of $\lambda$,
allowing the conditional Gaussian on $\alpha$ to be wide when the
data warrant a wiggly $g$. A half-normal alternative would have a
lighter upper tail, placing less prior mass on very large $\lambda$
than the half-Cauchy. Near zero, both priors can assign
non-negligible density.

The combination of the proper conditional prior on $\alpha$ given
$\lambda$ and the proper hyperprior on $\lambda$ is sufficient,
together with the bounded mixture likelihood (M3), for a proper
joint posterior on $(\alpha, \lambda)$. The package documents the
proper-posterior conditions in its derivation appendix; the
appropriate forward reference is M3.

The Cauchy citation here is to Gelman's weakly-informative-prior
rationale rather than to a specific prescription that the
half-Cauchy belongs on a precision parameter — Gelman's exposition
canonically places the half-Cauchy on a scale parameter, e.g., a
hierarchical-Gaussian standard deviation. The half-Cauchy on
$\lambda$ used here is paper-faithful to the published Lee & Sui
(2025) implementation, and the bijection $\sigma_\alpha =
\lambda^{-1/2}$ relates the two parameterisations.

## A smoothness sweep

The visual character of the family $\{g_{\alpha}\}$ under varying
implicit smoothness is most easily appreciated by sampling $\alpha$
from the conditional prior at several values of $\lambda$ and
plotting the corresponding $g$. The figure below shows three such
draws on the package's default `paper_realdata` grid. The standard
deviations correspond to $\lambda \in \{1, 4, 100\}$ — high-variance
(loose) at $\lambda = 1$, moderate at $\lambda = 4$, and tight at
$\lambda = 100$.

```{r sweep-figure, fig.cap = "Three log-spline priors g_alpha(theta) at decreasing alpha variance: low precision (lambda = 1, top), medium precision (lambda = 4, middle), and high precision (lambda = 100, bottom). Higher precision pulls the density toward the spline-uniform reference; lower precision admits multimodality."}
set.seed(2026)

grid_obj <- make_efron_grid(
  theta_hat = seq(-1.0, 1.0, length.out = 9),
  sigma     = rep(0.15, 9),
  L         = 101L,
  M         = 6L
)
theta_grid <- grid_obj$grid
B          <- grid_obj$B
M          <- grid_obj$M

log_sum_exp <- function(x) {
  m <- max(x)
  m + log(sum(exp(x - m)))
}
g_alpha <- function(alpha) {
  log_w <- as.numeric(B %*% alpha)
  exp(log_w - log_sum_exp(log_w))
}

alpha_low  <- rnorm(M, sd = 1.0)   # lambda = 1
alpha_med  <- rnorm(M, sd = 0.5)   # lambda = 4
alpha_high <- rnorm(M, sd = 0.1)   # lambda = 100
g_low  <- g_alpha(alpha_low)
g_med  <- g_alpha(alpha_med)
g_high <- g_alpha(alpha_high)

if (has_ggplot2) {
  df_sweep <- rbind(
    data.frame(theta = theta_grid, mass = g_low,
               panel = "lambda = 1 (loose)"),
    data.frame(theta = theta_grid, mass = g_med,
               panel = "lambda = 4 (moderate)"),
    data.frame(theta = theta_grid, mass = g_high,
               panel = "lambda = 100 (tight)")
  )
  df_sweep$panel <- factor(
    df_sweep$panel,
    levels = c("lambda = 1 (loose)",
               "lambda = 4 (moderate)",
               "lambda = 100 (tight)")
  )
  ggplot2::ggplot(df_sweep,
                  ggplot2::aes(x = theta, y = mass)) +
    ggplot2::geom_line(colour = "black", linewidth = 0.5) +
    ggplot2::facet_wrap(~ panel, ncol = 1, scales = "free_y") +
    ggplot2::labs(
      x = expression(theta),
      y = expression(g[alpha](theta)),
      title = NULL
    ) +
    ggplot2::theme_minimal(base_size = 11)
} else {
  par(mfrow = c(3, 1), mar = c(4, 4, 1.5, 1))
  plot(theta_grid, g_low,  type = "l", lwd = 1,
       xlab = expression(theta), ylab = expression(g[alpha](theta)),
       main = "lambda = 1 (loose)")
  plot(theta_grid, g_med,  type = "l", lwd = 1,
       xlab = expression(theta), ylab = expression(g[alpha](theta)),
       main = "lambda = 4 (moderate)")
  plot(theta_grid, g_high, type = "l", lwd = 1,
       xlab = expression(theta), ylab = expression(g[alpha](theta)),
       main = "lambda = 100 (tight)")
}
```

The three panels exhibit the expected behaviour. At $\lambda = 100$
(bottom), $\alpha$ is tightly shrunk toward zero, $B\alpha$ is small
in magnitude, and the softmax sits close to the spline-uniform
reference: the density is broad, gently sloped, and lacks distinct
modes. At $\lambda = 4$ (middle), some shape structure emerges but
the density remains unimodal or nearly so. At $\lambda = 1$ (top),
$\alpha$ has substantial dispersion, $B\alpha$ varies markedly across
the grid, and the resulting $g$ shows asymmetry and multimodality.
The contrast across the panels is what motivates the data-driven
choice of $\lambda$: the family is rich enough to express both
near-uniform and visibly multimodal priors, and the half-Cauchy
hyperprior lets the marginal likelihood select among them rather
than fixing a single smoothness level by hand.

## The four-level Bayesian hierarchy

Composing the components specified above yields the full
hierarchical model on $K$ sites. Reading from the data upward:
$$
\begin{aligned}
\hat\theta_i \mid \theta_i &\;\sim\;
   \mathcal{N}\!\left(\theta_i, \sigma_i^2\right),
   \qquad i = 1, \ldots, K, \\
\theta_i \mid g &\;\stackrel{\mathrm{iid}}{\sim}\; g, \\
g_{\alpha}(\theta_l) &\;=\;
   \frac{\exp\!\big(B_l^{\top}\alpha\big)}
        {\sum_{j=1}^{L} \exp\!\big(B_j^{\top}\alpha\big)},
   \qquad l = 1, \ldots, L, \\
\alpha \mid \lambda &\;\sim\;
   \mathcal{N}\!\left(\mathbf{0},\, \lambda^{-1}\,\mathbf{I}_M\right), \\
\lambda &\;\sim\; \mathrm{Cauchy}^{+}(0, 5).
\end{aligned}
$$
The four levels are: the sampling model that relates the observed
estimates to the latent effects (level 1); the categorical draw of
the latent effects from the discrete prior $g$ on the grid (level 2);
the log-spline parameterisation of $g$ in terms of the spline
coefficients $\alpha$ (level 3); and the smoothness hyperprior on
$\lambda$ that regularises $\alpha$ (level 4).

The latent $\theta_i$ are integrated out of the joint analytically:
the per-site marginal likelihood is the finite mixture
$p(\hat\theta_i \mid g, \sigma_i) = \sum_{j=1}^{L} g_j\,
\mathcal{N}(\hat\theta_i;\, \theta_j, \sigma_i^2)$, and the sampler
runs over the unconstrained pair $(\alpha, \lambda)$ in dimension
$M + 1$ (default $7$) regardless of $K$. Vignette M3 unpacks the
hierarchy level by level, derives the marginal likelihood, and
characterises the small-$K$ prior-dominance regime that follows from
the heavy-tailed hyperprior on $\lambda$.

## Computational structure for HMC

The log-posterior of $(\alpha, \lambda)$ under the hierarchy above
is everywhere differentiable. The log-spline parameterisation
contributes a term linear in $\alpha$ (through $B\alpha$) plus the
log-partition $\phi(\alpha)$, whose gradient is $\nabla\phi(\alpha) =
\sum_j g_j(\alpha)\, B_j$ and whose Hessian is the covariance matrix
of $B_J$ under $g_{\alpha}$ — both available in closed form. The
conditional prior on $\alpha$ contributes a quadratic-in-$\alpha$
term, and the half-Cauchy on $\lambda$ contributes a smooth
$\log(1 + \lambda^2/25)$ term. Combined with the mixture log-sum-exp
of the likelihood (M3), the entire target is twice continuously
differentiable in $(\alpha, \lambda)$, and reverse-mode autodiff in
Stan produces the gradient at cost $O(KL + LM)$ per leapfrog step.

This differentiability is the algorithmic reason the log-spline
prior is HMC-friendly, and it is the reason `bayesEfron` is
implemented in Stan rather than in a Metropolis-style sampler. The
Stan source `inst/stan/efron_re.stan` exploits the structure via
Stan's `log_softmax`, `log_sum_exp`, and `normal` primitives, all of
which are differentiable through reverse-mode autodiff. Vignette M5
walks through the Stan source line by line and documents the ten
numerical-stability disciplines that the source obeys.

## Reading map

This vignette specified the prior. The remaining methodological
vignettes carry the construction into implementation and
verification.

- **M3 — The random-effects hierarchy.** Unpacks the four levels,
  derives the per-site mixture likelihood, characterises the
  per-site posterior weights via the max-shift trick, distinguishes
  `theta_mean` (estimator) from `theta_rep` (parameter draw), and
  states the proper-posterior conditions.
- **M4 — Grid construction.** Documents the four grid recipes
  exposed by `make_efron_grid()` — `paper_realdata`,
  `paper_simulation`, `paper_sensitivity`, and the experimental
  KL-target recipe — and the trade-offs between grid width and
  resolution.
- **M5 — Stan model and cache.** The byte-locked Stan source,
  the numerical-stability disciplines (`log_softmax`,
  `log_sum_exp`, max-shift normalisation, two-pass variance), and
  the cache that prevents redundant compilation.
- **M6 — Verification and calibration.** End-to-end coverage
  verification on the locked benchmark fixtures and the
  release-blocking acceptance band on the per-site replicate
  intervals.

A reader who has followed M1 and M2 has the complete model
specification in hand; the remaining vignettes connect that
specification to its computational realisation and its empirical
calibration.
