---
title: "M3 · Bayesian random-effects hierarchy"
description: "The four-level hierarchy, posterior factorisation, and the package's metadata payload."
output:
  rmarkdown::html_vignette:
    fig_width: 7
    fig_height: 4.5
vignette: >
  %\VignetteIndexEntry{M3 · Bayesian random-effects hierarchy}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  echo     = TRUE,
  collapse = TRUE,
  comment  = "#>",
  message  = FALSE,
  warning  = FALSE,
  fig.width  = 7,
  fig.height = 4.5
)
library(bayesEfron)
```

## Scope of this vignette

M1 motivated the empirical-Bayes deconvolution problem and M2 specified
the log-spline prior on the mixing distribution $g$. M3 composes those
pieces into the full four-level Bayesian hierarchy that `bayesEfron` v0.1
fits. The vignette has four jobs. First, it states the hierarchy as a
single equation block so the entire generative model is visible at one
glance. Second, it draws the corresponding directed acyclic graph and
walks through the conditional dependencies. Third, it derives the
marginal of inferential interest — the per-site posterior $p(\theta_i
\mid \hat{\boldsymbol\theta})$ — and explains how the package's
generated quantities relate to that marginal and to functionals of $g$.
Fourth, it maps each of the thirteen named fields of `fit$metadata` back
to the node in the DAG from which it originates.

## Notation reminder

The notation is unchanged from M1 and M2. $K$ is the number of sites;
$\theta_i$ is the latent effect at site $i$ and $\hat\theta_i$ its
observed estimate; $\sigma_i$ is the known within-study standard error;
$g$ is the discrete mixing distribution on a fixed length-$L$ grid
$\{\theta_1, \ldots, \theta_L\}$; $B \in \mathbb{R}^{L \times M}$ is the
fixed natural-cubic-spline basis evaluated at the grid; $\alpha \in
\mathbb{R}^M$ is the spline-coefficient vector that determines $g$
through the softmax map of M2; and $\lambda > 0$ is the smoothness
precision that governs the Gaussian conditional prior on $\alpha$. No
new symbols are introduced.

## The four-level hierarchy

The full generative model for $K$ sites is, reading from the data
upward:

$$
\begin{aligned}
\text{Level 1 (likelihood):} \quad
\hat\theta_i \mid \theta_i
  &\;\sim\; \mathcal{N}\!\left(\theta_i, \sigma_i^2\right),
  \qquad i = 1, \ldots, K, \\
\text{Level 2 (latent effects):} \quad
\theta_i \mid g
  &\;\stackrel{\mathrm{iid}}{\sim}\; g, \\
\text{Level 3 (mixing-distribution prior):} \quad
g_{\alpha}(\theta_l)
  &\;=\; \frac{\exp\!\big(B_l^{\top}\alpha\big)}
            {\sum_{j=1}^{L} \exp\!\big(B_j^{\top}\alpha\big)},
  \quad l = 1, \ldots, L, \\
\text{Level 4 (hyperprior):} \quad
\alpha \mid \lambda
  &\;\sim\; \mathcal{N}\!\left(\mathbf{0},\, \lambda^{-1}\mathbf{I}_M\right), \\
\lambda
  &\;\sim\; \mathrm{Cauchy}^{+}(0, 5).
\end{aligned}
$$

Three points about the level structure are worth recording before
discussing the DAG.

The first concerns the role of the grid. The grid
$\boldsymbol{\theta} = (\theta_1, \ldots, \theta_L)$ at Level 3 is fixed
before sampling by `make_efron_grid()` (vignette M4) and is **not** a
parameter of the model. The basis matrix $B$ is likewise fixed, being
$B = \mathrm{ns}(\boldsymbol\theta;\ \mathrm{df} = M,\ \mathrm{intercept}
= \mathrm{FALSE})$. The Level 3 line above therefore specifies the
deterministic map $\alpha \mapsto g$ rather than a stochastic step:
conditional on $\alpha$, the prior $g$ is a function, not a draw.

The second is that the latent $\theta_i$ at Level 2 are integrated out
analytically. The per-site marginal likelihood,
$$
p(\hat\theta_i \mid g, \sigma_i)
  \;=\; \sum_{j=1}^{L} g_j\, \mathcal{N}\!\left(\hat\theta_i;\, \theta_j, \sigma_i^2\right),
$$
is a finite mixture of $L$ Gaussians, and it is this quantity that
enters the Stan model block. The Hamiltonian Monte Carlo sampler runs
over the unconstrained pair $(\alpha, \lambda)$ in dimension $M + 1$
(default $7$) regardless of $K$. The per-site posteriors on $\theta_i$
are recovered in the generated-quantities block after sampling.

The third is the variance-versus-standard-deviation convention on
$\alpha \mid \lambda$. The display above writes the conditional prior in
**variance** form, so each component of $\alpha$ has variance $1/\lambda$
and standard deviation $\lambda^{-1/2}$. The Stan source declares the
same prior in standard-deviation form (`alpha ~ normal(0,
inv_sqrt(lambda))`); the two are identical. The reference paper for
the package's prior writes the conditional prior in
standard-deviation form, which the Stan source retains for
paper-faithfulness. All three notations describe the same
distribution.

## DAG diagram

The directed acyclic graph corresponding to the hierarchy is shown
below. Each node is a random or fixed quantity; each arrow encodes a
conditional dependence of the head node on the tail node.

```{r dag-figure, fig.cap = "Directed acyclic graph of the four-level Bayesian random-effects hierarchy fit by bayesEfron. The smoothness precision lambda controls the Gaussian conditional prior on the spline coefficients alpha; alpha and the fixed basis B together determine the discrete mixing distribution g; latent site effects theta_i are drawn iid from g; observed estimates hat_theta_i are Gaussian around theta_i with known standard errors sigma_i. Random quantities are shown as circles, fixed inputs as squares. The plate over i = 1, ..., K indicates K independent replicates of the per-site sub-graph.", fig.width = 7, fig.height = 4.5}
old_par <- par(no.readonly = TRUE)
on.exit(par(old_par), add = TRUE)
par(mar = c(0.5, 0.5, 0.5, 0.5))

plot.new()
plot.window(xlim = c(0, 10), ylim = c(0, 6.5), asp = 1)

draw_circle <- function(x, y, r = 0.45, lwd = 1.3) {
  ang <- seq(0, 2 * pi, length.out = 200)
  polygon(x + r * cos(ang), y + r * sin(ang), border = "black",
          col = "white", lwd = lwd)
}
draw_square <- function(x, y, r = 0.45, lwd = 1.3) {
  polygon(x + c(-r, r, r, -r), y + c(-r, -r, r, r),
          border = "black", col = "white", lwd = lwd)
}

# Coordinates: top to bottom.
x_lambda <- 2.0; y_lambda <- 5.7
x_alpha  <- 4.0; y_alpha  <- 5.7
x_g      <- 6.0; y_g      <- 5.7
x_B      <- 8.0; y_B      <- 5.7

x_theta  <- 5.0; y_theta  <- 3.3
x_hat    <- 5.0; y_hat    <- 1.1
x_sigma  <- 7.5; y_sigma  <- 1.1

# Plate enclosing the per-site sub-graph.
plate_x <- c(3.6, 8.4, 8.4, 3.6)
plate_y <- c(0.3, 0.3, 4.1, 4.1)
polygon(plate_x, plate_y, border = "grey55", lty = 2, lwd = 1)
text(8.25, 0.55, expression(italic(i) == 1 * "," ~ ldots * "," ~ italic(K)),
     cex = 0.85, col = "grey35", adj = c(1, 0))

# Nodes.
draw_circle(x_lambda, y_lambda)
text(x_lambda, y_lambda, expression(lambda), cex = 1.25)

draw_circle(x_alpha,  y_alpha)
text(x_alpha,  y_alpha,  expression(alpha),  cex = 1.25)

draw_circle(x_g, y_g)
text(x_g, y_g, expression(italic(g)), cex = 1.25)

draw_square(x_B, y_B)
text(x_B, y_B, expression(italic(B)), cex = 1.1)

draw_circle(x_theta, y_theta)
text(x_theta, y_theta, expression(theta[italic(i)]), cex = 1.1)

draw_circle(x_hat, y_hat)
text(x_hat, y_hat, expression(hat(theta)[italic(i)]), cex = 1.1)

draw_square(x_sigma, y_sigma)
text(x_sigma, y_sigma, expression(sigma[italic(i)]), cex = 1.1)

# Arrows.
arrow_between <- function(x0, y0, x1, y1, r = 0.45, length = 0.10) {
  dx <- x1 - x0; dy <- y1 - y0; d <- sqrt(dx^2 + dy^2)
  ux <- dx / d;  uy <- dy / d
  arrows(x0 + r * ux, y0 + r * uy, x1 - r * ux, y1 - r * uy,
         length = length, angle = 22, lwd = 1.2, col = "black")
}
arrow_between(x_lambda, y_lambda, x_alpha, y_alpha)
arrow_between(x_alpha,  y_alpha,  x_g,     y_g)
arrow_between(x_B,      y_B,      x_g,     y_g)
arrow_between(x_g,      y_g,      x_theta, y_theta)
arrow_between(x_theta,  y_theta,  x_hat,   y_hat)
arrow_between(x_sigma,  y_sigma,  x_hat,   y_hat)

# Legend.
legend(
  "topleft", bty = "n", cex = 0.8,
  legend = c("random quantity", "fixed input"),
  pch    = c(1, 0), pt.cex = 1.3
)
```

The diagram resolves the conditional dependence structure of the joint
distribution. The hyperprior on $\lambda$ controls the conditional
Gaussian prior on $\alpha$; the spline coefficients $\alpha$, together
with the fixed basis $B$, deterministically produce the discrete prior
$g$; iid draws from $g$ produce the latent site effects $\theta_i$; and
each observed estimate $\hat\theta_i$ is Gaussian around its latent
counterpart $\theta_i$ with the known standard error $\sigma_i$. The
plate captures the conditional independence of the $K$ per-site
sub-graphs given the upper-level quantities $(\lambda, \alpha, g)$:
$\theta_i$ and $\theta_{i'}$ are conditionally independent given $g$,
and $\hat\theta_i$ and $\hat\theta_{i'}$ are conditionally independent
given $(\theta_i, \theta_{i'})$ and the corresponding standard errors.
Random quantities are circles; fixed inputs ($\sigma_i$ supplied by the
user, $B$ constructed by `make_efron_grid()`) are squares. The graph is
acyclic and the joint factorises as $p(\lambda)\, p(\alpha \mid
\lambda)\, \prod_i p(\theta_i \mid g)\, p(\hat\theta_i \mid \theta_i,
\sigma_i)$.

## The marginal of inferential interest

The hierarchy is fit so that the per-site posterior $p(\theta_i \mid
\hat{\boldsymbol\theta})$ — and its functionals — can be summarised. At
fixed $g$ the per-site posterior factorises by Bayes's rule applied to
the convolution:
$$
p(\theta_i \mid \hat\theta_i, \sigma_i, g)
  \;=\; \frac{p(\hat\theta_i \mid \theta_i, \sigma_i)\; g(\theta_i)}
             {p(\hat\theta_i \mid \sigma_i, g)},
$$
with denominator the per-site marginal likelihood
$p(\hat\theta_i \mid \sigma_i, g) = \sum_j g_j\, \mathcal{N}(\hat\theta_i;
\theta_j, \sigma_i^2)$ already encountered above. On the grid this
specialises to a discrete posterior on $\{\theta_1, \ldots, \theta_L\}$
with weights
$$
w_{ij}
  \;\propto\;
  \exp\!\big\{\log g_j + \log \mathcal{N}(\hat\theta_i;\, \theta_j, \sigma_i^2)\big\},
  \qquad \sum_j w_{ij} = 1.
$$
The Stan source computes $w_{ij}$ in the generated-quantities block
using the max-shift normalisation discussed in M5 to prevent
underflow when individual log-components differ by tens of units.

The crucial subtlety is that the package does **not** report $p(\theta_i
\mid \hat\theta_i, \sigma_i, g)$ at a fixed $g$. It reports the marginal
posterior
$$
p(\theta_i \mid \hat{\boldsymbol\theta})
  \;=\; \int p(\theta_i \mid \hat\theta_i, \sigma_i, g(\alpha))\;
        p(\alpha, \lambda \mid \hat{\boldsymbol\theta})\,
        \mathrm{d}\alpha\,\mathrm{d}\lambda,
$$
which integrates the at-fixed-$g$ posterior over the joint posterior of
$(\alpha, \lambda)$. The integration is performed Monte Carlo via the
HMC draws: at each iteration $s = 1, \ldots, S$ the sampler produces
$(\alpha^{(s)}, \lambda^{(s)})$, the deterministic map yields
$g^{(s)} = \mathrm{softmax}(B \alpha^{(s)})$, the per-site weights
$w_{ij}^{(s)}$ are recomputed at $g^{(s)}$, and the per-site posterior
summaries (mean, standard deviation, MAP, replicate) are computed at
that iteration's $g^{(s)}$. Posterior summaries on $\theta_i$ are then
Monte Carlo averages or quantiles over the $S$ iterations. The
integration over $(\alpha, \lambda)$ is therefore implicit in the
across-iteration aggregation.

The shrinkage interpretation of the per-site posterior follows from the
weights formula. Each $w_{ij}$ is the product of the prior mass $g_j$
at grid point $\theta_j$ and the Gaussian sampling kernel
$\mathcal{N}(\hat\theta_i; \theta_j, \sigma_i^2)$, normalised across
$j$. When $\sigma_i$ is small the kernel is sharp at
$\hat\theta_i$ and the weights concentrate near grid points close to
$\hat\theta_i$, so the posterior on $\theta_i$ sits close to
$\hat\theta_i$ regardless of $g$. When $\sigma_i$ is large the kernel
is broad and the prior $g$ dominates the weights, so the posterior on
$\theta_i$ is pulled toward the mass of $g$. The strength of shrinkage
toward $g$ is therefore site-specific and continuous in $\sigma_i$,
which is the standard Robbins-Efron empirical-Bayes shrinkage pattern
adapted to a heteroscedastic sampling design.

## Functionals of $g$

The package also reports posterior summaries of two functionals of $g$:
its mean and its variance. At a single iteration $s$ with $g^{(s)} =
\mathrm{softmax}(B \alpha^{(s)})$, these functionals are deterministic
sums on the grid:
$$
\mathbb{E}_{g^{(s)}}[\theta]
  \;=\; \sum_{l=1}^{L} \theta_l\, g_l^{(s)},
  \qquad
  \mathrm{Var}_{g^{(s)}}[\theta]
  \;=\; \sum_{l=1}^{L} \big(\theta_l - \mathbb{E}_{g^{(s)}}[\theta]\big)^2\, g_l^{(s)}.
$$
The standard deviation is $\sqrt{\mathrm{Var}_{g^{(s)}}[\theta]}$. Stan
computes these as the generated quantities `mean_g`, `var_g`, and
`sd_g`, respectively, using the two-pass form $\mathbb{E}[\theta^2] -
\mathbb{E}[\theta]^2$ on the grid (vignette M5 documents the numerical
discipline).

The metadata fields `mean_g_summary`, `var_g_summary`, and
`sd_g_summary` carry posterior summaries of these draw-wise quantities
across the $S$ HMC iterations. A posterior median of `mean_g_summary` is
therefore the median across iterations of $\sum_l \theta_l\, g_l^{(s)}$
— an expectation under the posterior of $g$, not a property of any
single $g$-curve. Two facts about this construction deserve emphasis.

First, the package does not return a posterior summary of an arbitrary
functional $T(g)$. The three functionals it does return (the first
moment, the variance, and the standard deviation) are sufficient for
the calibration claims the package makes and parsimonious enough to fit
in the 13-field metadata contract; richer functionals (quantiles of
$g$, tail probabilities, density values at user-specified anchors) are
available through `fit$draws` for users willing to compute them
externally.

Second, the integration over the joint posterior of $(\alpha, \lambda)$
is the load-bearing distinction between the package's reports of $g$
and a fixed-$g$ plug-in. The plug-in summaries from an empirical-Bayes
estimator condition on $\hat g = g_{\hat\alpha}$ as if it were known;
the posterior summaries here propagate the uncertainty in
$(\alpha, \lambda)$ into every functional of $g$ that the package
returns.

## The 13-field metadata contract from the hierarchy perspective

The `fit$metadata` list has exactly thirteen named fields and four
durable attributes. The named fields are
`model_family`, `grid_method`, `seed`, `cmdstan_version`,
`stan_file_sha256`, `data_list`, `runtime_seconds`,
`mean_g_summary`, `var_g_summary`, `theta_summary`,
`theta_rep_draws`, `effective_params_summary`, and
`log_marginal_likelihood_summary`. The durable attributes are
`sd_g_summary`, `diagnostics`, `diagnostic_skipped`, and
`sampler_diagnostics_failed`. The provenance from the DAG nodes is
direct.

**Two fields are per-site posterior summaries on $\theta_i$**,
derived from the marginal $p(\theta_i \mid \hat{\boldsymbol\theta})$
of the Level 1 + Level 2 sub-graph after integration over
$(\alpha, \lambda)$:

- `theta_summary` — posterior point summaries (mean, standard
  deviation, posterior intervals) of $\theta_i$ for each
  $i = 1, \ldots, K$.
- `theta_rep_draws` — full draws of the posterior predictive
  replicate $\theta_i^{\mathrm{rep}}$ (see next section), retained
  because the release-blocking coverage diagnostic is built from
  across-draw quantiles of these replicates rather than from
  across-draw quantiles of the posterior mean estimator.

**Two named fields plus one attribute are posterior summaries of
functionals of $g$**, derived from the Level 3 sub-graph after
integration over $(\alpha, \lambda)$:

- `mean_g_summary` — posterior summary of $\mathbb{E}_g[\theta]$
  (named field).
- `var_g_summary` — posterior summary of $\mathrm{Var}_g[\theta]$
  (named field).
- `sd_g_summary` — posterior summary of $\mathrm{SD}_g[\theta]$,
  stored as a durable attribute of `fit$metadata`. The attribute
  layout is documented on `?bayes_efron_fit`.

**Two fields are model-quality diagnostics** on the joint
posterior:

- `effective_params_summary` — posterior summary of the diagnostic
  effective-parameter count $p_{\mathrm{eff}} = K - \sum_i
  \mathrm{Var}[\theta_i \mid \hat\theta_i, g] / \sigma_i^2$. The
  quantity is exposed for inspection only; vignette M5 discusses
  the $\sigma_i \to 0$ amplification that limits its usefulness as
  a model-comparison statistic.
- `log_marginal_likelihood_summary` — posterior summary of $\sum_i
  \log p(\hat\theta_i \mid g, \sigma_i)$, the data's log-marginal
  under the per-iteration $g$. Future versions (v0.2+) will build a
  publication-ready LOO-CV facility on top of this quantity.

**Seven fields are reproducibility and bookkeeping metadata**
rather than nodes in the DAG:

- `model_family` — the model identifier (`"RE"` for v0.1).
- `grid_method` — the grid recipe used to construct
  $\boldsymbol\theta$ and $B$ (vignette M4).
- `data_list` — the seven-field Stan data block that was actually
  sent to CmdStan (`K`, `theta_hat`, `sigma`, `L`, `grid`, `M`,
  `B`).
- `runtime_seconds` — sampler wall-clock.
- `seed`, `cmdstan_version`, `stan_file_sha256` — the integer seed
  used for the run (auto-generated if `seed = NULL` at the call
  site), the CmdStan version string, and the SHA-256 of the
  byte-locked Stan source. Together with `data_list` and a matching
  toolchain environment, these support practical re-play of the
  fit.

The remaining three durable attributes (`diagnostics`,
`diagnostic_skipped`, `sampler_diagnostics_failed`) surface the
sampler-diagnostic findings; they are consumed by `summary()` and
`diagnose()` rather than read directly. Together with
`sd_g_summary`, they make up the four attribute payloads stated at
the top of this section.

## Posterior predictive replicates

The release-blocking interval estimand is the posterior predictive
replicate $\theta_i^{\mathrm{rep}}$. At each MCMC iteration $s$, with
$g^{(s)}$ available and per-site weights $w_{ij}^{(s)}$ computed,
$\theta_i^{\mathrm{rep}}$ is drawn from the per-site posterior on the
grid:
$$
\theta_i^{\mathrm{rep}, (s)}
  \;\sim\;
  \mathrm{Categorical}_{\boldsymbol\theta}\!\big(\mathbf{w}^{(i, s)}\big),
  \qquad
  w_{ij}^{(s)} \propto \exp\!\big\{\log g_j^{(s)} +
    \log \mathcal{N}(\hat\theta_i;\, \theta_j, \sigma_i^2)\big\}.
$$
This is a draw from the distribution $p(\theta \mid \hat\theta_i,
\sigma_i, g^{(s)}) \propto g^{(s)}(\theta)\, p(\hat\theta_i \mid \theta,
\sigma_i)$ on the grid, for fixed site $i$ and at the current
iteration's $g^{(s)}$.

The distinction from the per-iteration posterior mean
$\theta_i^{\mathrm{mean}, (s)} = \sum_j w_{ij}^{(s)}\, \theta_j$ is
load-bearing. Quantiles of `theta_mean` across MCMC iterations give a
credible interval for the **estimator**; they omit the conditional
posterior dispersion of $\theta_i$ around its mean at each iteration.
Quantiles of `theta_rep` across MCMC iterations integrate over both
the across-iteration uncertainty in $(\alpha, \lambda, g)$ and the
within-iteration posterior on $\theta_i$, giving a calibrated
parameter-uncertainty interval. The package stores the full draws as
`theta_rep_draws` precisely so that the calibrated intervals can be
recomputed from the fit object without re-sampling. Vignette M6
reports the empirical coverage of these intervals on the locked
Lee-Sui benchmark.

## Reading map

This vignette stated the hierarchy, diagrammed it, derived the marginal
and the functionals of $g$ that the package returns, and mapped the
thirteen metadata fields back to the DAG. The remaining methodological
vignettes carry the construction into implementation and verification.

- **M4 — Grid construction.** The four grid recipes exposed by
  `make_efron_grid()` (`paper_realdata`, `paper_simulation`,
  `paper_sensitivity`, `kl_target_experimental`), the construction of
  the basis matrix $B$, the rank-postcondition guard against
  identifiability failure, and the trade-offs between grid width and
  resolution.
- **M5 — Stan model and cache.** The byte-locked Stan source
  `inst/stan/efron_re.stan`, the ten numerical-stability disciplines
  (`log_softmax`, `log_sum_exp`, max-shift normalisation, two-pass
  variance), and the two-tier cache that prevents redundant
  compilation.
- **M6 — Verification and calibration.** End-to-end coverage
  verification on the locked benchmark fixtures, the release-blocking
  acceptance band on the per-site replicate intervals
  ($[0.87, 0.92]$ at nominal $0.90$), and the diagnostic surface of
  the fitted object.

A reader who has followed M1–M3 now has the complete model
specification and a clear picture of the inferential payload
`fit$metadata` carries; M4–M6 close the loop from specification to
implementation and verified calibration.
