---
title: "M5 · Stan model, sampler, and cache"
description: "The locked Stan model, the seven-field data block, and the two-tier compile cache."
output:
  rmarkdown::html_vignette:
    fig_width: 7
    fig_height: 4.5
vignette: >
  %\VignetteIndexEntry{M5 · Stan model, sampler, and cache}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  echo     = TRUE,
  collapse = TRUE,
  comment  = "#>",
  message  = FALSE,
  warning  = FALSE,
  fig.width  = 7,
  fig.height = 4.5
)
library(bayesEfron)
```

```{r read-stan-source, include = FALSE}
stan_path <- system.file("stan", "efron_re.stan", package = "bayesEfron")
stan_source <- readLines(stan_path)

block_lines <- function(src, start_pattern) {
  start <- grep(start_pattern, src)[1]
  if (is.na(start)) {
    stop("Stan source block not found: ", start_pattern, call. = FALSE)
  }
  rel_close <- grep("^[}]", src[seq.int(start, length(src))])[1]
  if (is.na(rel_close)) {
    stop("Unterminated Stan source block: ", start_pattern, call. = FALSE)
  }
  src[seq.int(start, start + rel_close - 1L)]
}
```

## Scope of this vignette

M2 specified the log-spline prior on $g$, M3 composed that prior into
the four-level random-effects hierarchy, and M4 fixed the discrete
support $\{\theta_1, \ldots, \theta_L\}$ and the natural-cubic-spline
basis matrix $B \in \mathbb{R}^{L \times M}$ on which $g$ lives. M5
closes the implementation arc: it walks through the locked Stan
program `inst/stan/efron_re.stan` block by block, documents the
sampler defaults that `bayes_efron_fit()` passes to CmdStan, and
specifies the two-tier compile cache. The cache is not an
optimisation curiosity; the cache key, by including the SHA-256 of
the Stan source and the CmdStan and `cmdstanr` versions, is the
mechanism that prevents stale binaries from silently changing
inferential output across releases.

## Notation reminder

The notation is unchanged from M1–M4. $K$ is the number of sites;
$\hat\theta_i$ and $\sigma_i$ are the per-site point estimate and
within-study standard error; $\theta_i$ is the latent effect; $g$ is
the discrete mixing distribution on the support points $\theta_1 <
\cdots < \theta_L$; $B \in \mathbb{R}^{L \times M}$ is the
natural-cubic-spline basis matrix evaluated at the grid; $M$ is the
spline dimension; $\alpha \in \mathbb{R}^M$ is the spline coefficient
vector; and $\lambda > 0$ is the smoothness precision.

## Reading the Stan source

The package ships a single Stan program, `inst/stan/efron_re.stan`,
which is byte-locked at release: its SHA-256 enters the cache key, so
any modification invalidates every compiled binary in every user's
cache directory. The lock is verifiable. Each Stan block displayed
below is read at vignette-build time from the installed file via the
`stan_source` vector created in the setup chunk, then sliced with the
`block_lines()` helper that locates the block opener and the matching
top-level closing brace. The displayed code is therefore byte-identical
to the file the sampler reads. Readers who wish to confirm this can
run `digest::digest(file = system.file("stan", "efron_re.stan",
package = "bayesEfron"), algo = "sha256")` and compare against
`fit$metadata$stan_file_sha256` on any fitted object. The header
comment in the source carries authorship, version, and a one-paragraph
plain-English description of the model; it is preserved verbatim under
the lock but is not reproduced here, since the substantive content
appears below.

## Data block

The Stan data block declares the seven inputs the sampler consumes
per fit. These are precisely the seven fields stored on
`fit$metadata$data_list`, and the contract between R and Stan runs
through them: the package's input-validation layer guarantees that
every field below carries its declared type, bounds, and lengths
before CmdStan is invoked.

```{stan, output.var = "data_block", eval = FALSE, code = block_lines(stan_source, "^data [{]")}
```

The seven fields divide into three groups. The site-level data are
$K \in \mathbb{Z}_{\geq 1}$, the number of sites; the
length-$K$ vector `theta_hat` of observed effect estimates
$\hat{\boldsymbol\theta} = (\hat\theta_1, \ldots, \hat\theta_K)$; and
the length-$K$ vector `sigma` of strictly positive within-study
standard errors $\boldsymbol\sigma = (\sigma_1, \ldots, \sigma_K)$.
The discretisation of the prior is governed by $L \in \mathbb{Z}_{\geq 1}$,
the grid length (default 101), and the length-$L$ vector `grid`
$(\theta_1, \ldots, \theta_L)$, both specified by the recipes
documented in M4. The spline basis is encoded by $M \in \mathbb{Z}_{\geq 1}$
degrees of freedom (default 6) and the $L \times M$ matrix `B`,
where $B_{\ell m}$ is the value of the $m$-th natural-cubic-spline
basis function at $\theta_\ell$. The model contains no additional
hyperparameters that vary across fits: the half-Cauchy scale on
$\lambda$ and the ridge form of the conditional prior on $\alpha$
are written into the model block as constants. Everything that
distinguishes one fit from another lives in these seven fields. The
metadata payload `data_list` is therefore a sufficient statistic for
replay: rebuilding it from `fit$metadata` and feeding it back through
`cmdstanr::cmdstan_model()$sample()` reproduces the run up to the
sampler seed.

## Parameters block

The unknowns over which Hamiltonian Monte Carlo integrates are the
$M$-vector of spline coefficients and the scalar smoothness
precision.

```{stan, output.var = "parameters_block", eval = FALSE, code = block_lines(stan_source, "^parameters [{]")}
```

There are exactly $M + 1$ sampled parameters: $\alpha \in \mathbb{R}^M$
and $\lambda > 0$. At the default $M = 6$ the unconstrained
parameter space has dimension seven. The latent site effects
$\theta_i$ are absent from this block by design. M3 derived that
$\theta_i$ is integrated out analytically when forming the marginal
likelihood of $\hat\theta_i$: with $g$ discrete on the grid, the
convolution $\hat\theta_i \mid \sigma_i, g \sim \sum_\ell g_\ell\,
\mathcal{N}(\theta_\ell, \sigma_i^2)$ has a closed form that the
model block evaluates directly. Posterior summaries of $\theta_i$
that the analyst consumes — the posterior mean $\mathrm{E}[\theta_i
\mid \hat{\boldsymbol\theta}]$, the MAP, the posterior standard
deviation, and the posterior replicate $\theta_i^{\mathrm{rep}}$ —
are reconstructed in the generated-quantities block from the sampled
$(\alpha, \lambda)$, rather than carried as additional latent
variables. The practical consequence is that the dimension of the
sampler problem grows only with $M$, not with $K$: scaling to
hundreds of sites adds no parameters and no warmup cost beyond the
$O(K)$ per-iteration likelihood evaluation. The non-negativity bound
on $\lambda$ is enforced through the standard Stan log-Jacobian
transform; no manual reparameterisation is performed.

## Transformed parameters block

The transformed-parameters block carries the deterministic map
$\alpha \mapsto g$ that M2 defined, computed once per sampler
iteration and made available to both the model and the
generated-quantities blocks.

```{stan, output.var = "tparams_block", eval = FALSE, code = block_lines(stan_source, "^transformed parameters [{]")}
```

The three quantities computed here are the length-$L$ vector
`log_w` $= B\alpha$ of unnormalised log-densities at the grid points,
the length-$L$ vector `log_g` $= \log\!\big(\mathrm{softmax}(B\alpha)\big)$
of normalised log-densities, and the simplex-typed length-$L$ vector
`g` $= \mathrm{softmax}(B\alpha)$. The Stan function
`log_softmax(x)` returns $x - \log\!\sum_j \exp(x_j)$, computed via the
log-sum-exp identity that subtracts $\max_j x_j$ before exponentiation,
so the normalisation is numerically stable even when $B\alpha$ contains
large positive or large negative entries. The model block consumes
`log_g` (log space, for the per-site log-marginal likelihood), and
the generated-quantities block consumes both `log_g` (for the
per-site posterior at each grid point) and `g` (for the moments
$\mathrm{E}[\theta]$, $\mathrm{Var}[\theta]$, $\mathrm{SD}[\theta]$).
Computing the simplex once per iteration and reusing it across both
downstream blocks avoids redundant softmax calls and concentrates the
sole linear-algebra cost — the $L \times M$ matrix-vector product
$B\alpha$ — in a single line.

## Model block

The model block specifies the joint $p(\hat{\boldsymbol\theta},
\alpha, \lambda)$ up to a normalising constant. It contains two
prior statements and one likelihood loop.

```{stan, output.var = "model_block", eval = FALSE, code = block_lines(stan_source, "^model [{]")}
```

The hyperprior on the smoothness precision is half-Cauchy with
location zero and scale five,
$\lambda \sim \mathrm{Half\text{-}Cauchy}(0, 5)$, weakly informative
in the sense of Gelman (2006, *Bayesian Analysis*): heavy tails
permit the data to dominate when the spline coefficients support
large magnitudes, while the truncation to $\lambda > 0$ excludes the
ill-posed limit. Conditional on $\lambda$, the prior on the spline
coefficients is independent Gaussian with mean zero and standard
deviation $\lambda^{-1/2}$, written in the Stan source as
`alpha ~ normal(0, inv_sqrt(lambda))`. The implied marginal prior on
$\alpha$ after integrating out $\lambda$ has heavier tails than a
fixed-precision Gaussian; this is the mechanism by which the model
adapts the effective smoothness of $g$ to the data without an
analyst-chosen ridge constant.

The likelihood loop over $i = 1, \ldots, K$ implements the discrete
mixture marginal of M3:
$$
p(\hat\theta_i \mid \sigma_i, g)
  \;=\; \sum_{\ell=1}^{L} g_\ell \, \varphi(\hat\theta_i;\, \theta_\ell,\, \sigma_i),
$$
where $\varphi(\cdot;\, \mu, \sigma)$ is the Gaussian density. The
loop builds a length-$L$ vector of log-components
$\log g_\ell + \log \varphi(\hat\theta_i;\, \theta_\ell,\, \sigma_i)$
and reduces it with `log_sum_exp`, which evaluates
$\log\!\sum_\ell \exp(x_\ell)$ as $m + \log\!\sum_\ell \exp(x_\ell - m)$
with $m = \max_\ell x_\ell$. The shift guarantees that the largest
exponentiated term is exactly one, so underflow is impossible for the
dominant component and the catastrophic cancellation that a naive
$\sum \exp$ would suffer at small $\sigma_i$ is avoided. The
contribution to `target` is the log marginal $\log p(\hat\theta_i
\mid \sigma_i, g)$; the cumulative `target` after the loop is the
log of the joint up to the normalising constant of $g$ already
absorbed by `log_softmax`. No manual Jacobian adjustment is required:
the only constrained parameter is $\lambda > 0$, and Stan applies
the half-real Jacobian automatically.

## Generated quantities block

The generated-quantities block computes nine summaries per posterior
draw. The release-blocking Tier 0 verification target
`tier0_gq_count` asserts exactly that count.

```{stan, output.var = "gq_block", eval = FALSE, code = block_lines(stan_source, "^generated quantities [{]")}
```

The nine generated quantities partition into three groups. The first
group records the moments of the estimated mixing distribution $g$:
`mean_g` $= \sum_\ell g_\ell \theta_\ell$, the posterior mean of $g$
viewed as a distribution; `var_g` $= \sum_\ell g_\ell (\theta_\ell -
\mathrm{mean\_g})^2$, the two-pass variance; and `sd_g` $=
\sqrt{\mathrm{var\_g}}$. Per-draw summaries of these three quantities
populate the metadata fields `mean_g_summary`, `var_g_summary`, and
the `sd_g_summary` attribute carried on `fit$metadata`. The second
group computes per-site posterior summaries. For each site $i$ the
log posterior on the grid $\log p(\theta_\ell \mid \hat\theta_i, g)$
is formed from `log_g` plus the per-grid-point Gaussian log-likelihood,
normalised via the max-shift `log_sum_exp` discipline. `theta_map`
selects the grid index that maximises this log posterior; `theta_mean`
is the posterior mean $\sum_\ell w_\ell \theta_\ell$ with $w_\ell$
the normalised posterior weights; `theta_sd` is the two-pass posterior
standard deviation; and `theta_rep` is a draw from the discrete
posterior via `categorical_rng(w)`, the per-draw replicate used for
posterior predictive intervals (the `theta_summary` and
`theta_rep_draws` metadata fields). The third group computes model
diagnostics: `effective_params` $= K - \sum_i \mathrm{Var}(\theta_i
\mid \hat\theta_i, g)/\sigma_i^2$, an effective-parameter count
analogous to a hierarchical $\mathrm{DIC}$ penalty, and
`log_marginal_likelihood` $= \sum_i \log p(\hat\theta_i \mid \sigma_i, g)$,
the log of the data marginal under the discrete mixture. These feed
the `effective_params_summary` and `log_marginal_likelihood_summary`
metadata fields.

## Sampler defaults

`bayes_efron_fit()` invokes `cmdstanr::CmdStanModel$sample()` with
defaults chosen for reproducibility rather than tightest performance.
The chain count defaults to `chains = 4L`, warmup to
`iter_warmup = 1000L`, post-warmup sampling to
`iter_sampling = 3000L`, and the NUTS target acceptance to
`adapt_delta = 0.9`. The fixed warmup is long enough to let the
sampler adapt the dual-averaging step size and the mass matrix
without budget pressure; the post-warmup count delivers
$4 \times 3000 = 12000$ retained draws at the default chain count,
which keeps per-site posterior tail quantiles stable for the
verification regimes covered in M6. The CmdStan initialisation is
fixed at `init = 0.5` and is intentionally not exposed as a
user-facing tuning argument in v0.1; the convention narrows the space
of release-blocking surprises that user inits can introduce. The
parallel-chain count defaults to `chains` but can be overridden by
the environment variable `BAYESEFRON_PARALLEL_CHAINS` (or the
unprefixed `PARALLEL_CHAINS`); the override is validated against
`chains` and rejected if it exceeds it, so a fit that requests four
chains never silently runs fewer. Seeds are auto-generated when
`seed = NULL` and recorded on `fit$metadata$seed`, supporting later
replay.

## The two-tier cache

Stan compilation is the dominant fixed cost of a fit, and the binary
produced by `cmdstanr` is identical across runs that share the same
source, compiler, options, and CmdStan/`cmdstanr` versions. The
package exploits both facts through two cache layers.

The in-session cache is an `environment` object attached to the
package namespace, keyed by the SHA-256 cache key documented below.
Once a model has been compiled (or restored from disk) within an
R session, subsequent fits in the same session look up the
`cmdstanr::CmdStanModel` handle in $O(1)$ time; no file I/O, no
compiler invocation, and no inter-process locking is performed. The
session cache is cleared automatically when the R session ends and
can be cleared explicitly via `bayes_efron_clear_cache("session")`.

The on-disk cache lives under a per-user cache root resolved from
the `BAYESEFRON_CACHE_ROOT` environment variable, with
`tools::R_user_dir("bayesEfron", which = "cache")` as the fallback.
The root holds a single format-version subdirectory (currently
`v1`) and within it the compiled binary, a copy of the Stan source,
and a JSON sidecar carrying the cache provenance:

```text
<BAYESEFRON_CACHE_ROOT>/
  v1/
    .lock                     # cross-process lock file
    <key>                     # compiled CmdStan executable
    <key>.stan                # copy of the Stan source used at compile time
    <key>.meta.json           # JSON sidecar with provenance fields
```

The 64-character hexadecimal `<key>` is the SHA-256 of the
canonicalised JSON serialisation of the cache-key payload. The
payload has eleven fields: the SHA-256 of the Stan source file, the
CmdStan version, the `cmdstanr` version, the R architecture string,
the `CXX` compiler identifier, the cache format version, the model
family (`"RE"` in v0.1), SHA-256 hashes of the normalised `cpp_options`
and `stanc_options` lists, a SHA-256 of the normalised
`~/.R/Makevars` snapshot, and the platform-and-OS-major identifier
formed by concatenating `R.version$platform` with the platform's
major-release token. The eleven-field composition is designed so that
any change capable of producing a numerically distinct binary changes
the key. The on-disk lookup is guarded by a cross-process advisory
lock at `.lock`: the lock is acquired before any directory write,
held until the write completes, and released by `on.exit` even when
the writer is interrupted. Stale locks left behind by killed
processes are detected by reading the PID, hostname, and process
start time the lock file records, and are cleared once the recorded
process is no longer alive.

## Cache lifecycle and maintenance

Both cache layers populate lazily. The first `bayes_efron_fit()` (or
explicit `bayes_efron_compile()`) call in a fresh environment
computes the cache key, finds no on-disk entry, acquires the lock,
compiles via `cmdstanr::cmdstan_model()`, writes the binary and
sidecar atomically through a staging directory, releases the lock,
and finally installs the handle in the session cache. The second
call in the same R session returns the session-cached handle without
touching the disk. A first call in a new R session that finds a
matching on-disk entry validates the sidecar against the recomputed
key, restores the binary, and installs the handle in the session
cache; compilation is skipped. Cache eviction is manual:
`bayes_efron_clear_cache()` accepts four scopes. `"lock_only"`
removes only the lock file (used to recover from a crashed
compilation that left a stale lock). `"session"` clears the
in-session environment without touching disk. `"compiled_models"`
removes the executables, sidecars, and source copies under the
format-version subdirectory but preserves the cache root itself.
`"all"` clears the in-session environment and removes the entire
cache root after verifying that the path is a directory the current
process owns. The four scopes correspond to four common operational
needs and are documented in detail in the function's reference page.

## Reading map

This vignette specified the Stan implementation block by block,
documented the sampler defaults, and laid out the two-tier compile
cache that supports reproducible inference at release. The final
methodological vignette closes the loop from specification to
verified calibration.

- **M6 — Verification and calibration.** The four-tier verification
  ladder (Tier 0 structural invariants such as the nine-field
  generated-quantities count documented above; Tier 1 internal
  consistency against an `optim`-based reference for `theta_map`;
  Tier 2 deferred cross-implementation comparisons against
  `deconvolveR`; Tier 3 the release-blocking aggregate coverage
  result on the Lee–Sui benchmark fixtures, with acceptance band
  $[0.87, 0.92]$ at nominal $0.90$). The cache-key composition
  documented here is what licenses Tier 3 reproducibility: a binary
  that produced the gating coverage result can be re-identified, on a
  fresh machine, from the eleven payload fields alone.

A reader who has followed M1–M5 holds the full specification from
hierarchy to compiled binary. M6 supplies the verification surface
that v0.1.0 release rests on.
