---
title: "M6 · Verification and calibration"
description: "The verification tier ladder and the Lee-Sui aggregate coverage evidence that gates v0.1.0."
output:
  rmarkdown::html_vignette:
    fig_width: 7
    fig_height: 5
vignette: >
  %\VignetteIndexEntry{M6 · Verification and calibration}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  echo     = TRUE,
  collapse = TRUE,
  comment  = "#>",
  message  = FALSE,
  warning  = FALSE,
  fig.width  = 7,
  fig.height = 5
)
library(bayesEfron)
has_ggplot2 <- requireNamespace("ggplot2", quietly = TRUE)
```

```{r tier3-evidence, include = FALSE}
## Canonical Tier 3 acceptance evidence for v0.1.0, transcribed from
## `tests/testthat/verification_targets.csv` (band) and from the
## accepted Step 7.7 full-live matrix (per-K aggregate coverage). The
## ledger file lives under `tests/testthat/` and is therefore not on
## the install path that this vignette sees at build time, so the
## numbers are reproduced here as named constants.
tier3 <- data.frame(
  K        = c(50, 100, 200, 500, 1500),
  coverage = c(0.8940, 0.8910, 0.8940, 0.8916, 0.876433333333333),
  n_sites  = c(1000, 2000, 4000, 10000, 30000),
  stringsAsFactors = FALSE
)
band_lower <- 0.87
band_upper <- 0.92
nominal    <- 0.90
```

## Scope of this vignette

M1 through M5 specified the empirical-Bayes deconvolution problem,
the log-spline parameterisation of the mixing distribution
$g(\theta)$, the random-effects hierarchy, the discrete support and
natural-cubic-spline basis, and the locked Stan program and compile
cache. M6 closes the methodological track. It documents the four-tier
verification ladder that the package's release decision rests on and
the all-twenty-replication Lee-Sui aggregate coverage evidence that
gates v0.1.0. The applied counterpart A7 walks through one $K = 50$
replication end to end; M6 covers all five $K$ values and the full
twenty replications per $K$ that make the release ledger.

## Notation reminder

Notation is unchanged from M1-M5. $K$ is the number of sites;
$\hat\theta_i$ and $\sigma_i$ are the per-site point estimate and
within-study standard error; $\theta_i$ is the latent site effect;
$g$ is the discrete mixing distribution on the grid
$\{\theta_1, \ldots, \theta_L\}$; $B \in \mathbb{R}^{L \times M}$ is
the natural-cubic-spline basis matrix; $\alpha \in \mathbb{R}^M$ is
the spline coefficient vector and $\lambda > 0$ is the smoothness
precision. The per-site posterior replicate
$\theta_i^{\mathrm{rep}}$, drawn from the discrete posterior on the
grid via `categorical_rng`, is the calibrated estimand on which the
release gate is set.

## Why verification matters

The package fits a fully Bayesian log-spline random-effects model on
heteroscedastic meta-analytic data, exposes per-site posterior
intervals on $\theta_i$ as its headline summary, and asks the analyst
to act on those intervals. A user reading a nominal-90% interval
expects the long-run rate at which the interval traps the latent
effect to be close to 90% under known data-generating conditions. The
package must produce defensible evidence that this is so, on data of
the family the model is designed for, before release.

Verification cannot be reduced to a single end-to-end coverage test.
A calibrated-interval coverage figure can fail through any of several
disjoint mechanisms: a transcription error in the spline basis, a
cache subsystem that returns a stale binary, a sampler-init
regression that lengthens warmup, a grid recipe that no longer
matches its committed reference, or a model-block edit that shifts
the marginal likelihood. Each mechanism is best detected at its own
granularity. The package therefore organises verification into four
tiers, each targeting a different class of failure mode at a
different numerical tolerance.

The pyramid is determined by tolerance band, not by wall-clock
runtime. Tier 0 carries the most tests at the tightest tolerance;
Tier 3 carries the fewest at the loosest. A slow integration test
that asserts a machine-precision invariant remains Tier 0, and a
fast unit test that asserts a 5% bound remains Tier 2. The
construction produces a clean localisation property: a Tier 0
failure points at a single line of code, since an algebraic
invariant cannot fail by 5%, only by being wrong; a Tier 3 failure
can implicate model, grid, cache, or inferential framework and
requires triage.

## Tier 0: deterministic invariants

Tier 0 collects the algebraic and structural invariants of the
package, asserted at machine precision (a relative or absolute
tolerance of $10^{-15}$) or at byte identity. Seventeen Tier 0
targets are recorded in `tests/testthat/verification_targets.csv`;
sixteen are release-blocking and active at v0.1.0, and one is
non-blocking. The targets divide into three groups.

The first group asserts algebraic identities on the prior and basis.
The simplex normalisation $\sum_\ell g_\ell = 1$ is checked at
$10^{-15}$ after every `softmax(B\alpha)` evaluation. The basis
matrix $B \in \mathbb{R}^{L \times M}$ is checked for finite entries
and, jointly with the constant vector, for full rank
$\operatorname{rank}([\mathbf{1}\;B]) = M + 1$ in the
intercept-augmented form that the identifiability argument of M2 and
M4 requires. The augmented form is necessary because a model with $B$
alone can be made unidentifiable when $\mathbf{1} \in \operatorname{col}(B)$;
the rank check rules that case out at the grid-construction
boundary.

The second group asserts byte-level identity for the locked artefacts
of the package. Three SHA-256 digests are stored in
`inst/locked-core-checksums.txt`, one each for `inst/stan/efron_re.stan`,
`R/grid.R`, and `R/data-prep.R`. The audit test recomputes the digest
of each file via `digest::digest(path, algo = "sha256", file = TRUE)`
and asserts byte equality against the ledger. Three byte-identity
fixtures store the output of the three replication-faithful grid
methods (`paper_realdata`, `paper_simulation`, `paper_sensitivity`)
as RDS payloads; a silent change to the grid formula would surface
immediately as a fixture mismatch.

The third group asserts structural counts and cache determinism. The
`bef_fit_re` metadata schema is locked at thirteen fields. The Stan
generated-quantities block produces exactly nine quantities. The
eleven-field cache-key payload of M5 produces a deterministic
SHA-256 for any fixed payload. The post-fit invariants $\theta_i^{\mathrm{sd}} \geq 0$
and $\theta_i^{\mathrm{sd}} < \infty$ are checked on the
generated-quantity output of every fit, closing the two-pass-variance
cancellation paths flagged in M5.

## Tier 1: closed-form and quadrature parity

Tier 1 collects cross-checks of R-side computations against
closed-form references and reference implementations from the base R
distribution, asserted at a tolerance of $10^{-6}$ (or $10^{-12}$ for
identities that are effectively algebraic). Seven Tier 1 targets are
release-blocking and all are active.

Three Tier 1 targets exercise the spline basis and the discrete prior
construction. The natural-cubic-spline basis produced by
`make_efron_grid()` is required to agree with the reference produced
by `splines::ns()` at $10^{-6}$ on a fixed test grid. The discrete
prior $g$ obtained from the `paper_simulation` method is required to
agree with the reference produced by `prop.table()` on a deterministic
input vector at $10^{-6}$. These two targets bind the grid and basis
construction documented in M4 to the canonical base-R implementations
that the rest of the R ecosystem cross-references.

Three further Tier 1 targets exercise the closed-form posterior
algebra of M3. At $K = 2$ the Normal-Normal posterior moments admit a
closed form; the package's per-site posterior reconstruction is
checked against that closed form at $10^{-6}$. The prior moments
$\operatorname{E}[g]$ and $\operatorname{Var}[g]$ are checked against
Gauss-Hermite quadrature on the corresponding integrals at the same
tolerance. The per-site MAP, $\theta_i^{\mathrm{map}}$, is checked
against the maximiser returned by `optim()` on the per-site
closed-form objective.

A final Tier 1 target checks that `posterior::quantile2()` returns
the same per-site interval endpoints as `base::quantile()` on a
fixed `draws_array`, at a tighter algebraic tolerance of $10^{-12}$.
The posterior summarisation pipeline of M5 passes through
`quantile2`; the cross-check rules out a silent change in the
draws-summary layer.

## Tier 2: deconvolution parity

Tier 2 collects cross-checks against alternative deconvolution
implementations, asserted at the 5% to 10% bands that reflect
genuine prior-class differences. Six Tier 2 targets are recorded;
none are release-blocking and all are deferred at v0.1.0.

The Tier 2 reference is the penalised maximum-likelihood non-parametric
deconvolution implemented in `deconvolveR` (Narasimhan and Efron,
2020, *Journal of Statistical Software*). The comparison is
restricted to homoscedastic z-score panels ($\sigma_i \equiv 1$) at
$K \in \{20, 50, 100\}$, because `deconvolveR` is designed for the
homoscedastic case and the heteroscedastic regime the package is
designed for would conflate two distinct sources of disagreement.
Within the homoscedastic regime the band tolerances are 5% for
per-site posterior means and 10% for the marginal moments of $g$;
the loose bounds reflect the full-Bayes-versus-penalised-MLE
prior-class difference rather than a numerical disagreement.

Tier 2 is deferred to v0.2 because the deferred targets are
informative rather than blocking: a Tier 2 disagreement is expected
behaviour at the band tolerances and does not, by itself, indicate a
package fault. The Tier 2 row schema is committed at v0.1.0 so that a
v0.2 reactivation requires no schema migration.

## Tier 3: live HMC calibration on the Lee-Sui benchmark

Tier 3 collects the release-blocking calibration targets: aggregate
empirical coverage of the nominal-90% interval on the per-site
posterior replicate $\theta_i^{\mathrm{rep}}$, computed over twenty
independent replications of the Lee-Sui benchmark at each of five
site counts $K \in \{50, 100, 200, 500, 1500\}$. The acceptance band
is $[0.87, 0.92]$ at the nominal level $0.90$. Five Tier 3 rows are
release-blocking and all five are active. Two additional Tier 3 rows
are non-blocking diagnostics deferred to v0.2.

The benchmark generator is the heteroscedastic twin-towers mixture
documented in Lee and Sui (2025, *Mathematics*) and used in the
Lee-Sui small-$K$ shrinkage analyses. Five locked fixtures store the input
arrays for each $K$:
`tests/testthat/_fixtures/lee_sui_K{50,100,200,500,1500}.rds`. Each
fixture is a `v2`-schema bundle that carries twenty replications,
each a length-$K$ triple `(theta_hat, sigma, theta_true)` together
with provenance metadata. The site counts span the small-$K$ regime
in which discrete-grid deconvolution is most sensitive to grid and
prior choices (K = 50, 100), the mid-range regime in which most
applied multisite analyses operate (K = 200, 500), and the
densely-sampled regime in which recovery of $g(\theta)$ becomes a
leading constraint on coverage (K = 1500).

The harness `tests/testthat/test-tier3-live-harness.R` loops over the
five $K$ values and the twenty replications per $K$. For each
replication it fits `bayes_efron_fit(theta_hat, sigma)` with the
v0.1 default configuration (four chains, 1000 warmup iterations,
3000 sampling iterations per chain, `init = 0.5`, `adapt_delta = 0.9`,
the project-canonical Stan sampler seed), extracts the per-site
posterior replicate draws $\theta_i^{\mathrm{rep}}$, computes per-site
90% intervals via `posterior::quantile2(probs = c(0.05, 0.95))`, and
records the per-site coverage rate as the fraction of sites for which
`lower <= theta_true <= upper`. The replication-level coverage is the
mean of per-site indicators on the replication's $K$ sites; the
aggregate $K$-level coverage is the mean of the twenty
replication-level rates. Because $K$ is fixed within a row, the
aggregate is equivalent to a pooled fraction over $20K$ site-level
indicators. The release-blocking assertion is that all five aggregate
rates lie in $[0.87, 0.92]$.

The band's two-sided structure is intentional. The lower bound
budgets a small but explicit deficit for under-coverage: intervals
that are mildly narrow under the generator remain in band provided
the deficit is bounded. The upper bound denies credit for
over-conservatism: a release that buys coverage by inflating intervals
well beyond the nominal level is also out of band. The pre-specified
band is recorded on each of the five Tier 3 rows as
`expected_value_lower = 0.87` and `expected_value_upper = 0.92`.

The five accepted aggregate coverage values that constitute the
v0.1.0 release evidence are reproduced in the constants block at the
top of this vignette and rendered in the table and figure below.
The Tier 3 evidence is intentionally benchmark-scoped: it underwrites
calibration of $\theta_i^{\mathrm{rep}}$ intervals on the locked
Lee-Sui all-twenty-replication fixtures under the v0.1 default fit
configuration. It does not underwrite a generic frequentist coverage
claim for arbitrary data-generating mechanisms, nor for new grid
choices, nor for alternate sampler settings; users whose data depart
materially from the twin-towers generator should treat the band as
suggestive and run a scenario-specific calibration test of their
own.

```{r tier3-table, echo = FALSE}
display <- data.frame(
  K                  = tier3$K,
  `Aggregate coverage` = sprintf("%.4f", tier3$coverage),
  `Site denominator` = tier3$n_sites,
  `In band`          = ifelse(
    tier3$coverage >= band_lower & tier3$coverage <= band_upper,
    "yes", "no"
  ),
  check.names = FALSE
)
knitr::kable(display, align = c("r", "r", "r", "c"))
```

## The K = 1500 acceptance

The $K = 1500$ row is the worst case of the five and the only Tier 3
row that was repaired before acceptance. Its accepted aggregate
coverage is $0.876433333333333$ on a denominator of
$20 \times 1500 = 30000$
site-level coverage indicators, against the same $[0.87, 0.92]$ band.
The result cleared the lower bound by approximately four-tenths of a
percentage point, the slimmest margin of the five. The pre-repair
$K = 1500$ result fell below the lower bound; the repaired result is
the one recorded in the ledger and reproduced above. The repair
exercise validated the assumption underlying the band's lower bound,
that the model can deliver in-band aggregate coverage in the
densely-sampled regime, while leaving the headroom small enough to
flag $K = 1500$ as the constraint case for any future change that
might affect calibration.

The accepted $K = 1500$ replication fits retained sampler diagnostic
flags on $\widehat{R}$, bulk effective sample size, and tail effective
sample size. The ledger treats these flags as calibration-tier
diagnostic metadata and does not treat them as Tier 3 acceptance
gates: a diagnostic flag at the conventional thresholds for
$\widehat{R}$ or effective sample size does not invalidate the
aggregate coverage acceptance, but the flag is preserved on the
accepted fits so that the diagnostic stream and the coverage stream
remain separable. The v0.1.0 release ledger therefore records the
$K = 1500$ acceptance with the diagnostic flags attached.

## Reproducibility

Tier 3 live verification is environment-variable gated. Default
`R CMD check` skips the live harness entirely. Setting
`BAYESEFRON_RUN_LIVE = 1` opts in to a smoke path that fits one
replication at $K = 50$ and checks that the harness executes without
error. Setting `BAYESEFRON_RUN_FULL_LIVE = 1` opts in to the
release-nightly path that fits all twenty replications at all five
$K$ values; on a reference runner the full matrix is the dominant
wall-clock cost of CI. Fresh full-live Tier 3 refits without a
replay matrix require the additional explicit opt-in
`BAYESEFRON_TIER3_OK_TO_REFIT = 1`, so that a routine full-live
invocation does not silently launch a fresh Stan recalibration.

The accepted evidence matrix can be replayed instead of refit. The
environment variable `BAYESEFRON_TIER3_FULL_LIVE_MATRIX` points to a
runner-local copy of the accepted Step 7.7 evidence matrix; when set,
the harness validates the matrix against the band and returns
without sampling. Replay-mode verification preserves the audit chain
on the accepted aggregate values while keeping the CI envelope
within tag-release budget. The replay path and the fresh-fit path
return the same assertion structure; the difference is the source of
the aggregate coverage numbers, not the assertion that those numbers
lie in $[0.87, 0.92]$.

The `.github/workflows/check.yml` configuration prepares a CI-local
replay copy of the accepted matrix for version-tag, scheduled, and
manual full-live runs. The configuration sets
`BAYESEFRON_RUN_FULL_LIVE = 1` and the matrix env var on those
trigger branches, so that release-tag verification confirms the
accepted matrix without an accidental fresh recalibration. The
pull-request smoke path is independent and runs on the smaller
envelope.

## A coverage-vs-band figure

The figure below shows the five accepted aggregate coverages against
the $[0.87, 0.92]$ band and the nominal $0.90$ reference line. The
horizontal axis is the site count $K$ on a logarithmic scale; the
vertical axis is the aggregate coverage. The shaded region is the
pre-specified acceptance band; the dashed line is the nominal level;
each point is the accepted aggregate coverage from the
twenty-replication full-live matrix. The figure has no inferential
content beyond the table above; it makes the position of each $K$
within the band, and the proximity of the $K = 1500$ point to the
lower bound, visible at a glance.

```{r coverage-figure-ggplot, eval = has_ggplot2, fig.cap = "Accepted aggregate coverage at nominal-90% per K, with the [0.87, 0.92] acceptance band shaded."}
library(ggplot2)
ggplot(tier3, aes(x = K, y = coverage)) +
  annotate(
    "rect",
    xmin = min(tier3$K) * 0.7, xmax = max(tier3$K) * 1.3,
    ymin = band_lower, ymax = band_upper,
    alpha = 0.15, fill = "steelblue"
  ) +
  geom_hline(yintercept = nominal, linetype = "dashed", colour = "grey40") +
  geom_point(size = 3) +
  geom_text(
    aes(label = sprintf("%.4f", coverage)),
    nudge_y = 0.0035, size = 3.2
  ) +
  scale_x_log10(breaks = tier3$K) +
  scale_y_continuous(
    breaks = c(band_lower, 0.88, 0.89, nominal, 0.91, band_upper),
    limits = c(0.86, 0.93)
  ) +
  labs(
    x = "K (sites, log scale)",
    y = expression("Aggregate coverage of nominal-90% " * theta^rep * " interval")
  ) +
  theme_minimal(base_size = 11)
```

```{r coverage-figure-base, eval = !has_ggplot2, fig.cap = "Accepted aggregate coverage at nominal-90% per K, with the [0.87, 0.92] acceptance band shaded (base graphics fallback)."}
op <- par(mar = c(4.5, 4.5, 1, 1))
plot(
  tier3$K, tier3$coverage,
  log = "x", pch = 19, cex = 1.4,
  xlab = "K (sites, log scale)",
  ylab = expression("Aggregate coverage of nominal-90% " * theta^rep * " interval"),
  ylim = c(0.86, 0.93),
  xlim = c(min(tier3$K) * 0.7, max(tier3$K) * 1.3),
  axes = FALSE
)
rect(
  xleft = min(tier3$K) * 0.7, xright = max(tier3$K) * 1.3,
  ybottom = band_lower, ytop = band_upper,
  col = grDevices::adjustcolor("steelblue", alpha.f = 0.15), border = NA
)
abline(h = nominal, lty = 2, col = "grey40")
points(tier3$K, tier3$coverage, pch = 19, cex = 1.4)
text(tier3$K, tier3$coverage + 0.0035, sprintf("%.4f", tier3$coverage), cex = 0.8)
axis(1, at = tier3$K, labels = tier3$K)
axis(2, at = c(band_lower, 0.88, 0.89, nominal, 0.91, band_upper), las = 1)
box()
par(op)
```

The figure renders identically through the ggplot2 backend (when
available) and a base-graphics fallback. The chunk-eval gating mirrors
the package's own plot dispatch documented in A6.

## Reading map

This vignette closes the methodological track. The applied
counterpart A7 walks through one $K = 50$ replication of the Lee-Sui
benchmark end to end and traces the same per-site interval that the
Tier 3 harness aggregates here; reading A7 alongside this section
gives the within-replication picture that complements the
across-$K$ release ledger.

- **A7 - Case study, Lee-Sui replication light.** One $K = 50$
  replication, with the per-site posterior summary, the caterpillar
  view, and the coverage interpretation on a single replication's
  fit.
- **NEWS.md.** The canonical release-note record of the accepted
  Tier 3 evidence, including the verbatim $K = 1500$ aggregate
  coverage value and the env-var gating contract.
- **`tests/testthat/verification_targets.csv`.** The ledger of all
  Tier 0, Tier 1, Tier 2, and Tier 3 targets, including the five
  release-blocking Tier 3 rows with the $[0.87, 0.92]$ band recorded
  on each row.

A reader who has followed M1 through M6 holds the full chain from
the empirical-Bayes formulation, through the log-spline
parameterisation, the random-effects hierarchy, the discrete grid and
basis, the locked Stan program and cache, and the four-tier
verification ladder that licences the v0.1.0 release.
