---
title: "The model"
description: "The deconvolution problem, the log-spline family for the distribution of effects, its prior, and the posterior distribution of a site effect."
output:
  rmarkdown::html_vignette:
    fig_width: 7
    fig_height: 4.5
bibliography: references.bib
link-citations: true
vignette: >
  %\VignetteIndexEntry{The model}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
library(bayesEfron)
options(bayesEfron.use_cli = FALSE, width = 80)
num <- function(x, digits = 2) formatC(x, format = "f", digits = digits)
fit <- raudenbush_fit
```

A meta-analysis or a multisite trial yields, for each of $K$ sites, an
estimate $\hat\theta_i$ of an effect and a standard error $\sigma_i$.
Inference about the true effects $\theta_i$ can be split into two steps:

(a) inference from the estimates to the distribution of true effects over
    sites, and
(b) inference from that distribution, together with a site's own estimate,
    to the true effect of the site.

Step (b) is an application of Bayes' rule once the distribution is known.
Step (a) is the difficult one. The model described here is that of
@lee2025, who give the derivations and a simulation study.

## The deconvolution problem

For sites $i = 1, \ldots, K$ the model is

$$
\hat\theta_i \mid \theta_i \sim N(\theta_i, \sigma_i^2), \qquad
\theta_i \sim g ,
$$

with the $\theta_i$ independent. The standard errors $\sigma_i$ come from
the design of each study; they are taken as known and usually differ from
site to site. This is the model of a random-effects meta-analysis
[@raudenbush1985], except that the distribution $g$ of true effects is left
unspecified.

Each estimate is the sum of a draw from $g$ and an error, so its density is
not $g$ but $g$ blurred by a normal density,

$$
f_i(x) = \int \phi(x;\, \theta, \sigma_i)\, g(\theta)\, d\theta ,
$$

where $\phi(x; \theta, \sigma)$ is the normal density with mean $\theta$
and standard deviation $\sigma$. Recovering $g$ from data whose
distributions are the $f_i$ is called deconvolution [@efron2016].

If $g$ were known, step (b) would be immediate. By Bayes' rule the
posterior density of a site's effect is

$$
p(\theta_i \mid \hat\theta_i) \propto
  g(\theta_i)\, \phi(\hat\theta_i;\, \theta_i, \sigma_i) ,
$$

the product of $g$ and the likelihood of the site's own estimate. When
$\sigma_i$ is small the likelihood is narrow and the posterior stays near
the estimate; when it is large the posterior is drawn toward the regions
where $g$ puts its probability. How far, and in which direction, depends on
the shape of $g$. That is the idea of empirical Bayes [@robbins1956]: the
other sites tell us about $g$, and $g$ tells us how to read each site. In
this sense $g$ is the prior distribution of each $\theta_i$, and the
printed output of the package labels it so; the distributions of $\alpha$
and $\lambda$ below are priors on the parameters of $g$.

It is also why $g$ is of interest in itself. From it follow the proportion
of sites whose effects are negligible or large, the effect to be expected
in a new site, and the amount of variation between sites.

The difficulty lies in step (a). Blurring loses information: distributions
$g$ that differ a great deal in detail can give nearly the same densities
$f_i$, so the data constrain the detail of $g$ only weakly, and less so the
larger the standard errors. Some assumption about $g$ is needed to obtain a
stable estimate from a few dozen or a few hundred sites.

## What to assume about g

The oldest assumption is that $g$ is normal. Its mean and variance are then
the two parameters of the random-effects model, and the posterior mean of a
site effect is a weighted average of the site's estimate and the overall
mean. The assumption is convenient and often adequate. It fails, however,
when the effects fall into groups, or when a few sites have effects much
larger than the rest, because a normal distribution has one peak, is
symmetric and has light tails.

At the other extreme no assumption is made, and $g$ is estimated by
maximum likelihood over all distributions [@kiefer1956; @laird1978]. The
maximizing distribution is discrete, with no more points of support than
there are sites [@lindsay1995], and it can be computed by convex
optimization [@koenker2014]. A discrete estimate is awkward when the
effects are thought to vary continuously, and quantities such as the
probability of an effect above a given size depend on where the few
points of support happen to fall.

Between the two are parametric families richer than the normal, such as
the mixtures of a point mass at zero with a normal or a Laplace
distribution in the ebnm package [@willwerscheid2025]. These suit problems
in which most effects are thought to be zero, although, like the normal,
they commit to a shape in advance.

@efron2016 proposed a family that is flexible in shape but has few
parameters, and in which $g$ is smooth. The model of this package uses that
family. @lee2025 [Section 2] discuss the alternatives at greater length.

## The log-spline family

Let $\tau_1 < \cdots < \tau_L$ be a grid of equally spaced points that
covers the values the true effects can plausibly take, and let $g_j$ be the
probability that $g$ gives to $\tau_j$. The family is

$$
g_j(\alpha) = \frac{\exp(B_j^\top \alpha)}
                   {\sum_{l=1}^{L} \exp(B_l^\top \alpha)},
\qquad j = 1, \ldots, L,
$$

where $B_j$ is the vector of $M$ natural cubic spline functions evaluated
at $\tau_j$ [see @hastie2009 for the splines] and $\alpha$ is a vector of $M$
coefficients. The logarithm of $g$ is thus a smooth function of the effect,
described by $M$ numbers, up to the constant that makes the probabilities
sum to one. With $\alpha = 0$ every grid point has the same probability.
Other values of $\alpha$ give distributions with one or several peaks, with
skewness, or with long tails; the larger $M$, the more shapes are within
reach.

A grid distribution stands in for a continuous one. With a fine grid the
sum $\sum_j g_j\, \phi(x; \tau_j, \sigma_i)$ is close to the integral that
defines $f_i$, so little is lost as long as the spacing of the grid is
small compared with the standard errors. The defaults are $L = 101$ points
and $M = 6$ spline functions, as in the application of @lee2025.
`vignette("choosing-a-grid")` discusses both choices and the placing of the
grid.

The spline basis has no constant term, because adding a constant to every
$B_j^\top \alpha$ leaves $g$ unchanged, so that the coefficient of a
constant column could not be determined by the data. The basis is computed
by `splines::ns(grid, df = M, intercept = FALSE)`, and `make_efron_grid()`
verifies that the columns of $B$ together with a constant column are
linearly independent, stopping with an error if they are not.

The columns of $B$ are used as `ns()` returns them, as in the code that
accompanies @lee2025. The deconvolveR package [@narasimhan2020] by default
centers each column and scales it to a sum of squares of one, so the
coefficients of the two packages are on different scales, which matters
for the prior distribution described next.

## The prior on the coefficients

Six coefficients estimated from a few dozen sites need restraint, or the
estimated $g$ follows the accidents of the sample. @efron2016 restrains
them with a penalty of fixed size in the likelihood. Here the restraint is
a prior distribution whose strength is itself estimated:

$$
\alpha_m \mid \lambda \sim N(0,\, 1/\lambda), \quad m = 1, \ldots, M,
\qquad \lambda \sim \mbox{half-Cauchy}(0, 5) .
$$

The second argument of the normal distribution is a variance, so the
coefficients have standard deviation $\lambda^{-1/2}$. A large $\lambda$
holds the coefficients near zero, and $g$ near the uniform distribution on
the grid. A small $\lambda$ lets them move, and $g$ take whatever shape the
likelihood favors. Because $\lambda$ has a prior of its own, its value is
estimated along with the coefficients. Six coefficients carry little
information about it, however, and its posterior distribution is wide, as
the output below shows.

```{r prior-lambda, include = FALSE}
# lambda = 1 / sd^2, so P(sd > s) = P(lambda < 1 / s^2) under half-Cauchy(0, 5)
prior_sd_above <- function(s) 2 / pi * atan(1 / (5 * s^2))
```

The half-Cauchy distribution is the choice of @lee2025. @gelman2006
proposed it as a prior for the standard deviation of the effects in a
hierarchical model with few groups. Here it is placed on the precision,
which implies a
different prior for the standard deviation $\lambda^{-1/2}$ of the
coefficients: one with median `r num(1 / sqrt(5))` that gives probability
`r num(prior_sd_above(1))` to values above 1 and
`r num(1 - prior_sd_above(0.1))` to values below 0.1. @lee2025
[Appendix D] replaced it by other families and other scales in a
simulation, fitted with a variational approximation, and report that the
estimates of the site effects and the coverage of their intervals changed
little.

For the teacher expectancy studies the posterior distribution of $\lambda$
is concentrated on small values, that is, on coefficients that are free to
move.

```{r posterior-lambda}
draws <- posterior::as_draws_matrix(fit)
lambda <- as.numeric(draws[, "lambda"])
quantile(lambda, c(0.05, 0.5, 0.95))
quantile(1 / sqrt(lambda), c(0.05, 0.5, 0.95))
```

The second line is the standard deviation of the coefficients.

## The full model and what is sampled

Put together, the model has four parts,

$$
\begin{aligned}
\hat\theta_i \mid \theta_i &\sim N(\theta_i, \sigma_i^2), \\
P(\theta_i = \tau_j \mid \alpha) &= g_j(\alpha), \\
\alpha_m \mid \lambda &\sim N(0,\, 1/\lambda), \\
\lambda &\sim \mbox{half-Cauchy}(0, 5),
\end{aligned}
$$

and each quantity depends only on the one before it:
$\lambda \rightarrow \alpha \rightarrow g \rightarrow \theta_i \rightarrow
\hat\theta_i$.

Because each $\theta_i$ takes values on the grid, it can be summed out and
need not be sampled, which gives the likelihood of the coefficients,

$$
p(\hat\theta \mid \alpha) = \prod_{i=1}^{K} \sum_{j=1}^{L}
  g_j(\alpha)\, \phi(\hat\theta_i;\, \tau_j, \sigma_i) .
$$

The sampler therefore works in $M + 1$ dimensions, $\alpha$ and $\lambda$,
whatever the number of sites. The number of sites enters only through the
cost of computing the likelihood, which is that of $K \times L$ normal
densities computed once and a matrix product at each step. Sampling is by
the No-U-Turn sampler of Stan [@hoffman2014; @carpenter2017]. The logarithm
of the likelihood is stored for each draw as `log_marginal_likelihood`.
The name refers to the true effects, which have been summed out; it is not
the marginal likelihood of the model, in which $\alpha$ and $\lambda$ would
be integrated out as well.

The result is a set of draws $\alpha^{(1)}, \ldots, \alpha^{(S)}$ from the
posterior distribution of the coefficients, and with them draws
$g(\alpha^{(s)})$ of the distribution of effects. Step (a) thus ends not
with one estimate of $g$ but with a posterior distribution over $g$.

```{r g-draws, fig.alt = "Forty posterior draws of the distribution of effects and their mean."}
grid <- fit$metadata$data_list$grid
g <- exp(draws[, sprintf("log_g[%d]", seq_along(grid))])
g <- matrix(as.numeric(g), nrow = nrow(g))
some <- round(seq(1, nrow(g), length.out = 40))
matplot(grid, t(g[some, ]), type = "l", lty = 1, col = "gray70",
        xlab = "Effect", ylab = "Probability of the grid point")
lines(grid, colMeans(g), lwd = 2)
```

The gray lines are 40 of the posterior draws of $g$ for the teacher
expectancy studies and the black line is their mean. All of them have a
peak near zero. They differ in how sharp the peak is and in how much
probability they give to the right tail, which 19 studies cannot settle.

## The posterior distribution of a site effect

For a given $\alpha$, Bayes' rule on the grid gives the posterior
probabilities of the effect of site $i$,

$$
w_{ij}(\alpha) = \frac{g_j(\alpha)\, \phi(\hat\theta_i;\, \tau_j, \sigma_i)}
  {\sum_{l=1}^{L} g_l(\alpha)\, \phi(\hat\theta_i;\, \tau_l, \sigma_i)},
$$

and from them a mean $\sum_j w_{ij} \tau_j$, a variance and a mode. This is
step (b) for one candidate $g$. An empirical Bayes analysis stops here, with
$\alpha$ replaced by an estimate.

The posterior distribution of $\theta_i$ given the data averages over the
uncertainty about $g$:

$$
P(\theta_i = \tau_j \mid \mbox{data}) =
  \int w_{ij}(\alpha)\, p(\alpha \mid \mbox{data})\, d\alpha
  \approx \frac{1}{S} \sum_{s=1}^{S} w_{ij}(\alpha^{(s)}) .
$$

For every draw $\alpha^{(s)}$ the Stan program computes the mean, the
standard deviation and the mode of $w_{i\cdot}(\alpha^{(s)})$, stored as
`theta_mean`, `theta_sd` and `theta_map`, and draws one value of
$\theta_i$ from it, stored as `theta_rep`. The values of `theta_rep` are
draws from the posterior distribution of $\theta_i$ given the data.
`coef()` returns the average of `theta_mean`, which is the posterior mean,
and `confint()` returns quantiles of `theta_rep`. The posterior mode,
`coef(type = "map")`, is the grid point at which the averaged probabilities
$P(\theta_i = \tau_j \mid \mbox{data})$ are largest.

The posterior variance has two parts,

$$
\mbox{Var}(\theta_i \mid \mbox{data}) =
  E\{\mbox{Var}(\theta_i \mid \alpha)\} +
  \mbox{Var}\{E(\theta_i \mid \alpha)\} ,
$$

where the expectation and the variance outside the braces are over the
posterior distribution of $\alpha$. The first part is the uncertainty
about the effect that would remain if $g$ were known. The second is the
uncertainty about $g$ as it bears on this site. An analysis that plugs in
an estimate of $g$ omits the second part, so its intervals tend to be too
narrow [@lee2025, Section 3.4 and Appendix A].

```{r variance-parts}
site_draws <- function(name) {
  x <- draws[, sprintf("%s[%d]", name, seq_len(nobs(fit)))]
  matrix(as.numeric(x), nrow = nrow(x))
}
theta_mean <- site_draws("theta_mean")
theta_sd <- site_draws("theta_sd")
parts <- data.frame(
  sei = raudenbush1985$sei,
  given_g = sqrt(colMeans(theta_sd^2)),
  about_g = apply(theta_mean, 2, sd)
)
parts$posterior_sd <- sqrt(parts$given_g^2 + parts$about_g^2)
round(parts[c(4, 10, 17, 18), ], 3)
```

```{r variance-parts-check, include = FALSE}
# The next paragraph says that `posterior_sd` is the column `sd` of the
# summary; stop if that is no longer so.
stopifnot(isTRUE(all.equal(parts$posterior_sd, as.data.frame(fit)$sd)))
```

The rows are four of the teacher expectancy studies. `given_g` is the
square root of the first part, `about_g` is the square root of the second
and `posterior_sd` is the posterior standard deviation. The second part is
small for study 18, whose estimate is precise and near the peak of $g$. It
is far from small for studies 4 and 10, whose large estimates fall in the
tail, where the draws of $g$ disagree. `posterior_sd` is the column `sd`
that `summary()` prints for each site.

In the simulations of @lee2025 [Section 6], in which the true effects of
1,500 sites formed two groups, 90% intervals from this model covered the
true effects 88% to 91% of the time. `vignette("implementation")` reports
a check of the same kind with this package for 50 to 1,500 sites.

## Summaries of g

Every draw of $g$ has a mean, a variance and a standard deviation,

$$
\mu(\alpha) = \sum_{j} g_j(\alpha)\, \tau_j, \qquad
v(\alpha) = \sum_{j} g_j(\alpha)\, \{\tau_j - \mu(\alpha)\}^2 ,
$$

stored as `mean_g`, `var_g` and `sd_g`. Their posterior distributions
answer the questions that the mean and the between-site standard deviation
answer in a random-effects model, without the assumption of normality;
`confint(fit, type = "g")` summarizes them. Other features of $g$, such as
the probability of an effect above a given value, are found from draws of
`predict()`.

The fit also reports an effective number of parameters, which is not in
the paper. For a given $\alpha$ it is

$$
\sum_{i=1}^{K} \frac{\mbox{Var}(\theta_i \mid \alpha)}{\sigma_i^2} .
$$

In the normal model each term is the weight that the posterior mean gives
to the site's own estimate, so the sum counts how many estimates' worth of
information the sites keep for themselves: near $K$ when the estimates are
hardly drawn together, near zero when they are drawn to a common value.
With a general $g$ a term can exceed one, and the sum is a rough guide. It
is a different quantity from the `p_loo` that `loo()` reports under the
same name (see `vignette("choosing-a-grid")`).

```{r effective}
unlist(fit$metadata$effective_params_summary)
```

For the `r nobs(fit)` studies the posterior mean is about
`r round(fit$metadata$effective_params_summary$mean)`. `logLik()` returns
the posterior mean of `log_marginal_likelihood` with this number as its
degrees of freedom. Because that is not a maximized log likelihood, `AIC()`
and `BIC()` computed from it do not have their usual meaning, and fits are
better compared with `loo()`.

## References
